---
title: "Geocoding and Mapping Prince George's County Churches"
author: "Derek Willis"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this notebook we'll take a CSV of churches in Prince George's County, Maryland that students gathered, use a free geocoding service to convert their street addresses into latitude/longitude coordinates, and then plot them on a map with `ggplot2`.

We'll use **`tidygeocoder`** with OpenStreetMap's **Nominatim** service, which is free and does not require an API key. Nominatim asks users to send no more than 1 request per second — `tidygeocoder` handles that rate-limiting automatically.

## Load libraries

If you haven't installed these packages, uncomment the install lines and run them once.

```{r}
# install.packages(c("tidyverse", "tidygeocoder", "sf", "tigris", "ggrepel"))

library(tidyverse)
library(tidygeocoder)   # free geocoding via Nominatim/OSM
library(sf)             # simple features for the basemap
library(tigris)         # county boundaries from the Census Bureau
library(ggrepel)        # non-overlapping labels (optional)

# Cache tigris downloads so we don't re-fetch the shapefile every run
options(tigris_use_cache = TRUE)
options(scipen = 999)
```

## Load the data

```{r}
churches <- read_csv("pg_churches.csv")

glimpse(churches)
```

The `address` column contains addresses students collected from church websites. They are not perfectly formatted — some have embedded newlines, some include "United States", some are missing the city or ZIP. We'll need to clean them before sending them to the geocoder.

## Clean the addresses

A geocoder works best with addresses on a single line and without extra noise. We'll do a few things:

1. Replace newlines and tabs with spaces.
2. Collapse repeated whitespace.
3. Strip stray punctuation like trailing commas.
4. Add ", USA" to the end so Nominatim doesn't confuse Maryland street names with places elsewhere in the world.

```{r}
churches_clean <- churches |>
  mutate(
    address_clean = address |>
      str_replace_all("[\r\n\t]+", " ") |>   # newlines/tabs -> space
      str_squish() |>                         # collapse repeated spaces
      str_remove_all(",\\s*$") |>             # drop trailing comma
      str_remove_all("\\s*PO Box\\s*\\d+\\s*") |>   # drop PO Box fragments
      str_squish(),
    # Add country if it isn't already there
    address_clean = if_else(
      str_detect(address_clean, regex("united states|usa", ignore_case = TRUE)),
      address_clean,
      paste0(address_clean, ", USA")
    )
  )

# Spot check
churches_clean |>
  select(name, address, address_clean) |>
  slice_head(n = 5)
```

## Geocode

`tidygeocoder::geocode()` takes a data frame, a column with addresses, and a `method`. We'll use `"osm"` (Nominatim). With ~250 rows and the required 1-second-per-request rate, this will take roughly **4–5 minutes** the first time you run it.

Because geocoding is slow, we cache the result to a local CSV. On subsequent runs the chunk just reads the cache.

```{r geocode, cache=TRUE}
cache_file <- "pg_churches_geocoded.csv"

if (file.exists(cache_file)) {
  churches_geo <- read_csv(cache_file)
} else {
  churches_geo <- churches_clean |>
    geocode(
      address = address_clean,
      method  = "osm",
      lat     = "latitude",
      long    = "longitude",
      # tidygeocoder will sleep between requests as required by Nominatim
      quiet   = FALSE
    )
  write_csv(churches_geo, cache_file)
}
```

## Check geocoding quality

Not every address will return a hit. Let's see how many we got and look at any failures.

```{r}
# How many succeeded?
churches_geo |>
  summarise(
    total      = n(),
    geocoded   = sum(!is.na(latitude)),
    failed     = sum(is.na(latitude)),
    pct_geocoded = round(100 * geocoded / total, 1)
  )

# Which ones failed?
churches_geo |>
  filter(is.na(latitude)) |>
  select(name, address)
```

We should also sanity-check that the coordinates we *did* get actually fall in or near Prince George's County. PG County sits roughly between latitude 38.65–39.10 and longitude -77.10 to -76.65. Anything way outside that bounding box is almost certainly a mis-geocode (a Nominatim hit on a similarly-named street in another state).

```{r}
churches_geo <- churches_geo |>
  mutate(
    in_pg_bbox = latitude  >= 38.6 & latitude  <= 39.1 &
                 longitude >= -77.1 & longitude <= -76.6
  )

churches_geo |>
  filter(!is.na(latitude), !in_pg_bbox) |>
  select(name, address, latitude, longitude)
```

For the map we'll keep only the rows that geocoded successfully and fall inside the bounding box. In a real reporting workflow you'd go back and manually fix the addresses that failed or landed in the wrong place — for the lab we'll flag them and move on.

```{r}
churches_map <- churches_geo |>
  filter(!is.na(latitude), in_pg_bbox)

nrow(churches_map)
```

## Get a basemap of Prince George's County

We'll pull the county boundary from the U.S. Census Bureau's TIGER/Line shapefiles via the `tigris` package. Maryland's state FIPS code is `24` and Prince George's County is `033`.

```{r}
md_counties <- counties(state = "MD", cb = TRUE, year = 2022, progress_bar = FALSE)

pg_county <- md_counties |>
  filter(GEOID == "24033")
```

## Plot the churches

A clean first map: the county outline, with one point per church.

```{r fig.width=9, fig.height=9}
ggplot() +
  geom_sf(data = pg_county, fill = "grey95", color = "grey40") +
  geom_point(
    data = churches_map,
    aes(x = longitude, y = latitude),
    color = "#cc3333",
    alpha = 0.7,
    size  = 2
  ) +
  labs(
    title    = "Churches in Prince George's County, Maryland",
    subtitle = paste0(nrow(churches_map), " churches geocoded via OpenStreetMap"),
    caption  = "Source: Student-collected church data; basemap from U.S. Census TIGER/Line",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank()
  )
```

## Color by denomination

The dataset has a `denomination` column. There are a lot of categories, so let's keep the top denominations and lump the rest into "Other" so the legend stays readable.

```{r fig.width=10, fig.height=9}
churches_map <- churches_map |>
  mutate(
    denomination = str_squish(denomination),
    denom_top    = fct_lump_n(denomination, n = 8, other_level = "Other")
  )

ggplot() +
  geom_sf(data = pg_county, fill = "grey95", color = "grey40") +
  geom_point(
    data = churches_map,
    aes(x = longitude, y = latitude, color = denom_top),
    alpha = 0.8,
    size  = 2.2
  ) +
  scale_color_brewer(palette = "Set1", name = "Denomination") +
  labs(
    title    = "Churches in Prince George's County by Denomination",
    subtitle = paste0(nrow(churches_map), " churches geocoded"),
    caption  = "Source: Student-collected church data; basemap from U.S. Census TIGER/Line",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    legend.position = "right"
  )
```

## Where are the clusters?

A simple density plot can highlight which parts of the county have the most churches in our dataset.

```{r fig.width=9, fig.height=9}
ggplot() +
  geom_sf(data = pg_county, fill = "grey95", color = "grey40") +
  stat_density_2d(
    data = churches_map,
    aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom  = "polygon",
    alpha = 0.4
  ) +
  geom_point(data = churches_map,
             aes(x = longitude, y = latitude),
             size = 0.8, color = "grey20") +
  scale_fill_viridis_c(name = "Density") +
  labs(
    title = "Where churches cluster in Prince George's County",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), axis.text = element_blank())
```

## Wrap-up: questions to take to reporting

A few things worth following up on once the data is mapped:

- Which addresses failed to geocode, and why? Were they entered incorrectly or do those churches not have a clear physical address?
- Are denominations spatially clustered? Catholic parishes, for example, are typically organized geographically by diocese — does that show up here?
- Which neighborhoods or municipalities have the highest density of churches per capita? (You'd join to ACS population data via `tidycensus` to answer that.)
- Are there ZIP codes or census tracts in PG County with no churches in our dataset at all? That could indicate either a real gap or an undercount in the student collection effort.

ANSWER HERE

## A note on the geocoder

OpenStreetMap's Nominatim is free and good enough for reporting work at this scale, but it has limits:

- **Rate limit:** 1 request/second. Don't try to push it.
- **Coverage:** OSM data quality varies. U.S. street addresses are generally well-covered, but rural addresses or very new construction may not be.
- **Terms of use:** For larger jobs (thousands of addresses, repeated production runs), set up a self-hosted Nominatim instance or use a paid service like Google or Mapbox. The `tidygeocoder` package supports many of those backends with the same `geocode()` interface — you just change the `method` argument.
