# Intro to APIs: The Census

There is truly an astonishing amount of data collected by the US Census Bureau. First, there's the Census that most people know -- the every 10 year census. That's the one mandated by the Constitution where the government attempts to count every person in the US. It's a mind-boggling feat to even try, and billions get spent on it. That data is used first for determining how many representatives each state gets in Congress. From there, the Census gets used to divide up billions of dollars of federal spending.

To answer the questions the government needs to do that, a ton of data gets collected. That, unfortunately, means the Census is exceedingly complicated to work with. The good news is, the Census has an API -- an application programming interface. What that means is we can get data directly through the Census Bureau via calls over the internet.

Let's demonstrate.

We're going to use a library called `tidycensus` which makes calls to the Census API in a very tidy way, and gives you back tidy data. That means we don't have to go through the process of importing the data from a file. I can't tell you how amazing this is, speaking from experience. The documentation for this library is [here](https://walker-data.com/tidycensus/). Another R library for working with Census APIs (there is more than one) is [this one](https://github.com/hrecht/censusapi) from Hannah Recht, a journalist with Kaiser Health News.

First we need to install `tidycensus` using the console: `install.packages("tidycensus", dependencies = TRUE)`. You also should install the `sf` and `rgdal` packages.

```{r}
#| output: false
library(tidyverse)
library(janitor)
library(tidycensus)
```

To use the API, you need an API key. To get that, you need to [apply for an API key with the Census Bureau](https://api.census.gov/data/key_signup.html). It takes a few minutes and you need to activate your key via email. Once you have your key, you need to set that for this session. Just FYI: Your key is your key. Do not share it around.

```{r, eval=FALSE}
census_api_key("YOUR KEY HERE", install=TRUE)
```

The two main functions in tidycensus are `get_decennial`, which retrieves data from the 2000, 2010 and 2020 Censuses, and `get_acs`, which pulls data from the American Community Survey, a between-Censuses annual survey that provides estimates, not hard counts, but asks more detailed questions. If you're new to Census data, there's [a very good set of slides from Kyle Walker](http://walker-data.com/umich-workshop/census-data-in-r/slides/#1), the creator of tidycensus, and he's working on a [book](https://walker-data.com/census-r/index.html) that you can read for free online.

It's important to keep in mind that Census data represents people - you, your neighbors and total strangers. It also requires some level of definitions, especially about race & ethnicity, that may or may not match how you define yourself or how others define themselves.

So to give you some idea of how complicated the data is, let's pull up just one file from the decennial Census. We'll use Summary File 1, or SF1. That has the major population and housing stuff.

```{r}
sf1 <- load_variables(2010, "sf1", cache = TRUE)

sf1
```

Note: There are thousands of variables in SF1. That's not a typo. Open it in your environment by double clicking. As you scroll down, you'll get an idea of what you've got to choose from.

If you think that's crazy, try the SF3 file from 2000.

```{r}
sf3 <- load_variables(2000, "sf3", cache = TRUE)

sf3
```

Yes. That's more than 16,000 variables to choose from. I told you. Astonishing.

So let's try to answer a question using the Census. What is the fastest growing state since 2000?

To answer this, we need to pull the total population by state in each of the decennial census. Here's 2000.

```{r, eval=FALSE}
p00 <- get_decennial(geography = "state", variables = "P001001", year = 2000)

```

Now 2010.

```{r, eval=FALSE}
p10 <- get_decennial(geography = "state", variables = "P001001", year = 2010)
```

Let's take a peek at 2010.

```{r, eval=FALSE}
p10
```

As you can see, we have a GEOID, NAME, then variable and value. Variable and value are going to be the same. Because those are named the same thing, to merge them together, we need to rename them.

```{r, eval=FALSE}
p10 |> select(GEOID, NAME, value) |> rename(Population2010=value) -> p2010

p00 |> select(GEOID, NAME, value) |> rename(Population2000=value) -> p2000
```

Now we join the data together.

```{r, eval=FALSE}
alldata <- p2000 |> inner_join(p2010)
```

And now we calculate the percent change.

```{r, eval=FALSE}
alldata |> mutate(change = ((Population2010-Population2000)/Population2000)*100) |> arrange(desc(change))
```

And just like that: Nevada.

You may be asking: hey, wasn't there a 2020 Census? Where's that data? The answer is that it's coming, slowly - the Census Bureau has a [schedule of releases](https://www.census.gov/programs-surveys/popest/about/schedule.html).

## The ACS

In 2010, the Census Bureau replaced SF3 with the American Community Survey. The Good News is that the data would be updated on a rolling basis. The bad news is that it's more complicated because it's more like survey data with a large sample. That means there's margins of error and confidence intervals to worry about. By default, using `get_acs` fetches data from the 5-year estimates, but you can specify 1-year estimates for jurisdictions with at least 65,000 people (many counties and cities).

Here's an example using the 5-year ACS estimates:

What is Maryland's richest county?

We can measure this by median household income. That variable is `B19013_001`, so we can get that data like this (I'm narrowing it to the top 20 for simplicity):

```{r}
md <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2024)

md <- md |> arrange(desc(estimate)) |> top_n(20, estimate)

md

```

Howard, Calvert, Montgomery, Anne Arundel, Charles. What do they all have in common? Lots of suburban flight from DC and Baltimore. But do the margins of error let us say one county is richer than the other. We can find this out visually using error bars. Don't worry much about the code here -- we'll cover that soon enough.

```{r}
md |>
  mutate(NAME = gsub(" County, Maryland", "", NAME)) |>
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red") +
  labs(title = "Household income by county in Maryland",
       subtitle = "2020-2024 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```

As you can see, some of the error bars are quite wide. Some are narrow. But if the bars overlap, it means the difference between the two counties is within the margin of error, and the differences aren't statistically significant. So is the difference between Calvert and Montgomery significant? Nope. Is the difference between Howard and everyone else significant? Yes it is.

Let's ask another question of the ACS -- has every Maryland county's median income recovered from the time of the global financial crisis to the current 5-year window?

Let's re-label our first household income data.

```{r}
md24 <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2024)
```

And now we grab the 2010 median household income.

```{r}
md10 <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2010)
```

What I'm going to do next is a lot, but each step is simple. I'm going to join the data together, so each county has one line of data. Then I'm going to rename some fields that repeat. Then I'm going to calculate the minimum and maximum value of the estimate using the margin of error. That'll help me later. After that, I'm going to calculate a percent change and sort it by that change.

```{r}
md10 |>
  inner_join(md24, join_by(GEOID, NAME)) |>
  rename(estimate2010=estimate.x, estimate2024=estimate.y) |>
  mutate(min2010 = estimate2010-moe.x, max2010 = estimate2010+moe.x, min2024 = estimate2024-moe.y, max2024 = estimate2024+moe.y) |>
  select(-variable.x, -variable.y, -moe.x, -moe.y) |>
  mutate(change = ((estimate2024-estimate2010)/estimate2010)*100) |>
  arrange(change)
```

So according to this, Somerset and Caroline counties had the smallest change between 2010 and 2024, while all other jurisdictions saw double-digit percentage increases.

But did they?

Look at the min and max values for both. Is the change statistically significant?

The ACS data has lots of variables, just like the decennial Census does. To browse them, you can do this:

```{r}
v24 <- load_variables(2024, "acs5", cache=TRUE)
```

And then view `v24` to see what kinds of variables are available via the API.

## "Wide" Results

Although one of the chief strengths of tidycensus is that it offers a, well, tidy display of Census data, it also has the ability to view multiple variables spread across columns. This can be useful for creating percentages and comparing multiple variables.

## Sorting Results

You'll notice that we've used `arrange` to sort the results of tidycensus functions, although that's done after we create a new variable to hold the data. There's another way to use `arrange` that you should know about, one that you can use for exploratory analysis. An example using median household income:

```{r}
md24 <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2024)
arrange(md24, desc(estimate))
```

In this case we don't save the sorted results to a variable, we can just see the output in the console.

## Combining with Other Data

Combining Census data with other types of data really unlocks its power for journalism. In this case, we'll use some 2019-20 data from Maryland high schools, including the number of graduates and the number of students who enrolled in college the next fall semester, plus the percentage. We also have the zip code of the high school, which will enable us to bring in Census data. Let's load the CSV file and make sure the zip code is a character column, because that's what we'll need to join.

```{r}
md_high_schools <- read_csv("data/md_high_schools.csv") |>
  mutate(zip = as.character(zip))
```

The Census Bureau has Zip Code Tabulation Areas - they are mostly identical to what we know as zip codes, with some small differences. In most cases we can use them interchangeably. Let's get ZCTA household income data from the ACS using tidycensus for each zip code in `md_high_schools`. Run the following codeblock to retrieve household income data for all Maryland ZCTAs first, then join them to `md_high_schools` using the GEOID column. We'll also create min and max values and drop any rows where we don't have income estimates. What's one question we could ask of this data now that we have enrollment percentages and household income for the high school's zip code? What might make the answer harder to interpret?

```{r}
md_zipcodes <- get_acs(geography="zcta", variables = "B19013_001", state='MD', year=2019)
md_high_schools_with_acs <- md_high_schools |> left_join(md_zipcodes, by=c('zip'='GEOID'))
md_high_schools_with_acs <- md_high_schools_with_acs |> mutate(min=(estimate-moe), max=(estimate+moe)) |> drop_na()
```

Now let's look at the average household income for high schools that have 60% or more graduates immediately enrolling in college, and then look at those with less than 40%. What do you see here?

```{r}
md_high_schools_with_acs |>
  filter(percent_of_high_school_graduates_enrolled >= 60) |>
  summarize(avg_income = mean(estimate), count =n())

md_high_schools_with_acs |>
  filter(percent_of_high_school_graduates_enrolled <= 40) |>
  summarize(avg_income = mean(estimate), count = n())
```

## Joining Census Data with U.S. Religion Census Data

Now that we've learned how to work with Census API data, let's combine it with data from the 2020 U.S. Religion Census. This will let us explore the relationship between demographic characteristics and religious adherence patterns.

The key to joining these datasets is the FIPS code - a standardized geographic identifier used by the Census Bureau. State FIPS codes are 2 digits, and county FIPS codes are 5 digits (state + county).

Let's start by reading in the U.S. Religion Census data at both the state and county levels. We'll use `clean_names()` to convert the column names to a consistent format: lowercase, with spaces and special characters replaced by underscores. One thing to note: R column names can't start with a number, so `2020 Population` becomes `x2020_population` -- the `x` is added automatically as a prefix.

```{r}
religion_state <- read_csv("data/usreligioncensus_state_2020.csv") |> clean_names()
religion_county <- read_csv("data/usreligioncensus_county_2020.csv") |> clean_names()

glimpse(religion_state)
glimpse(religion_county)
```

Notice that the state data has a `state_code` column (FIPS code) and the county data has a `fips` column (5-digit county FIPS). These are the keys we'll use to join with Census data.

### State-Level Analysis

Let's get state-level population data from the 2020 Census and join it with the religion census data to explore which states have the highest percentage of religious adherents.

```{r}
# Get 2020 state population from Census
state_pop_2020 <- get_decennial(
  geography = "state",
  variables = "P1_001N",  # Total population
  year = 2020
)

# Clean the religion data and prepare for joining
religion_state_clean <- religion_state |>
  # Make sure state_code is character for joining
  mutate(state_code = as.character(state_code)) |>
  # Pad single-digit state codes with leading zero
  mutate(state_code = str_pad(state_code, width = 2, pad = "0"))

# Join the datasets
state_religion_combined <- state_pop_2020 |>
  inner_join(religion_state_clean, join_by(GEOID == state_code))

# View the combined data
state_religion_combined |>
  select(NAME, value, state_name, adherents, adherents_as_percent_of_population, diversity_index_value) |>
  arrange(desc(adherents_as_percent_of_population))
```

The U.S. Religion Census calculates a diversity index that measures how religiously diverse each state is. Let's examine religious diversity by state to see if there are any patterns:

```{r}
# Most diverse states
state_religion_combined |>
  select(NAME, value, diversity_index_value, adherents, adherents_as_percent_of_population) |>
  arrange(desc(diversity_index_value)) |>
  head(10)

# Compare to least diverse
state_religion_combined |>
  select(NAME, value, diversity_index_value, adherents, adherents_as_percent_of_population) |>
  arrange(diversity_index_value) |>
  head(10)
```

### County-Level Analysis: Maryland Example

Now let's work with county-level data to explore Maryland counties specifically. We'll join ACS income data with religion census data to see if there's a relationship between median household income and religious adherence.

The religion census data includes a `congregations_per_100_000_population` column, but that rate isn't very intuitive for Maryland counties -- many of which have well under 100,000 people. Garrett County, for example, has about 29,000 people and 96 congregations, which works out to 333 per 100,000. But Garrett County's population is very far from 100,000 people, so that number is hard to grasp. A per-1,000 rate is more useful here: Garrett County has about 3.3 congregations per 1,000 residents, or roughly 1 for every 300 people.

```{r}
# Get Maryland county median income
md_income <- get_acs(
  geography = "county",
  variables = c(medincome = "B19013_001"),
  state = "MD",
  year = 2021
)

# Filter religion data for Maryland counties only
md_religion <- religion_county |>
  filter(state_name == "Maryland") |>
  mutate(fips = as.character(fips)) |>
  mutate(congregations_per_1k = congregations / x2020_population * 1000)

# Join the data
md_combined <- md_income |>
  inner_join(md_religion, join_by(GEOID == fips))

# Examine the relationship between income and religious adherence
md_combined |>
  select(NAME, estimate, county_name, adherents_as_percent_of_population, diversity_index_value, congregations_per_1k) |>
  arrange(desc(estimate))
```

We can dig deeper into the Maryland data by looking at which counties have the most religious diversity and how that relates to other characteristics:

```{r}
# Most diverse Maryland counties
md_combined |>
  select(NAME, estimate, diversity_index_value, adherents_as_percent_of_population, largest_body_by_number_of_adherents) |>
  arrange(desc(diversity_index_value)) |>
  head(10)

# Least diverse Maryland counties
md_combined |>
  select(NAME, estimate, diversity_index_value, adherents_as_percent_of_population, largest_body_by_number_of_adherents) |>
  arrange(diversity_index_value) |>
  head(10)
```

### Exploring Congregations Per Capita

Let's explore whether wealthier counties have more or fewer congregations per capita, using our per-1,000 rate:

```{r}
md_combined |>
  select(NAME, estimate, congregations_per_1k, adherents_as_percent_of_population) |>
  arrange(desc(congregations_per_1k)) |>
  head(10)
```

This combination of Census demographic data with U.S. Religion Census data opens up many possibilities for analyzing the relationship between community characteristics and religious patterns. The FIPS code system makes it straightforward to join these datasets at both state and county levels, allowing for rich comparative analysis.
