# Filters and selections

More often than not, we have more data than we want. Sometimes we need to be rid of that data. In `dplyr`, there's two ways to go about this: filtering and selecting.

**Filtering creates a subset of the data based on criteria**. All records where the amount is greater than 150,000. All records that match "College Park". Something like that. **Filtering works with rows -- when we filter, we get fewer rows back than we start with.**

**Selecting simply returns only the fields named**. So if you only want to see city and amount, you select those fields. When you look at your data again, you'll have two columns. If you try to use one of your columns that you had before you used select, you'll get an error. **Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with.**

Let's continue to work with the Maryland churches data we used in the previous chapter. First, we need to load the tidyverse:

```{r}
#| output: false
library(tidyverse)
```

```{r}
md_churches <- read_csv("data/churches_md.csv")
```

If we want to see only those churches located in a particular city, we can use the `filter` function to isolate just those records. Filter works with something called a comparison operator. We need to filter all records equal to "COLLEGE PARK". The comparison operators in R, like most programming languages, are == for equal to, != for not equal to, \> for greater than, \>= for greater than or equal to and so on.

**Be careful: `=` is not `==` and `=` is not "equal to". `=` is an assignment operator in most languages -- how things get named.**

```{r}
cp_churches <- md_churches |> filter(CITY == "COLLEGE PARK")

head(cp_churches)
```

And just like that, we have just College Park results, which we can verify looking at the head, the first six rows.

We also have more data than we might want. For example, we may only want to work with the church name and street.

To simplify our dataset, we can use select.

```{r}
selected_cp_churches <- cp_churches |> select(NAME, STREET)

head(selected_cp_churches)
```

And now we only have two columns of data for whatever analysis we might want to do.

## Combining filters

So let's say we wanted to see all the churches in College Park that are independent - in other words, that aren't affiliated with a larger denomination. The data we have comes from the Internal Revenue Service and has a [code book](https://www.irs.gov/pub/irs-soi/eo-info.pdf) describing the values of the columns. The AFFILIATION column says that a code of 3 means independent. We can do this a number of ways. The first is we can chain together a whole lot of filters.

```{r}
cp_churches_ind <- md_churches |> filter(CITY == "COLLEGE PARK") |> filter(AFFILIATION == 3)

nrow(cp_churches_ind)
```

That gives us 28 records But that's repetitive, no? We can do better using a single filter and boolean operators -- AND and OR. In this case, AND is `&` and OR is `|`.

The difference? With AND, all conditions must be true to be included. With OR, any of those conditions things can be true and it will be included.

Here's the difference.

```{r}
cp_churches_ind <- md_churches |> filter(CITY == "COLLEGE PARK" & AFFILIATION == 3)

nrow(cp_churches_ind)
```

So AND gives us the same answer we got before. What does OR give us?

```{r}
cp_churches_ind <- md_churches |> filter(CITY == "COLLEGE PARK" | AFFILIATION == 3)

nrow(cp_churches_ind)
```

So there's 4,875 rows that are EITHER College Park churches OR are independent. OR is additive; AND is restrictive.

A general tip about using filter: it's easier to work your way towards the filter syntax you need rather than try and write it once and trust the result. Each time you modify your filter, check the results to see if they make sense. This adds a little time to your process but you'll thank yourself for doing it because it helps avoid mistakes.
