# Combining and joining

Often, as data journalists, we're looking at data across time or at data stored in multiple tables. And to do that, we need to merge, reshape, and join data together.

We might have data in multiple files that needs to be combined. We might have data in a "wide" format that needs to be converted to "long" format for analysis. And we might have datasets that need to be joined together on common elements.

Let's explore these techniques using FEC contribution data from Maryland clergy and United Methodist church membership data.

## Combining data (stacking)

Let's say that we have Maryland clergy contribution data from five different election cycles in five different files. They have the same record layout. We can combine them into a single dataframe.

Let's do what we need to import them properly. I've merged it all into one step for each of the datasets.

```{r}
#| output: false
library(tidyverse)
```

```{r}
clergy_contribs_2016 <- read_csv("data/clergy_contribs_2016.csv")
clergy_contribs_2018 <- read_csv("data/clergy_contribs_2018.csv")
clergy_contribs_2020 <- read_csv("data/clergy_contribs_2020.csv")
clergy_contribs_2022 <- read_csv("data/clergy_contribs_2022.csv")
clergy_contribs_2024 <- read_csv("data/clergy_contribs_2024.csv")
```

All of these datasets have the same number of columns, all with the same names, so if we want to merge them together to compare them over time, we need to stack them together. The verb here, in R, is `bind_rows`. You tell the function what you want to combine and it does it, assuming that you've got column names in common containing identically formatted data.

Since we have five dataframes, we're going to need to pass them as a list, meaning they'll be enclosed inside the `list` function.

```{r}
clergy_contribs_combined <- bind_rows(list(clergy_contribs_2016, clergy_contribs_2018, clergy_contribs_2020, clergy_contribs_2022, clergy_contribs_2024))
```

And boom, like that, we have more than 42,000 rows of data together instead of five dataframes. Now we can ask more interesting questions like how clergy giving patterns have changed over time.

There are plenty of uses for `bind_rows`: any regularly updated data that comes in the same format like crime reports, award recipients, or campaign contributions across multiple election cycles.

## Reshaping data with pivot functions

Now that we have all our contribution data in one place, let's explore how to reshape it. Sometimes data is easier to analyze or visualize in different formats. R provides two key functions for this: `pivot_wider` and `pivot_longer`.

Before we can reshape the data, we need to summarize it. Let's calculate the total amount each person contributed in each election cycle.

```{r}
clergy_totals_by_cycle <- clergy_contribs_combined |>
  group_by(contributor_name, two_year_transaction_period) |>
  summarize(
    total_amount = sum(contribution_receipt_amount),
    num_contributions = n(),
    .groups = 'drop'
  )

glimpse(clergy_totals_by_cycle)
```

### Converting from long to wide format

Right now our data is in "long" format - each row represents one person's contributions in one election cycle. But what if we want to see all of someone's contributions across cycles in a single row? That's where `pivot_wider` comes in.

Before we pivot, we need to select only the columns we want to reshape. We'll keep `contributor_name`, `two_year_transaction_period`, and `total_amount`. This ensures we get exactly one row per contributor in the wide format.

```{r}
clergy_wide <- clergy_totals_by_cycle |>
  select(contributor_name, two_year_transaction_period, total_amount) |>
  pivot_wider(
    names_from = two_year_transaction_period,
    values_from = total_amount,
    values_fill = 0
  )

glimpse(clergy_wide)
```

Notice how each election cycle (2016, 2018, 2020, 2022, 2024) is now its own column. This format makes it easy to compare someone's giving patterns across cycles at a glance. For example, we could find contributors who gave in all five election cycles:

```{r}
clergy_wide |>
  filter(`2016` > 0 & `2018` > 0 & `2020` > 0 & `2022` > 0 & `2024` > 0) |>
  arrange(desc(`2024`))
```

### Converting from wide to long format

Sometimes you need to convert data from wide back to long format. This is common when preparing data for visualization with ggplot2, which generally prefers long format data.

```{r}
clergy_long <- clergy_wide |>
  pivot_longer(
    cols = c(`2016`, `2018`, `2020`, `2022`, `2024`),
    names_to = "election_cycle",
    values_to = "total_contributions"
  )

glimpse(clergy_long)
```

Notice that `pivot_longer` takes:

-   `cols`: which columns to pivot (the cycle years)
-   `names_to`: what to call the new column that will hold the old column names
-   `values_to`: what to call the new column that will hold the values

This long format is now ready for analysis or visualization by election cycle.

## Joining data

More complicated is when you have two separate tables that are connected by a common element or elements. But there's a verb for that, too: `join`.

Let's start by reading in some United Methodist church clergy data. We have two files: one with basic information about clergy, and another with more detailed church assignment data. Clergy can be assigned to multiple churches over their career, but they are the same person.

```{r}
umdata_people <- read_csv('data/umdata_people.csv')
umdata_people_all <- read_csv('data/umdata_people_all.csv')

glimpse(umdata_people)
glimpse(umdata_people_all)
```

You should notice that `umdata_people_all` has two additional columns: `ConferenceId` and `ConferenceName`. These contain the detailed church assignment information. When we join these datasets, we'll select only these unique columns from `umdata_people_all` to avoid creating duplicate columns with .x and .y suffixes.

### Understanding the relationship between datasets

Before joining data, it's crucial to understand how the tables relate to each other.

```{r}
# Count unique people in each dataset
umdata_people |>
  summarize(unique_people = n_distinct(GCFAId))

umdata_people_all |>
  summarize(unique_people = n_distinct(GCFAId))

# Count total rows
nrow(umdata_people)
nrow(umdata_people_all)
```

The `umdata_people_all` dataset has many more rows than unique people because each person can have multiple church assignments over time. This is called a one-to-many relationship.

### Types of joins

There are different kinds of joins. It's better if you think of two tables sitting next to each other. A `left_join` takes all the records from the left table and only the records that match in the right one. A `right_join` does the same thing. An `inner_join` takes only the records where they are equal. There's one other join -- a `full_join` which returns all rows of both, regardless of if there's a match -- but I've never once had a use for a full join.

An `inner_join` keeps only the records that exist in BOTH datasets. Let's find the people who appear in both datasets.

Since both datasets share most columns, we'll select only the unique information from `umdata_people_all` (ConferenceId and ConferenceName) to avoid duplicate columns with .x and .y suffixes.

```{r}
people_matched <- umdata_people |>
  inner_join(
    umdata_people_all |> select(GCFAId, ConferenceId, ConferenceName),
    join_by(GCFAId)
  )

nrow(people_matched)
glimpse(people_matched)
```

A `left_join` keeps ALL records from the left (first) dataset and only matching records from the right (second) dataset. This is the most commonly used join type.

```{r}
people_with_conferences <- umdata_people |>
  left_join(
    umdata_people_all |> select(GCFAId, ConferenceId, ConferenceName),
    join_by(GCFAId)
  )

nrow(people_with_conferences)
glimpse(people_with_conferences)
```

Notice that we're using `join_by()` to specify the column to join on. Both datasets have a column called `GCFAId` (General Council on Finance and Administration ID), which uniquely identifies each person in the United Methodist system.

### Understanding one-to-many joins

When you join a table with unique records to a table with multiple records per key, the result will have multiple rows for each match. Let's examine this:

```{r}
# Look at the first person
sample_person <- umdata_people |> slice(1)
sample_gcfaid <- sample_person$GCFAId

people_with_conferences |>
  filter(GCFAId == sample_gcfaid) |>
  select(FirstName, LastName, GCFAId, Conferences, ConferenceName, AccountStatus)
```

Now we can use the joined data to answer questions. For example, we can find out how many different conference assignments people have:

```{r}
conference_counts <- people_with_conferences |>
  group_by(GCFAId, FirstName, LastName) |>
  summarize(
    num_conferences = n(),
    conferences_list = paste(unique(ConferenceName), collapse = "; "),
    .groups = 'drop'
  ) |>
  arrange(desc(num_conferences))

head(conference_counts, 10)
```

### Anti-joins for finding missing data

An `anti_join` keeps records from the left dataset that DON'T have matches in the right dataset. This is useful for finding missing data:

```{r}
people_not_in_all <- umdata_people |>
  anti_join(umdata_people_all, join_by(GCFAId))

nrow(people_not_in_all)
```

In this case, all people in `umdata_people` also appear in `umdata_people_all`, so we get zero rows. But if there were people missing from the larger dataset, this would help us find them.

### Handling data type mismatches

Sometimes joins look like they should work but don't. Often this is due to the two columns you're joining on having different data types: joining a `<chr>` column to a `<dbl>` column, for example. Let's walk through an example of that using some demographic data by zip code.

```{r}
maryland_zcta <- read_csv('data/maryland_zcta.csv')
glimpse(maryland_zcta)
```

You can see that ZCTA5N, the column representing the Zip Code Tabulation Area, is a numeric column. But should it be? Do we ever want to know the average zip code in Maryland? Zip codes and ZCTAs look like numbers but really are character columns. Let's change that so that we can be sure to join them correctly with other data where the zip codes are not numbers. We'll use `mutate`:

```{r}
maryland_zcta <- maryland_zcta |> mutate(across(ZCTA5N, as.character))
glimpse(maryland_zcta)
```

What's happening here is that we're telling R to take all of the values in the ZCTA5N column and make them "as.character". If we wanted to change a column to numeric, we'd do "as.numeric". When you join two dataframes, the join columns *must* be the same datatype.
