# Mutating data

Often the data you have will prompt questions that it doesn't immediately answer. Religious affiliation data, for example, might have raw counts of priests and parishes but we often don't use those to make comparisons across time periods unless the numbers are small. We need ratios and percentages!

To do that in R, we can use `dplyr` and `mutate` to calculate new metrics in a new field using existing fields of data. That's the essence of `mutate` - using the data you have to answer a new question.

So first we'll import the tidyverse so we can read in our data and begin to work with it.

```{r}
#| output: false
library(tidyverse)
```

Now we'll import a dataset of Catholic Church statistics from selected years that is in the data folder in this chapter's pre-lab directory. We'll use this to explore ways to create new information from existing data.

```{r}
cara_stats <- read_csv('data/cara.csv')
```

Let's add a column called `catholics_per_parish` to calculate how many Catholics there are for each parish in each year. This helps us understand how parish size and resource distribution have changed over time. The code to calculate a ratio is pretty simple. Remember, with `summarize`, we used `n()` to count things. With `mutate`, we use very similar syntax to calculate a new value -- a new column of data -- using other values in our dataset.

To calculate this ratio, we need both the Catholic population and the number of parishes. We'll use mutate to create the new column. The key here is to save the dataframe to itself so that our changes stick.

```{r}
cara_stats <- cara_stats |>
  mutate(
    catholics_per_parish = catholic_population / parishes
  )
```

But what do you see right away? Do those numbers look like we expect them to? They might be large numbers with lots of decimal places. Let's make them easier to read by rounding to the nearest whole number:

```{r}
cara_stats <- cara_stats |>
  mutate(
    catholics_per_parish = round(catholic_population / parishes, 0)
  )
```

Now, does this ordering do anything for us? No. Let's fix that with arrange.

```{r}
cara_stats <- cara_stats |>
  mutate(
    catholics_per_parish = round(catholic_population / parishes, 0)
  ) |> 
  arrange(desc(catholics_per_parish))
```

So now we have results ordered by `catholics_per_parish` with the highest ratio first. To see the lowest ratio first, we can reverse that `arrange` function - we don't need to recalculate the column:

```{r}
cara_stats |>
  arrange(catholics_per_parish)
```

The U.S. Catholic Church had the fewest Catholics per parish in 1965, with about 2,494 Catholics per parish. By 2024, that ratio had increased to 4,137 Catholics per parish - a sign of both parish closures and changing demographics. Using `mutate()` allows us to create a comparison that helps us understand those trends.

## Adding multiple calculations

We can also add multiple calculations at once. Let's calculate both the Catholics per parish and the percentage of parishes without a resident priest:

```{r}
cara_stats <- cara_stats |>
  mutate(
    catholics_per_parish = round(catholic_population / parishes, 0),
    pct_no_resident_priest = (parishes_without_resident_priest / parishes) * 100
  ) |> 
  arrange(desc(pct_no_resident_priest))
```

This shows us how the challenge of staffing parishes with resident priests has grown over time - rising from less than 3% of parishes in 1965 to more than 21% in 2024.

## Another use of mutate

Mutate is also useful for standardizing data - for example, making different spellings of, say, campaign spending recipients.

Let's load our Maryland campaign expenditures involving churches into a `md_churches` dataframe, and focus in particular on the `Payee Name` column.

```{r}
church_expenses <- read_csv("data/md_church_expenditures.csv")
church_expenses
```

You'll notice that the payee names have a mix of styles: title case, upper case and lower case. R will treat two names that are spelled the same but have different cases as distinct things, and that will mean that any aggregates we create based on Payee Name won't be accurate.

So how can we fix that? Mutate - it's not just for math! And a function called `str_to_upper` that will convert a character column into all uppercase.

```{r}
standardized_church_expenses <- church_expenses |>
  mutate(
    payee_upper = str_to_upper(`Payee Name`)
)
```

That creates a new column, `payee_upper`, with all uppercase letters. Now we can use that column to group and summarize our data.

There are lots of potential uses for standardization - addresses, zip codes, anything that can be misspelled or abbreviated.

## A more powerful use

Mutate is even more useful when combined with some additional functions. Let's keep rolling with our expenditure data. Take a look at our new `payee_upper` column: it contains the name of the recipient. It would be useful to have a separate `denomination` column indicating what sort of church it is. We can check to see if a denomination name is contained in that column and then populate a new column with the value we want, using the functions `str_detect` and `case_when`. The `case_when` function handles multiple variations, such as if a payee contains "Baptist" or "Catholic", etc. Crucially, we can tell R to populate the new column with `NA` if it doesn't find a match.

```{r}
standardized_church_expenses <- standardized_church_expenses |>
  mutate(
    denomination = case_when(
        str_detect(payee_upper, "BAPTIST") ~ "Baptist",
        str_detect(payee_upper, "CATHOLIC") ~ "Catholic",
        str_detect(payee_upper, "METHODIST") ~ "Methodist",
        str_detect(payee_upper, "PRESBYTERIAN") ~ "Presbyterian",
        str_detect(payee_upper, "LUTHERAN") ~ "Lutheran",
        str_detect(payee_upper, "EPISCOPAL") ~ "Episocopal",
        str_detect(payee_upper, "CHURCH OF CHRIST") ~ "Church of Christ",
        str_detect(payee_upper, "UNITARIAN") ~ "Unitarian",
        str_detect(payee_upper, "TEMPLE") ~ "Jewish",
        .default = NA
      )
  )
```

There's a lot going on here, so let's unpack it. It starts out as a typical mutate statement, but `case_when` introduces some new things. Each line checks to see if the pattern is contained in the `payee_upper` column, followed by `~` and then a value for the new column for records that match that check. You can read it like this: "If we find 'BAPTIST' in the payee_upper column, then put 'Baptist' in the denomination column", and so on for other denominations, and if we don't match any word we're looking for, make denomination `NA`.

We can then use our new `denomination` column in group_by statements to make summarizing easier.

```{r}
standardized_church_expenses |>
  group_by(denomination) |>
  summarize(total = sum(`Transaction Amount`)) |>
  arrange(desc(total))
```

Most expenditures seem to have been paid to Baptist churches, although the large amount of NA transactions suggests that there are many churches we couldn't categorize.

Mutate is there to make your data more useful and to make it easier for you to ask more and better questions of it.
