---
title: "Reverse Engineering: Religious Diversity in the United States"
author: "Student Name"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze Dale E. Jones's report "Religious Diversity in the United States" and reproduce key quantitative findings using data from the 2020 U.S. Religion Census.

**Your task:**

1.  Read the report carefully (ReligiousDiversity-Jones1.txt in this folder)
2.  Identify key facts and statistics that were produced through data analysis
3.  Write well-commented, well-structured R code to reproduce those findings
4.  Focus on replicable quantitative sentences (not the maps)

## The Data

You have two Excel files from the 2020 U.S. Religion Census:

- **2020_USRC_Group_Detail.xlsx**: Breakdowns by individual religious group at the national, state, county, and metro levels. Each row is one religious group in one geographic area.
- **2020_USRC_Summaries.xlsx**: Aggregate totals (all groups combined) at each geographic level, including population, total congregations, and total adherents.

## Load Libraries and Data

```{r}
library(tidyverse)
library(readxl)
```

```{r}
# Load all the sheets we'll need

# Group-level detail by geographic area
groups_nation <- read_excel("2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by Nation")
groups_state <- read_excel("2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by State")
groups_county <- read_excel("2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by County")
groups_metro <- read_excel("2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by Metro")

# Summary totals by geographic area
summary_state <- read_excel("2020_USRC_Summaries.xlsx", sheet = "2020 State Summary")
summary_county <- read_excel("2020_USRC_Summaries.xlsx", sheet = "2020 County Summary")
summary_metro <- read_excel("2020_USRC_Summaries.xlsx", sheet = "2020 Metro Summary")

# Preview
glimpse(groups_state)
glimpse(summary_state)
```

## Finding 1: Religious Body Counts in Specific Areas

**Report claims:**
- Los Angeles County has 171 religious bodies
- New York-Newark-Jersey City metro has 198 religious bodies
- Pennsylvania has 245 religious bodies

A "religious body" is a distinct denomination/group present in the area (with at least one congregation or adherent count reported).

```{r}
# Count distinct religious bodies in Los Angeles County
# Los Angeles County FIPS = 06037
la_bodies <- groups_county %>%
  filter(`County Name` == "Los Angeles County" | FIPS == "06037") %>%
  filter(!is.na(Congregations) | !is.na(Adherents)) %>%
  distinct(`Group Code`) %>%
  nrow()

cat("Los Angeles County religious bodies:", la_bodies, "(report claims: 171)\n")

# Count in New York-Newark-Jersey City metro
# Need to find the CBSA code for this metro
ny_metro <- groups_metro %>%
  filter(str_detect(`Metro Name`, "New York-Newark")) %>%
  filter(!is.na(Congregations) | !is.na(Adherents)) %>%
  distinct(`Group Code`) %>%
  nrow()

cat("New York-Newark-Jersey City metro religious bodies:", ny_metro, "(report claims: 198)\n")

# Count in Pennsylvania
pa_bodies <- groups_state %>%
  filter(`State Name` == "Pennsylvania") %>%
  filter(!is.na(Congregations) | !is.na(Adherents)) %>%
  distinct(`Group Code`) %>%
  nrow()

cat("Pennsylvania religious bodies:", pa_bodies, "(report claims: 245)\n")
```

**Your interpretation:** Do your counts match the report's claims? If slightly different, why might that be?

## Finding 2: Catholic Dominance in Key Areas

**Report claims:**
- Catholics are 47% of Pennsylvania's religious adherents
- Catholics are 61% of the New York metro area's adherents
- Catholics are 62% of all adherents in Los Angeles County

```{r}
# Pennsylvania: Catholic share of total adherents
pa_total_adherents <- summary_state %>%
  filter(`State Name` == "Pennsylvania") %>%
  pull(Adherents)

pa_catholic_adherents <- groups_state %>%
  filter(`State Name` == "Pennsylvania", `Group Code` == "081") %>%
  pull(Adherents)

cat("PA total adherents:", format(pa_total_adherents, big.mark = ","), "\n")
cat("PA Catholic adherents:", format(pa_catholic_adherents, big.mark = ","), "\n")
cat("Catholic share of PA adherents:", round(pa_catholic_adherents / pa_total_adherents * 100, 0), "% (report claims: 47%)\n\n")

# New York metro: Catholic share
ny_metro_name <- groups_metro %>%
  filter(str_detect(`Metro Name`, "New York-Newark")) %>%
  pull(`Metro Name`) %>%
  first()

ny_metro_total <- summary_metro %>%
  filter(str_detect(`Metro Name`, "New York-Newark")) %>%
  pull(Adherents)

ny_metro_catholic <- groups_metro %>%
  filter(str_detect(`Metro Name`, "New York-Newark"), `Group Code` == "081") %>%
  pull(Adherents)

cat("NY metro total adherents:", format(ny_metro_total, big.mark = ","), "\n")
cat("NY metro Catholic adherents:", format(ny_metro_catholic, big.mark = ","), "\n")
cat("Catholic share of NY metro adherents:", round(ny_metro_catholic / ny_metro_total * 100, 0), "% (report claims: 61%)\n\n")

# Los Angeles County: Catholic share
la_total <- summary_county %>%
  filter(`County Name` == "Los Angeles County" | FIPS == "06037") %>%
  pull(Adherents)

la_catholic <- groups_county %>%
  filter((`County Name` == "Los Angeles County" | FIPS == "06037") & `Group Code` == "081") %>%
  pull(Adherents)

cat("LA County total adherents:", format(la_total, big.mark = ","), "\n")
cat("LA County Catholic adherents:", format(la_catholic, big.mark = ","), "\n")
cat("Catholic share of LA County adherents:", round(la_catholic / la_total * 100, 0), "% (report claims: 62%)\n")
```

**Your interpretation:** Do the Catholic percentages match the report?

## Finding 3: State Unaffiliated Percentages

**Report claims:**
- Two-thirds of Oregon, New Hampshire, and Maine populations are unaffiliated
- New Hampshire: 73% unaffiliated
- Maine: 69% unaffiliated
- Oregon: 67% unaffiliated
- Alaska: 65% unaffiliated

The "unaffiliated" population = total population minus total adherents.

```{r}
# Calculate unaffiliated percentage for each state
state_unaffiliated <- summary_state %>%
  mutate(
    unaffiliated = `2020 Population` - Adherents,
    unaffiliated_pct = round(unaffiliated / `2020 Population` * 100, 0)
  ) %>%
  select(`State Name`, `2020 Population`, Adherents, unaffiliated, unaffiliated_pct) %>%
  arrange(desc(unaffiliated_pct))

# Show the top 10 most unaffiliated states
state_unaffiliated %>% slice_head(n = 10)

# Check the specific states from the report
cat("\nSpecific state claims:\n")
for (st in c("New Hampshire", "Maine", "Oregon", "Alaska")) {
  val <- state_unaffiliated %>% filter(`State Name` == st) %>% pull(unaffiliated_pct)
  cat(st, ":", val, "% unaffiliated\n")
}
```

**Your interpretation:** Do the unaffiliated percentages match? Is it true that about two-thirds of OR, NH, and ME populations are unaffiliated?

## Finding 4: Large Metro Area Diversity Comparison

**Report claims:**
- 56 metropolitan areas have 1 million+ population
- Over one-third (20 of 56) rank with the top 20% of counties for diversity
- Only 2 metros would rank with moderate diversity
- None would be categorized with the least diverse counties

The report also provides specific compositions:
- Denver: two-thirds unaffiliated, nearly half the remaining is Catholic
- Memphis: one-third unaffiliated, one-third Evangelical Protestant, ~one-sixth Black Protestant

```{r}
# Identify metro areas with 1M+ population
big_metros <- summary_metro %>%
  filter(`2020 Population` >= 1000000)

cat("Metro areas with 1M+ population:", nrow(big_metros), "(report claims: 56)\n\n")

# Check Denver composition
denver_pop <- big_metros %>%
  filter(str_detect(`Metro Name`, "Denver")) %>%
  pull(`2020 Population`)

denver_adherents <- big_metros %>%
  filter(str_detect(`Metro Name`, "Denver")) %>%
  pull(Adherents)

denver_unaffiliated_pct <- round((denver_pop - denver_adherents) / denver_pop * 100, 0)

denver_catholic <- groups_metro %>%
  filter(str_detect(`Metro Name`, "Denver"), `Group Code` == "081") %>%
  pull(Adherents)

cat("Denver:\n")
cat("  Unaffiliated:", denver_unaffiliated_pct, "% (report claims: ~two-thirds)\n")
cat("  Catholic adherents:", format(denver_catholic, big.mark = ","), "\n")
cat("  Catholic share of all adherents:", round(denver_catholic / denver_adherents * 100, 0), "% (report claims: ~half of affiliated)\n\n")

# Check Memphis composition
memphis_pop <- big_metros %>%
  filter(str_detect(`Metro Name`, "Memphis")) %>%
  pull(`2020 Population`)

memphis_adherents <- big_metros %>%
  filter(str_detect(`Metro Name`, "Memphis")) %>%
  pull(Adherents)

memphis_unaffiliated_pct <- round((memphis_pop - memphis_adherents) / memphis_pop * 100, 0)
cat("Memphis:\n")
cat("  Unaffiliated:", memphis_unaffiliated_pct, "% (report claims: ~one-third)\n")

# Show the top groups in Memphis
memphis_groups <- groups_metro %>%
  filter(str_detect(`Metro Name`, "Memphis")) %>%
  filter(!is.na(Adherents)) %>%
  arrange(desc(Adherents)) %>%
  slice_head(n = 10) %>%
  mutate(pct_of_pop = round(Adherents / memphis_pop * 100, 1)) %>%
  select(`Group Name`, Adherents, pct_of_pop)

memphis_groups
```

**Your interpretation:** Does Denver's composition match "two-thirds unaffiliated"? Does Memphis show the one-third/one-third split the report describes?

## Finding 5: State Diversity Rankings (Table 2)

**Report claims (most diverse states):**
1. Louisiana (37% unaffiliated, 27% Catholic, 25% Evangelical Protestant)
2. South Carolina
3. Texas
4. North Dakota
5. Idaho

**Report claims (least diverse):**
- Alaska (65% unaffiliated), Oregon (67%), Utah (65% Latter-day Saints), Maine (69%), New Hampshire (73%)

For Table 2, the key compositions can be verified by looking at the largest groups in each state and the unaffiliated share.

```{r}
# For Louisiana: check the top religious groups and unaffiliated share
la_pop <- summary_state %>%
  filter(`State Name` == "Louisiana") %>%
  pull(`2020 Population`)

la_adherents <- summary_state %>%
  filter(`State Name` == "Louisiana") %>%
  pull(Adherents)

la_unaffiliated_pct <- round((la_pop - la_adherents) / la_pop * 100, 0)

la_top_groups <- groups_state %>%
  filter(`State Name` == "Louisiana", !is.na(Adherents)) %>%
  arrange(desc(Adherents)) %>%
  slice_head(n = 5) %>%
  mutate(pct_of_pop = round(Adherents / la_pop * 100, 0)) %>%
  select(`Group Name`, Adherents, pct_of_pop)

cat("Louisiana:\n")
cat("  Unaffiliated:", la_unaffiliated_pct, "% (report claims: 37%)\n")
cat("  Top groups:\n")
print(la_top_groups)

# For Utah: check Latter-day Saints share
ut_pop <- summary_state %>%
  filter(`State Name` == "Utah") %>%
  pull(`2020 Population`)

# LDS group code is typically 160
ut_lds <- groups_state %>%
  filter(`State Name` == "Utah", str_detect(`Group Name`, "Latter-day|LDS|Church of Jesus Christ")) %>%
  pull(Adherents) %>%
  sum(na.rm = TRUE)

cat("\nUtah:\n")
cat("  LDS adherents:", format(ut_lds, big.mark = ","), "\n")
cat("  LDS share of population:", round(ut_lds / ut_pop * 100, 0), "% (report claims: 65%)\n")

# DC composition
dc_pop <- summary_state %>%
  filter(str_detect(`State Name`, "District|Columbia|DC")) %>%
  pull(`2020 Population`)

dc_adherents <- summary_state %>%
  filter(str_detect(`State Name`, "District|Columbia|DC")) %>%
  pull(Adherents)

dc_unaffiliated <- round((dc_pop - dc_adherents) / dc_pop * 100, 0)

cat("\nDistrict of Columbia:\n")
cat("  Unaffiliated:", dc_unaffiliated, "% (report claims: 44%)\n")

dc_top_groups <- groups_state %>%
  filter(str_detect(`State Name`, "District|Columbia|DC"), !is.na(Adherents)) %>%
  arrange(desc(Adherents)) %>%
  slice_head(n = 10) %>%
  mutate(pct_of_pop = round(Adherents / dc_pop * 100, 1)) %>%
  select(`Group Name`, Adherents, pct_of_pop)

cat("  Top groups:\n")
print(dc_top_groups)
```

**Your interpretation:** Does Louisiana show ~37% unaffiliated, ~27% Catholic, ~25% Evangelical Protestant? Does Utah show ~65% Latter-day Saints? Does DC show four groups each at 10%+?

## Summary

Summarize what you were able to reproduce and any discrepancies you found. Note that some findings in the report require mapping individual religious groups to the 13 tradition categories (Black Protestant, Buddhist, Catholic, Evangelical Protestant, Hindu, Islam, Jewish, Jehovah's Witnesses, Latter-day Saints, Mainline Protestant, Orthodox Christian, Other Christian, Other Eastern), which requires an external crosswalk from ARDA.

## Challenges and Limitations

Describe any challenges you encountered:
- What data was difficult to obtain or prepare?
- What assumptions did you have to make?
- How did you handle the distinction between individual groups and broader traditions?
- How did you calculate the unaffiliated population?

## Data Sources

-   2020 U.S. Religion Census: <https://www.usreligioncensus.org/>
-   Population data: U.S. Census Bureau (included in the Religion Census files)
