---
title: "Reverse Engineering: Religious Diversity in the United States"
author: "Student Name"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze Dale E. Jones's report "Religious Diversity in the United States" and reproduce key quantitative findings using data from the 2020 U.S. Religion Census.

**Your task:**

1. Read the report carefully (ReligiousDiversity-Jones1.txt in the parent religion-data folder)
2. Identify key facts and statistics that were produced through data analysis
3. Write well-commented, well-structured R code to reproduce those findings
4. Focus on replicable quantitative sentences (not the maps)

**Important:** This is NOT a step-by-step walkthrough. You are expected to independently:
- Identify which claims in the report can be verified with data
- Determine what data you need and how to obtain it
- Clean and analyze the data appropriately
- Implement Simpson's Diversity Index calculation
- Reproduce the findings with your own code

## Understanding Simpson's Diversity Index

Before analyzing the data, make sure you understand the diversity measure used in the report:

**Formula:** D = 1 - ∑(n/N)²

Where:
- N = total number of instances (e.g., total adherents in a county)
- n = instances in each category (e.g., Catholic adherents in that county)
- For each category: divide n by N, square the result, sum all squared results
- Subtract this sum from 1 to get the diversity index
- Result ranges from 0 (no diversity) to 1 (maximum diversity)

## Load Libraries

```{r}
# Load the tidyverse and any other libraries you need


```

## Load Data

```{r}
# Load the 2020 U.S. Religion Census data
# You will need to download this from usreligioncensus.org
# and determine which files/tables to use


```

## Implement Simpson's Diversity Index

```{r}
# Write a function to calculate Simpson's Diversity Index
# This function should:
# 1. Take a dataframe with categories and their counts/proportions
# 2. Calculate (n/N)² for each category
# 3. Sum those squared values
# 4. Subtract from 1 to get the diversity index


```

## Questions to Guide Your Analysis

As you read the report, consider these questions:

1. **What are the main quantitative claims?** Look for specific percentages, counts, rankings, and diversity index values.

2. **What data sources were used?** The report uses the 2020 U.S. Religion Census with different categorization schemes and geographic aggregations.

3. **What is the methodology?** The report applies Simpson's Diversity Index to different religious categorizations (372 bodies, 215 bodies, 6 world religions, 43 families, 13 traditions, 13 traditions + unaffiliated).

4. **What geographic levels are analyzed?** Counties, metropolitan areas (1 million+ population), and states.

5. **What specific claims can you verify?** Focus on:
   - Counts of religious bodies in specific areas (e.g., "Los Angeles County has 171 religious bodies")
   - Percentage claims (e.g., "Catholics are 47% of Pennsylvania's religious adherents")
   - Rankings (e.g., Table 1's most/least diverse metros, Table 2's most/least diverse states)
   - Specific diversity index values or comparisons

## Your Analysis

Write your code below to reproduce key findings from the report. For each finding:
- State the claim from the report
- Show your code with clear comments
- Present your results
- Compare to what the report stated

### Finding 1: Religious Body Counts

Verify claims about the number of different religious bodies in specific areas.

**Claims to verify:**
- Los Angeles County has 171 religious bodies
- New York-Newark-Jersey City metro has 198 religious bodies
- Pennsylvania has 245 religious bodies

```{r}
# Your code here


```

### Finding 2: Religious Composition of Specific Areas

Verify percentage claims about religious adherents in specific areas.

**Claims to verify:**
- Catholics are 47% of Pennsylvania's religious adherents
- Catholics are 61% of the New York metro area's adherents
- Catholics are 62% of all adherents in Los Angeles County
- Two-thirds of Oregon, New Hampshire, and Maine population is unaffiliated

```{r}
# Your code here


```

### Finding 3: Metropolitan Area Diversity Rankings (Table 1)

Reproduce the rankings from Table 1, which shows the most and least religiously diverse metropolitan areas with 1 million or more population.

**Most diverse (rank 1-5):**
1. Memphis, TN-MS-AR
2. Charlotte-Concord-Gastonia, NC-SC
3. Richmond, VA
4. Milwaukee-Waukesha, WI
5. Cleveland-Elyria, OH

**Least diverse (rank 52-56):**
52. Tucson, AZ
53. Sacramento-Roseville-Folsom, CA
54. San Francisco-Oakland-Berkeley, CA
55. Portland-Vancouver-Hillsboro, OR-WA
56. Denver-Aurora-Lakewood, CO

```{r}
# Your code here
# You'll need to:
# 1. Identify the 56 metropolitan areas with 1 million+ population in 2020
# 2. Calculate Simpson's Diversity Index using 13 traditions + unaffiliated
# 3. Rank them and compare to the table


```

### Finding 4: Religious Composition of Specific Metropolitan Areas

Verify the composition claims for specific metros.

**Claims to verify:**
- Denver: two-thirds unaffiliated, nearly half the remaining is Catholic
- Salt Lake City: half claimed by Latter-day Saints, one-third unaffiliated
- Memphis: one-third unaffiliated, one-third Evangelical Protestant, one-sixth Black Protestant, Mainline Protestants and Catholics each ~5%

```{r}
# Your code here


```

### Finding 5: State Diversity Rankings (Table 2)

Reproduce the rankings from Table 2, which shows the most and least religiously diverse states.

**Most diverse (rank 1-5):**
1. Louisiana (37% unaffiliated, 27% Catholic, 25% Evangelical Protestant)
2. South Carolina (45% unaffiliated, 20% Evangelical Protestant, 18% Mainline Protestant)
3. Texas (45% unaffiliated, 24% Evangelical Protestant, 20% Catholic)
4. North Dakota (45% unaffiliated, 21% Catholic, 21% Mainline Protestant)
5. Idaho (47% unaffiliated, 25% Latter-day Saints, 12% Evangelical Protestant)

**Least diverse (rank 46-50):**
46. Alaska (65% unaffiliated)
47. Oregon (67% unaffiliated)
48. Utah (65% Latter-day Saints)
49. Maine (69% unaffiliated)
50. New Hampshire (73% unaffiliated)

```{r}
# Your code here


```

### Finding 6: District of Columbia Composition

Verify the claim that DC would be the most diverse if it were a state.

**Claims to verify:**
- DC: 44% unaffiliated
- Evangelical Protestants: 15%
- Catholics: 12%
- Black Protestants: 10%
- Mainline Protestants: 10%

```{r}
# Your code here


```

### Finding 7: Metropolitan Area Diversity Categories

Verify the claim about how many large metros fall into different diversity categories.

**Claims to verify:**
- Over one-third (20 of 56) of the nation's largest metro areas rank with the top 20% of counties
- Only 2 metros would rank with moderate diversity
- None would be categorized with the least diverse counties

```{r}
# Your code here
# You'll need to compare metro diversity indices to county diversity indices


```

### Finding 8: World Religions Diversity Comparison

Verify the claim about maximum diversity when using only world religions.

**Claims to verify:**
- Highest diversity based on world religions is only 0.51
- This compares to 0.62 needed to reach "moderate diversity" on the all religious bodies scale

```{r}
# Your code here
# You'll need to calculate diversity using 6 world religions and compare to the scale


```

## Additional Findings

Identify and reproduce other quantitative claims from the report.

```{r}
# Your analysis here


```

## Summary of Results

Summarize what you were able to reproduce and any discrepancies you found.

## Challenges and Limitations

Describe any challenges you encountered:
- What data was difficult to obtain or prepare?
- What assumptions did you have to make?
- What findings could not be reproduced and why?
- How did you handle metropolitan area definitions?
- How did you calculate the unaffiliated population?

## Data Sources

- 2020 U.S. Religion Census: https://www.usreligioncensus.org/
- Population data: U.S. Census Bureau
- Metropolitan area definitions: Office of Management and Budget (OMB)
