---
title: "Reverse Engineering: The Most Republican Jobs in America"
author: "Student Name"
date: "2026-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from VRScores, a database that links workplaces with voting records, to reproduce key findings from Ryan Burge's article: ["The Most Republican Jobs in America? Start with the Church."](https://www.graphsaboutreligion.com/p/the-most-republican-jobs-in-america) published on his Substack newsletter "Graphs about Religion" on December 22, 2025.

The article uses data from VRScores, which matched voter registration data (from a company called L2) with employment data (from Revelio Labs) to estimate the partisan composition of workers across more than 1,000 industries classified by NAICS codes.

A copy of the article has been saved as an HTML file in this folder.

**Important note:** The article uses 2024 data as the most recent year, but our dataset covers 2012-2020. We will use 2020 as our most recent year. Your exact numbers will differ from the article's, but the patterns and relative rankings should be similar.

## The Data

The data comes from VRScores and covers NAICS industry codes from 2012 through 2020. Each row represents one industry in one year.

Key variables include:
- `naics_code`: The NAICS industry classification code
- `naics_desc`: Text description of the industry
- `employee_count`: Number of workers in that industry-year
- `democrat_pct_imp` / `republican_pct_imp`: Imputed share of workers who are Democrats/Republicans
- `overall_margin_imp`: Republican share minus Democratic share (positive = Republican lean)
- `political_diversity_imp`: A diversity index (higher = more politically diverse)

## Load Libraries and Data

```{r}
library(tidyverse)

naics <- read_csv("data/naics_combined.csv")
glimpse(naics)
```

## Finding 1: Religious Organizations Are the Most Republican Among Large Industries

**Article claim:** Among large employment sectors, religious organizations (NAICS 813110) are the only group to the right of center, with Republicans outnumbering Democrats by about 7 percentage points. Higher education is the most left-leaning (about 32 points more Democratic).

```{r}
# Filter to 2020 data
naics_2020 <- naics %>% filter(year == 2020)

# Get the top 15 industries by employee count
top_15 <- naics_2020 %>%
  arrange(desc(employee_count)) %>%
  slice_head(n = 15) %>%
  select(naics_code, naics_desc, employee_count, democrat_pct_imp, republican_pct_imp, overall_margin_imp) %>%
  arrange(overall_margin_imp)

# Make sure religious organizations (813110) are included
religious_org <- naics_2020 %>% filter(naics_code == 813110)
cat("Religious Organizations (NAICS 813110):\n")
cat("  Employee count:", format(religious_org$employee_count, big.mark = ","), "\n")
cat("  Republican %:", round(religious_org$republican_pct_imp * 100, 1), "%\n")
cat("  Democrat %:", round(religious_org$democrat_pct_imp * 100, 1), "%\n")
cat("  Margin (R-D):", round(religious_org$overall_margin_imp * 100, 1), "points\n")

# Show all top 15 with their margins
top_15 %>%
  mutate(
    margin_pct = round(overall_margin_imp * 100, 1),
    rep_pct = round(republican_pct_imp * 100, 1),
    dem_pct = round(democrat_pct_imp * 100, 1)
  ) %>%
  select(naics_desc, employee_count, rep_pct, dem_pct, margin_pct) %>%
  arrange(desc(margin_pct))

# How many of the top 15 lean Republican (positive margin)?
cat("\nIndustries among top 15 that lean Republican:",
    sum(top_15$overall_margin_imp > 0), "\n")
```

**Your interpretation:** Is religious organizations the only large industry that leans Republican? How does it compare to higher education?

## Finding 2: Religious Organizations Have the Highest Republican Share Among the 50 Largest Industries

**Article claim:** Looking at the 50 largest industries, religious organizations have the highest Republican share -- just below 50%. The most Democratic-heavy large industries include internet services, clothing retail, media/streaming, and higher education.

```{r}
# Get the 50 largest industries in 2020
top_50 <- naics_2020 %>%
  arrange(desc(employee_count)) %>%
  slice_head(n = 50)

# If religious orgs aren't in the top 50, add them
if (!813110 %in% top_50$naics_code) {
  top_50 <- bind_rows(top_50, religious_org)
}

# Rank by Republican share
top_50_ranked <- top_50 %>%
  mutate(rep_pct = round(republican_pct_imp * 100, 1)) %>%
  arrange(desc(rep_pct)) %>%
  mutate(rank = row_number()) %>%
  select(rank, naics_desc, employee_count, rep_pct)

# Show top 10 most Republican
cat("Top 10 most Republican industries (among 50 largest):\n")
top_50_ranked %>% slice_head(n = 10)

# Show top 10 most Democratic (bottom of Republican ranking)
cat("\nTop 10 most Democratic industries (among 50 largest):\n")
top_50_ranked %>% slice_tail(n = 10)

# Where does religious organizations rank?
rel_rank <- top_50_ranked %>% filter(str_detect(naics_desc, "[Rr]eligious"))
cat("\nReligious Organizations rank:", rel_rank$rank, "of", nrow(top_50_ranked),
    "with", rel_rank$rep_pct, "% Republican\n")
```

**Your interpretation:** Do religious organizations have the highest Republican share among large industries? Is it close to 50%?

## Finding 3: Partisan Lean of Religious Organizations Over Time

**Article claim:** The article compares partisan compositions across time periods to show how industries' political leanings have shifted.

```{r}
# Compare religious organizations across all available years
religious_over_time <- naics %>%
  filter(naics_code == 813110) %>%
  select(year, employee_count, democrat_pct_imp, republican_pct_imp, overall_margin_imp) %>%
  mutate(
    dem_pct = round(democrat_pct_imp * 100, 1),
    rep_pct = round(republican_pct_imp * 100, 1),
    margin = round(overall_margin_imp * 100, 1)
  )

religious_over_time %>% select(year, employee_count, dem_pct, rep_pct, margin)

# How has the margin changed from 2012 to 2020?
margin_2012 <- religious_over_time$margin[religious_over_time$year == 2012]
margin_2020 <- religious_over_time$margin[religious_over_time$year == 2020]
cat("\nMargin change from 2012 to 2020:", margin_2020 - margin_2012, "points\n")
cat("2012 margin:", margin_2012, "\n")
cat("2020 margin:", margin_2020, "\n")
```

**Your interpretation:** Has the Republican lean of religious organizations grown, shrunk, or stayed stable over time?

## Finding 4: How Many Industries Lean Republican vs. Democratic?

**Article claim:** The article emphasizes that religious organizations are an outlier -- most large industries lean Democratic. Among the biggest employers, the church stands nearly alone on the right side of the partisan divide.

```{r}
# Among the top 50 largest industries in 2020, how many lean R vs D?
top_50_lean <- naics_2020 %>%
  arrange(desc(employee_count)) %>%
  slice_head(n = 50) %>%
  mutate(lean = case_when(
    overall_margin_imp > 0 ~ "Republican",
    overall_margin_imp < 0 ~ "Democratic",
    TRUE ~ "Even"
  ))

cat("Among the 50 largest industries in 2020:\n")
top_50_lean %>% count(lean)

# What percentage lean Republican?
n_rep <- sum(top_50_lean$lean == "Republican")
cat("\nRepublican-leaning:", n_rep, "of 50 (", round(n_rep / 50 * 100, 0), "%)\n")
cat("Democratic-leaning:", sum(top_50_lean$lean == "Democratic"), "of 50\n")

# What is the average margin among these top 50?
cat("\nAverage margin (top 50):", round(mean(top_50_lean$overall_margin_imp) * 100, 1), "points (negative = Dem lean)\n")
```

**Your interpretation:** Are most large industries Democratic-leaning? How unusual is it for religious organizations to lean Republican?

## Finding 5: Political Diversity of Religious Organizations vs. Other Industries

**Article claim:** The article notes that religious organizations are not only partisan but also have relatively low political diversity compared to many other industries.

```{r}
# Compare the political diversity of religious organizations to the average
avg_diversity <- naics_2020 %>%
  summarise(avg = mean(political_diversity_imp, na.rm = TRUE)) %>%
  pull(avg)

rel_diversity <- naics_2020 %>%
  filter(naics_code == 813110) %>%
  pull(political_diversity_imp)

cat("Religious Orgs political diversity score:", round(rel_diversity, 3), "\n")
cat("Average across all industries:", round(avg_diversity, 3), "\n")
cat("Religious Orgs vs. average:", ifelse(rel_diversity > avg_diversity, "More diverse", "Less diverse"), "\n")

# Where does religious organizations rank in political diversity among top 50?
top_50_diversity <- naics_2020 %>%
  arrange(desc(employee_count)) %>%
  slice_head(n = 50) %>%
  arrange(desc(political_diversity_imp)) %>%
  mutate(diversity_rank = row_number()) %>%
  select(naics_desc, political_diversity_imp, diversity_rank)

rel_div_rank <- top_50_diversity %>% filter(str_detect(naics_desc, "[Rr]eligious"))
cat("\nAmong top 50 industries, religious orgs rank", rel_div_rank$diversity_rank, "in political diversity\n")

# Show the 5 most and 5 least politically diverse large industries
cat("\n5 most politically diverse (among top 50):\n")
top_50_diversity %>% slice_head(n = 5)

cat("\n5 least politically diverse (among top 50):\n")
top_50_diversity %>% slice_tail(n = 5)
```

**Your interpretation:** Are religious organizations more or less politically diverse than the typical large industry? What does low political diversity mean in practice?

## Summary

Write a paragraph summarizing your findings. Were you able to replicate the article's key claims? What patterns held up in 2020 data that the article found in 2024 data? What limitations did you encounter?

## Data Sources

- VRScores: Kagan, Max; Frake, Justin; Hurst, Reuben (2025). "VRscores: A New Measure and Dataset of Workforce Politics Using Voter Registrations." SSRN Working Paper No. 5104795.
- Article: Burge, Ryan. "The Most Republican Jobs in America? Start with the Church." Graphs about Religion, December 22, 2025.
