---
title: "Reverse Engineering: Religion and Voting"
author: "Student Name"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from the Cooperative Election Study (CES) to reproduce key findings from Ryan Burge's article: ["Why Religion, Not Income, Predicts the American Vote"](https://www.graphsaboutreligion.com/p/why-religion-not-income-predicts).

The article argues that religious identity is a far stronger predictor of how Americans vote than household income. A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1.  **Read the article carefully** and identify the key facts, figures, and claims the author presents
2.  **Examine the CES data** provided in the `data/` folder
3.  **Write your own code** to reproduce those key findings
4.  **Document your work** with clear comments explaining your analysis

## The Data

The Cooperative Election Study (CES) surveys roughly 60,000 adults each year. The data file `ces_2024_religion.csv` contains:

-   **weight**: Survey weight (use this for all calculations)
-   **trad**: Broad religious tradition (e.g., "White Evangelical", "Jewish", "Atheist", "Black Protestant")
-   **income**: Household income on a 1-16 scale (1 = Less than \$10,000 ... 16 = \$500,000+)
-   **pid7**: 7-point party ID (1=Strong Democrat, 4=Independent, 7=Strong Republican)
-   **vote_2024**: 2024 presidential vote (1=Democrat/Harris, 2=Republican/Trump)
-   **race**: Race/ethnicity (1=White, 2=Black, 3=Hispanic, etc.)
-   **educ**: Education level (1=No HS through 6=Post-grad)

**Important:** Use `weighted.mean()` instead of `mean()` for all calculations.

## Load Libraries and Data

```{r}
library(tidyverse)

ces <- read_csv("data/ces_2024_religion.csv")

glimpse(ces)
```

## Finding 1: Average Income Varies Widely by Religious Tradition

**Article claim:** Hindus report the highest average income (\~\$85,000). Jewish traditions and Episcopalians average above \$75,000. Most Protestant traditions cluster around \$50,000-\$60,000.

```{r}
# Calculate weighted mean income category by religious tradition
# Note: the income variable is on a 1-16 scale, not dollar amounts
# To approximate dollar amounts, you'd need to assign midpoints to each category

income_by_trad <- ces %>%
  group_by(trad) %>%
  summarise(
    n = n(),
    avg_income_cat = round(weighted.mean(income, weight, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  filter(n >= 250) %>%
  arrange(desc(avg_income_cat))

income_by_trad
```

**Your interpretation:** Which religious traditions have the highest and lowest average income? How large are the gaps?

## Finding 2: Republican Identification Does Not Track with Income

**Article claim:** Reform Jews have fewer than 20% Republican identification despite having some of the highest incomes. About 25% of Hindus identify as Republican. Nearly 75% of non-denominational Protestants are Republican despite having moderate incomes.

```{r}
# Calculate the percentage who identify as Republican (pid7 >= 5) by tradition
republican_by_trad <- ces %>%
  mutate(is_republican = pid7 >= 5) %>%
  group_by(trad) %>%
  summarise(
    n = n(),
    rep_pct = round(weighted.mean(is_republican, weight, na.rm = TRUE) * 100, 1),
    avg_income_cat = round(weighted.mean(income, weight, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  filter(n >= 250) %>%
  arrange(desc(rep_pct))

republican_by_trad

# If income drove partisanship, high-income groups should be more Republican
# Check the correlation
cat("\nCorrelation between avg income and Republican %:",
    round(cor(republican_by_trad$avg_income_cat, republican_by_trad$rep_pct), 2), "\n")
```

**Your interpretation:** Is there a strong relationship between income and Republican identification? Or does religion override income?

## Finding 3: Trump Vote Among White Evangelicals by Income

**Article claim:** Among White Evangelicals making under \$40,000, about two-thirds (\~67%) voted for Trump. Among those earning \$80,000-\$120,000, about 78% voted for Trump -- only a \~10 percentage point increase across a large income range.

```{r}
# Calculate Trump vote percentage by income bracket among White Evangelicals
# Income categories: 1-4 roughly correspond to < $40k, 8-10 roughly to $80k-$120k

sbc_by_income <- ces %>%
  filter(!is.na(vote_2024)) %>%
  mutate(
    voted_trump = vote_2024 == 2,
    income_bracket = case_when(
      income <= 4 ~ "Under $40k",
      income <= 7 ~ "$40k-$70k",
      income <= 10 ~ "$70k-$120k",
      income <= 13 ~ "$120k-$250k",
      income >= 14 ~ "$250k+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_bracket)) %>%
  group_by(trad, income_bracket) %>%
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

sbc_by_income

# What is the range from lowest to highest income bracket?
cat("\nRange of Trump support across income brackets:",
    max(sbc_by_income$trump_pct) - min(sbc_by_income$trump_pct), "percentage points\n")
```

**Your interpretation:** Does income make much difference in how White Evangelicals vote? Is the gap really only about 10 points?

## Finding 4: The 60-Point Gap Between White Evangelicals and Atheists at Same Income

**Article claim:** A White Evangelical and an atheist with the exact same income are separated by about 60 percentage points in their likelihood of voting Republican.

```{r}
# Compare Trump vote by income bracket for two very different religious traditions
gap_analysis <- ces %>%
  filter(trad %in% c("White Evangelical", "Atheist"), !is.na(vote_2024)) %>%
  mutate(
    voted_trump = vote_2024 == 2,
    income_bracket = case_when(
      income <= 5 ~ "Under $50k",
      income <= 9 ~ "$50k-$100k",
      income >= 10 ~ "$100k+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_bracket)) %>%
  group_by(trad, income_bracket) %>%
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = trad, values_from = c(trump_pct, n))

gap_analysis

# Calculate the gap at each income level
gap_analysis %>%
  mutate(gap = `trump_pct_White Evangelical` - trump_pct_Atheist)
```

**Your interpretation:** Is the religion gap really about 60 points even when comparing people at the same income level?

## Finding 5: White Evangelicals vs. Black Protestants - Maximum Religion Gap

**Article claim:** White evangelicals give Trump over 80% support while Black Protestants give him under 10% -- a gap of over 70 percentage points that persists even when comparing people with the same household income.

```{r}
# Compare the two most extreme religious traditions
extremes <- ces %>%
  filter(trad %in% c("White Evangelical", "Black Protestant"), !is.na(vote_2024)) %>%
  mutate(voted_trump = vote_2024 == 2) %>%
  group_by(trad) %>%
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

extremes

cat("\nGap:", extremes$trump_pct[extremes$trad == "White Evangelical"] -
    extremes$trump_pct[extremes$trad == "Black Protestant"], "percentage points\n")

# Now break it down by income to show the gap persists
extremes_by_income <- ces %>%
  filter(trad %in% c("White Evangelical", "Black Protestant"), !is.na(vote_2024)) %>%
  mutate(
    voted_trump = vote_2024 == 2,
    income_bracket = case_when(
      income <= 5 ~ "Under $50k",
      income <= 9 ~ "$50k-$100k",
      income >= 10 ~ "$100k+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_bracket)) %>%
  group_by(trad, income_bracket) %>%
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

extremes_by_income
```

**Your interpretation:** Does the gap between White evangelicals and Black Protestants persist at every income level? What does this tell us about religion vs. income as predictors of voting?

## Finding 6: Income Barely Moves the Needle Within Religious Traditions

**Article claim:** "For Latter-day Saints, the difference between the highest earners and the lowest is one percentage point. For white evangelicals, it's 0.6%. For Black Protestants, just 0.5%."

```{r}
# Run a logistic regression predicting Trump vote from income and religious tradition
df <- ces %>%
  filter(!is.na(vote_2024)) %>%
  mutate(voted_trump = as.integer(vote_2024 == 2))

model <- glm(voted_trump ~ income + trad, data = df, family = binomial, weights = weight)

# For each tradition, calculate the predicted probability of voting Trump
# at the lowest income (1 = less than $10k) and highest income (16 = $500k+)
traditions_to_check <- c("Mormon/LDS", "White Evangelical", "Black Protestant")

income_effect <- map_dfr(traditions_to_check, function(t) {
  pred_low <- predict(model, newdata = data.frame(income = 1, trad = t), type = "response")
  pred_high <- predict(model, newdata = data.frame(income = 16, trad = t), type = "response")

  tibble(
    trad = t,
    pred_lowest_income = round(pred_low * 100, 1),
    pred_highest_income = round(pred_high * 100, 1),
    difference = round((pred_high - pred_low) * 100, 1)
  )
})

income_effect
```

**Your interpretation:** How much does income change the predicted Trump vote within these traditions? Does this support the article's claim that religion dominates income as a predictor?

For comparison, here's what the raw weighted means look like at the income extremes — note the very small sample sizes, which is why the regression approach above is more reliable:

```{r}
# Weighted mean Trump vote % at lowest and highest income levels, by tradition
ces %>%
  filter(!is.na(vote_2024), trad %in% c("Mormon/LDS", "White Evangelical", "Black Protestant")) %>%
  mutate(
    voted_trump = as.integer(vote_2024 == 2),
    income_group = case_when(
      income == 1 ~ "Lowest ($<10k)",
      income == 16 ~ "Highest ($500k+)",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_group)) %>%
  group_by(trad, income_group) %>%
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = income_group, values_from = c(trump_pct, n)) %>%
  mutate(diff = `trump_pct_Highest ($500k+)` - `trump_pct_Lowest ($<10k)`)
```

## Summary

Write a summary of your findings. Did your analysis confirm the article's central claim that religion is a stronger predictor of voting than income?
