---
title: "Reverse Engineering: Religion and Voting"
author: "Student Name"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from the Cooperative Election Study (CES) to reproduce key findings from Ryan Burge's article: ["Why Religion, Not Income, Predicts the American Vote"](https://www.graphsaboutreligion.com/p/why-religion-not-income-predicts).

The article argues that religious identity is a far stronger predictor of how Americans vote than household income. A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1.  **Read the article carefully** and identify the key facts, figures, and claims the author presents
2.  **Examine the CES data** provided in the `data/` folder
3.  **Write your own code** to reproduce those key findings
4.  **Document your work** with clear comments explaining your analysis

## The Data

The Cooperative Election Study (CES) surveys roughly 60,000 adults each year. The data file `ces_2024_religion.csv` contains:

-   **weight**: Survey weight (use this for all calculations)
-   **trad**: Broad religious tradition (e.g., "White Evangelical", "Jewish", "Atheist", "Black Protestant")
-   **income**: Household income on a 1-16 scale (1 = Less than \$10,000 ... 16 = \$500,000+)
-   **pid7**: 7-point party ID (1=Strong Democrat, 4=Independent, 7=Strong Republican)
-   **vote_2024**: 2024 presidential vote (1=Democrat/Harris, 2=Republican/Trump)
-   **race**: Race/ethnicity (1=White, 2=Black, 3=Hispanic, etc.)
-   **educ**: Education level (1=No HS through 6=Post-grad)

**Important:** Use `weighted.mean()` instead of `mean()` for all calculations.

## Load Libraries and Data

```{r}
library(tidyverse)

ces <- read_csv("data/ces_2024_religion.csv")

glimpse(ces)
```

## Finding 1: income barely budges the needle within traditions, but traditions themselves are separated by enormous political gaps,

```{r}
# Create income groups matching the article's analysis
ces <- ces |>
  mutate(
    income_group = case_when(
      income <= 4 ~ "Under $40K",
      income <= 8 ~ "$40K-$80K",
      income <= 11 ~ "$80K-$120K",
      income >= 12 ~ "$120K+"
    ),
    trump_vote = case_when(
      vote_2024 == 2 ~ 1,
      vote_2024 == 1 ~ 0
    )
  )

# Calculate weighted Trump vote share by religious tradition and income group
tradition_income <- ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  group_by(trad, income_group) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

# Show the results - within each tradition, income makes little difference
tradition_income |>
  pivot_wider(names_from = income_group, values_from = c(trump_pct, n)) |>
  arrange(desc(`trump_pct_Under $40K`))
```

**Your interpretation:** 

## Finding 2: A Southern Baptist and an atheist with the exact same income are separated by sixty percentage points in their likelihood of voting Republican."


```{r}
# Southern Baptists are identified by baptist == 1 (SBC code) in the data
# Compare Trump vote share for Southern Baptists vs Atheists at each income level
sbc_vs_atheist <- ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  filter((baptist == 1 & trad == "White Evangelical") | trad == "Atheist") |>
  mutate(group = if_else(trad == "Atheist", "Atheist", "Southern Baptist")) |>
  group_by(group, income_group) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

sbc_vs_atheist |>
  pivot_wider(names_from = group, values_from = c(trump_pct, n))

# Overall comparison
ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  filter((baptist == 1 & trad == "White Evangelical") | trad == "Atheist") |>
  mutate(group = if_else(trad == "Atheist", "Atheist", "Southern Baptist")) |>
  group_by(group) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )
```

Now using logistic regression, which is what the article actually did — predicting Trump vote from income + religious tradition, then comparing predicted probabilities at the same income level:

```{r}
# Logistic regression: predict Trump vote from income + religious tradition
model <- glm(trump_vote ~ income + trad, data = ces, weights = weight, family = binomial)

# Create prediction data: Southern Baptists and Atheists at each income level
# We use "White Evangelical" as the trad for Southern Baptists since the model uses trad
prediction_data <- expand.grid(
  income = 1:16,
  trad = c("White Evangelical", "Atheist")
)

prediction_data$predicted_prob <- predict(model, newdata = prediction_data, type = "response") * 100

# Show predicted Trump vote at median income (8) for both groups
prediction_data |>
  filter(income == 8) |>
  mutate(group = if_else(trad == "Atheist", "Atheist", "Southern Baptist (White Evangelical)")) |>
  select(group, income, predicted_prob)

# Calculate the gap at each income level
prediction_data |>
  pivot_wider(names_from = trad, values_from = predicted_prob) |>
  mutate(gap = `White Evangelical` - Atheist)
```

**Your interpretation:**

## Finding 3: Trump pulled support from over 80% of white evangelicals and under 10% of Black Protestants

```{r}
# Calculate overall weighted Trump vote share for White Evangelicals and Black Protestants
ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  filter(trad %in% c("White Evangelical", "Black Protestant")) |>
  group_by(trad) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

# Break it down by income group to show the pattern holds across income levels
ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  filter(trad %in% c("White Evangelical", "Black Protestant")) |>
  group_by(trad, income_group) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = trad, values_from = c(trump_pct, n))
```

Now using the regression model to verify this holds "even when comparing respondents with the same household income":

```{r}
# Predicted Trump vote probability for White Evangelicals and Black Protestants across all income levels
prediction_data_f3 <- expand.grid(
  income = 1:16,
  trad = c("White Evangelical", "Black Protestant")
)

prediction_data_f3$predicted_prob <- predict(model, newdata = prediction_data_f3, type = "response") * 100

# Summarize the range of predicted probabilities for each tradition across all income levels
# If the article is right, White Evangelical should stay above 80% and Black Protestant below 10%
prediction_data_f3 |>
  group_by(trad) |>
  summarise(
    min_predicted = min(predicted_prob),
    max_predicted = max(predicted_prob),
    median_predicted = median(predicted_prob),
    .groups = "drop"
  )
```

**Your interpretation:**

## Finding 4: For Latter-day Saints, the difference between the highest earners and the lowest is one percentage point. For white evangelicals, it’s 0.6%. For Black Protestants, just 0.5%.


```{r}
# Compare Trump vote share at the extremes of the income scale for key traditions
# income == 1 is "Less than $10K" and income == 16 is "$500K+"
extreme_income <- ces |>
  filter(!is.na(trump_vote), !is.na(weight)) |>
  filter(trad %in% c("Mormon/LDS", "White Evangelical", "Black Protestant")) |>
  mutate(income_extreme = case_when(
    income == 1 ~ "Lowest (Under $10K)",
    income == 16 ~ "Highest ($500K+)"
  )) |>
  filter(!is.na(income_extreme)) |>
  group_by(trad, income_extreme) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

extreme_income |>
  pivot_wider(names_from = income_extreme, values_from = c(trump_pct, n))

# Calculate the gap between highest and lowest earners
extreme_income |>
  select(trad, income_extreme, trump_pct) |>
  pivot_wider(names_from = income_extreme, values_from = trump_pct) |>
  mutate(gap = `Highest ($500K+)` - `Lowest (Under $10K)`)
```

Now using the regression model — the article explicitly says these gaps come from "predicted probabilities" at the top vs bottom of the income distribution:

```{r}
# Predicted probabilities at income extremes (1 = under $10K, 16 = $500K+)
prediction_data_f4 <- expand.grid(
  income = c(1, 16),
  trad = c("Mormon/LDS", "White Evangelical", "Black Protestant")
)

prediction_data_f4$predicted_prob <- predict(model, newdata = prediction_data_f4, type = "response") * 100

# Calculate the gap between highest and lowest income for each tradition
prediction_data_f4 |>
  mutate(income_label = if_else(income == 1, "Lowest", "Highest")) |>
  select(trad, income_label, predicted_prob) |>
  pivot_wider(names_from = income_label, values_from = predicted_prob) |>
  mutate(gap = Highest - Lowest)
```

## Finding 5: Take Southern Baptists. Among those making under $40,000 a year, about two-thirds voted for Donald Trump. Among those earning between $80,000 and $120,000, that number rises to about 78%.


```{r}
# Southern Baptists (baptist == 1) Trump vote by income group
ces |>
  filter(!is.na(trump_vote), !is.na(weight), !is.na(baptist)) |>
  filter(baptist == 1) |>
  group_by(income_group) |>
  summarise(
    trump_pct = weighted.mean(trump_vote, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )
```

**Your interpretation:** 
