---
title: "Reverse Engineering: Changing Faiths"
author: "Student Name"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## About this assignment

In this assignment, you will reverse engineer findings from a FiveThirtyEight analysis by Leah Libresco: ["Evangelical Protestants Are The Biggest Winners When People Change Faiths."](http://fivethirtyeight.com/datalab/evangelical-protestants-are-the-biggest-winners-when-people-change-faiths/) The article uses data from Pew Research Center's 2014 Religious Landscape Study to model what would happen to America's religious makeup if people kept switching faiths at their current rates. A copy of the article has been saved as an HTML file in this folder for offline reference.

The original code and data are from FiveThirtyEight's GitHub repository: <https://github.com/fivethirtyeight/data/tree/master/pew-religions>

Your job is to read the article and the original Python code, then write your own R code to replicate specific findings. The original analysis was written in Python using NumPy matrix operations. You will reproduce the key results using R and the tidyverse.

To complete this assignment, you need to: \* run existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## About the data

The data comes from Pew's 2014 Religious Landscape Study, which surveyed more than 35,000 Americans about their religious beliefs and practices. For this analysis, the key question is: what religion were you raised in, and what religion are you now?

From those responses, the original analysis built a **transition matrix** — a table where each row is the religion someone was raised in, and each column is the religion they hold as an adult. The values are probabilities. For example, if the row is "Catholic" and the column is "Evangelical Protestant," the value tells you the probability that someone raised Catholic is now an evangelical Protestant.

You have four data files in the `data/` folder:

| File | Description |
|------|-------------|
| `transition_matrix.csv` | 12x12 matrix of switching probabilities. Each row is a "raised in" religion; each column is a "current" religion. Each row sums to approximately 1.0. |
| `initial_distribution.csv` | The starting share of each religion in the U.S. population (sums to ~1.0). These are roughly the shares of Americans who were *raised in* each tradition. |
| `fertility_rates.csv` | Average number of children per woman for each religious group. |
| `current.csv` | The output of the original Python simulation: 101 rows showing the religious distribution at each "year" (iteration 0 through 100) as the model approaches equilibrium. This is what you'll be checking your work against. |

The 12 religious categories are: Buddhist, Catholic, Evangelical Protestant, Hindu, Historically Black Protestant, Jehovah's Witness, Jewish, Mainline Protestant, Mormon, Muslim, Orthodox Christian, and Unaffiliated.

## Load libraries and data

**Task** Run the codeblock below to load the tidyverse and all four data files.

```{r}
library(tidyverse)

transition <- read_csv("data/transition_matrix.csv")
initial <- read_csv("data/initial_distribution.csv")
fertility <- read_csv("data/fertility_rates.csv")
simulation <- read_csv("data/current.csv",
                       col_names = c("Buddhist", "Catholic", "Evangelical Protestant",
                                     "Hindu", "Historically Black Protestant",
                                     "Jehovahs Witness", "Jewish", "Mainline Protestant",
                                     "Mormon", "Muslim", "Orthodox Christian", "Unaffiliated"))
```

**Task** Run the codeblocks below to examine the data.

```{r}
# The transition matrix: each row is a "raised in" religion, each column is where they ended up
transition
```

```{r}
# The starting distribution
initial
```

```{r}
# Fertility rates
fertility
```

```{r}
# First and last rows of the simulation output
head(simulation, 1)
tail(simulation, 1)
```

## Questions

Your task is to replicate five specific findings from the article by examining the data. For each question, find the relevant numbers in the datasets and verify whether they match what the article reports.

### Q1: Retention rates

The article reports: "Hindus, Mormons and Jews have the highest retention rates, at 80%, 76% and 75% respectively. Evangelical Protestants have a retention rate of 65%, while Catholics have a rate of 59%. Mainline Protestants have the lowest rate, at 45%."

The retention rate for each religion is the value in the transition matrix where the row and column are the same religion (the diagonal). For example, to find the retention rate for Catholics, look at the row for Catholic and the column for Catholic.

Find the retention rates for all 12 religions in the transition matrix. Do your numbers match what the article reports? Write a sentence reporting the retention rate for Hindus, Mormons, Jews, Evangelical Protestants, Catholics, and Mainline Protestants.

**A1.**

```{r}
```

### Q2: Catholic to Evangelical Protestant conversions

The article states: "About 10 percent of people raised Catholic now identify as evangelical Protestants ... while only 2 percent of people raised as evangelicals are now Catholic."

In the transition matrix, find the value in the Catholic row and the Evangelical Protestant column (this shows what percent of people raised Catholic are now Evangelical Protestant). Then find the value in the Evangelical Protestant row and the Catholic column.

Do these numbers match the article's claim? Write a sentence stating what you found.

**A2.**

```{r}
```

### Q3: Mainline to Evangelical Protestant conversions

The article reports: "About 19 percent of people raised as mainline Protestants now identify as evangelical, while only 12 percent of evangelicals have become mainline Protestants."

Using the transition matrix, find these two conversion rates. Do they match what the article says? Write a sentence stating what you found.

**A3.**

```{r}
```

### Q4: Starting religious distribution

The article describes what share of Americans were raised in each religious tradition. Look at the `initial_distribution.csv` file, which shows these starting percentages.

According to this data, what are the three largest religious traditions people were raised in? What percent of Americans were raised Catholic? What percent were raised as Evangelical Protestants? What percent were raised Unaffiliated (with no religion)? Write 2-3 sentences describing the starting distribution.

**A4.**

```{r}
```

### Q5: Final distribution with fertility rates

The article's model projects what would happen if current conversion patterns and fertility rates continued for many generations. The file `current.csv` contains this simulation output. The last row (row 101) shows the projected equilibrium.

The article states: "Mormons... would eventually reach 22% of the population" and "Muslims would reach 8%" while "the unaffiliated would actually shrink from 23% to 19%."

Look at the last row of the simulation data. What percentages does it show for Mormons, Muslims, and Unaffiliated? Do these match the article's claims? Also identify which religious group ends up with the largest share. Write 2-3 sentences describing what you found.

**A5.**

```{r}
```

