---
title: "Reverse Engineering: When Are Half Your Members Going to be Dead?"
author: "Student Name"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from the 2024 Cooperative Election Study (CCES) to reproduce key findings from Ryan Burge's article: ["When Are Half Your Members Going to be Dead?"](https://www.graphsaboutreligion.com/p/when-are-half-your-members-going) published on his Substack newsletter "Graphs about Religion."

The article examines the age distribution of the 20 largest denominations in the Cooperative Election Study, calculating mean and modal ages, Baby Boomer shares, and projections for when each denomination will lose significant portions of its current membership.

A copy of the article has been saved as an HTML file in this folder.

**Important note:** The article combines 2022 and 2024 CCES data, but our dataset only contains 2024 responses. Your exact numbers may differ slightly from the article's, but the patterns and relative rankings should be similar.

## The Data

The data comes from the 2024 Cooperative Election Study and contains 60,000 respondents.

Key variables: - `year`: Survey year (2024 for all rows) - `birthyr`: Year of birth - `weight`: Survey weight for the respondent - `baptist`, `methodist`, `nondenom`, `lutheran`, `presbyterian`, `pentecostal`, `episcopal`, `christian`, `congregational`, `holiness`, `reformed`, `adventist`: Denomination sub-category codes (most are NA)

## Load Libraries and Data

```{r}
library(tidyverse)

cces <- read_csv("cces24.csv")
glimpse(cces)
```

## Create the Denomination Variable

```{r}
cces <- cces %>%
  mutate(
    age = 2024 - birthyr,
    trad = case_when(
      baptist == 1 ~ "Southern Baptist",
      baptist == 2 ~ "ABCUSA",
      baptist == 3 ~ "Natl. Baptist Convention",
      baptist == 5 ~ "Independent Baptist",
      baptist == 9 ~ "Free Will Baptist",
      baptist == 90 ~ "Other Baptist",
      methodist == 1 ~ "United Methodist",
      methodist == 3 ~ "African Methodist Episcopal",
      methodist == 90 ~ "Other Methodist",
      nondenom == 1 ~ "Nondenom Evangelical",
      nondenom == 3 ~ "Nondenom Charismatic",
      nondenom == 4 ~ "Interdenominational",
      nondenom == 5 ~ "Community Church",
      nondenom == 90 ~ "Other Nondenom",
      lutheran == 1 ~ "ELCA",
      lutheran == 2 ~ "Lutheran - Missouri Synod",
      presbyterian == 1 ~ "PCUSA",
      presbyterian == 2 ~ "PCA",
      pentecostal == 1 ~ "Assemblies of God",
      pentecostal == 6 ~ "Church of God in Christ",
      episcopal == 1 ~ "TEC",
      christian == 1 ~ "Church of Christ",
      christian == 2 ~ "Disciples of Christ",
      congregational == 1 ~ "United Church of Christ",
      holiness == 1 ~ "Church of the Nazarene",
      adventist == 1 ~ "Seventh Day Adventist",
      TRUE ~ NA_character_
    )
  )

# Find the top 20 denominations by count, excluding "Other" categories
top20 <- cces %>%
  filter(!is.na(trad)) %>%
  filter(!str_detect(trad, "^Other")) %>%
  count(trad) %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%
  pull(trad)

cat("Top 20 denominations:\n")
top20
```

## Finding 1: Mean Age by Denomination

**Article claim:** The average adult member of an Episcopal church (TEC) is 61 years old, Southern Baptists average 58, and the Assemblies of God average 54. Most denominations have mean ages of at least 55.

```{r}
# Calculate weighted mean age for each of the top 20 denominations
mean_ages <- cces %>%
  filter(trad %in% top20) %>%
  group_by(trad) %>%
  summarise(
    n = n(),
    mean_age = round(weighted.mean(age, weight, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_age))

mean_ages

# Key comparisons from the article
cat("\nKey mean ages:\n")
for (denom in c("TEC", "Southern Baptist", "Assemblies of God", "United Methodist", "ELCA", "PCUSA")) {
  val <- mean_ages %>% filter(trad == denom) %>% pull(mean_age)
  if (length(val) > 0) cat(denom, ":", val, "\n")
}

# How many have mean ages of at least 55?
cat("\nDenominations with mean age >= 55:", sum(mean_ages$mean_age >= 55), "of", nrow(mean_ages), "\n")
```

**Your interpretation:** Does the data confirm the mean ages reported in the article? Are most denominations at 55+?

## Finding 2: Modal Age by Denomination

**Article claim:** In many mainline Protestant denominations, the modal age is between 67 and 69. The Assemblies of God has a higher modal age (69) than United Methodists (67), yet the AG's mean age is about 7 years younger (54 vs. 61).

```{r}
# Calculate the modal (most common) age for each denomination
# Mode = the age with the highest count (unweighted for simplicity)
mode_ages <- cces %>%
  filter(trad %in% top20) %>%
  group_by(trad, age) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(trad) %>%
  filter(n == max(n)) %>%
  summarise(modal_age = min(age), .groups = "drop")  # min in case of ties

# Join with mean ages
age_comparison <- mean_ages %>%
  left_join(mode_ages, by = "trad") %>%
  mutate(mean_mode_gap = modal_age - mean_age) %>%
  arrange(desc(modal_age))

age_comparison

# Highlight the AG vs UMC comparison
cat("\nAssemblies of God vs. United Methodist:\n")
ag <- age_comparison %>% filter(trad == "Assemblies of God")
umc <- age_comparison %>% filter(trad == "United Methodist")
if (nrow(ag) > 0 & nrow(umc) > 0) {
  cat("AG  - Mean:", ag$mean_age, ", Mode:", ag$modal_age, ", Gap:", ag$mean_mode_gap, "\n")
  cat("UMC - Mean:", umc$mean_age, ", Mode:", umc$modal_age, ", Gap:", umc$mean_mode_gap, "\n")
  cat("Mean age difference:", umc$mean_age - ag$mean_age, "years (UMC older)\n")
}
```

**Your interpretation:** Does the AG really have a higher modal age but lower mean age than the UMC? What does this tell us about the shape of their age distributions?

## Finding 3: Share of Baby Boomers by Denomination

**Article claim:** About half of United Methodists and Episcopalians were born during the Baby Boomer window (1946-1964). For 16 of the 20 denominations, Boomers make up at least 40% of their adult membership. In the general U.S. adult population, Baby Boomers make up about 20%.

```{r}
# Calculate the weighted share of Baby Boomers (born 1946-1964) in each denomination
boomer_shares <- cces %>%
  filter(trad %in% top20) %>%
  mutate(is_boomer = birthyr >= 1946 & birthyr <= 1964) %>%
  group_by(trad) %>%
  summarise(
    boomer_pct = round(weighted.mean(is_boomer, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(boomer_pct))

boomer_shares

# How many denominations have at least 40% Boomers?
cat("\nDenominations with >= 40% Boomers:", sum(boomer_shares$boomer_pct >= 40), "of", nrow(boomer_shares), "\n")

# What is the Boomer share among ALL respondents (general population)?
general_boomer <- cces %>%
  mutate(is_boomer = birthyr >= 1946 & birthyr <= 1964) %>%
  summarise(boomer_pct = round(weighted.mean(is_boomer, weight, na.rm = TRUE) * 100, 1))
cat("General population Boomer share:", general_boomer$boomer_pct, "%\n")

# Key denominations
cat("\nUnited Methodist Boomer share:", boomer_shares$boomer_pct[boomer_shares$trad == "United Methodist"], "%\n")
if ("TEC" %in% boomer_shares$trad) {
  cat("TEC Boomer share:", boomer_shares$boomer_pct[boomer_shares$trad == "TEC"], "%\n")
}
```

**Your interpretation:** Does the data confirm that roughly half of mainline Protestants are Boomers? Are Boomers about twice as represented in these denominations compared to the general population?

## Finding 4: Share of Young Adults (18-29) and Episcopalian Age Skew

**Article claim:** Just 6% of Episcopalians and 6% of Southern Baptists are under the age of 30. For every young adult Episcopalian, there are about ten retirees. The Assemblies of God has 11% of members aged 18-29.

```{r}
# Calculate the percentage of members under 30 for each denomination
young_adult_shares <- cces %>%
  filter(trad %in% top20) %>%
  mutate(
    under_30 = age < 30,
    over_60 = age >= 60
  ) %>%
  group_by(trad) %>%
  summarise(
    under_30_pct = round(weighted.mean(under_30, weight, na.rm = TRUE) * 100, 1),
    over_60_pct = round(weighted.mean(over_60, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(retiree_to_young_ratio = round(over_60_pct / under_30_pct, 1)) %>%
  arrange(under_30_pct)

young_adult_shares

# Key comparisons
cat("\nKey under-30 shares:\n")
for (denom in c("TEC", "Southern Baptist", "Assemblies of God", "United Methodist")) {
  val <- young_adult_shares %>% filter(trad == denom)
  if (nrow(val) > 0) cat(denom, ": ", val$under_30_pct, "% under 30, ratio of retirees to young:", val$retiree_to_young_ratio, "\n")
}
```

**Your interpretation:** Does the data confirm just 6% of Episcopalians and Southern Baptists are under 30? Is the retiree-to-young-adult ratio really about 10:1 for Episcopalians?

## Finding 5: Percentage of Members Already Over 75

**Article claim:** The article highlights that some denominations already have a significant share of members who are 75 or older, meaning a substantial portion of their membership will be gone within a decade. Mainline denominations like TEC, PCUSA, and ELCA have the highest shares of very old members.

```{r}
# Calculate the percentage of members aged 75+ for each denomination
over_75 <- cces %>%
  filter(trad %in% top20) %>%
  mutate(is_over_75 = age >= 75) %>%
  group_by(trad) %>%
  summarise(
    over_75_pct = round(weighted.mean(is_over_75, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(over_75_pct))

over_75

# Compare to the general population
general_over_75 <- cces %>%
  mutate(is_over_75 = age >= 75) %>%
  summarise(over_75_pct = round(weighted.mean(is_over_75, weight, na.rm = TRUE) * 100, 1))
cat("\nGeneral population aged 75+:", general_over_75$over_75_pct, "%\n")

# Which denominations have more than double the general population share?
threshold <- general_over_75$over_75_pct * 2
cat("Denominations with 75+ share more than double the general population (",
    threshold, "%):\n")
over_75 %>% filter(over_75_pct > threshold)

# Assemblies of God vs United Methodist comparison
cat("\nAssemblies of God 75+ share:", over_75$over_75_pct[over_75$trad == "Assemblies of God"], "%\n")
cat("United Methodist 75+ share:", over_75$over_75_pct[over_75$trad == "United Methodist"], "%\n")
```

**Your interpretation:** Which denominations have the highest share of members aged 75+? How does this compare to the general population? What does this tell us about which denominations face the most urgent demographic challenges?

## Summary

Write a paragraph summarizing your findings. Were you able to replicate the article's key claims? What patterns held up? What limitations did you encounter with using only 2024 data instead of the combined 2022-2024 data?

## Data Sources

-   Cooperative Election Study, 2024
-   Article: Burge, Ryan. "When Are Half Your Members Going to be Dead?" Graphs about Religion.
