---
title: "Reverse Engineering: Non-Christian Faith Dispersion"
author: "Student Name"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze Mike McMullen's report "The Rise and Geographic Dispersion of Non-Christian Faith Traditions in the United States" and reproduce key quantitative findings using data from the 2020 U.S. Religion Census.

**Your task:**

1.  Read the report carefully (PDF file in this folder)
2.  Identify key facts and statistics that were produced through data analysis
3.  Write well-commented, well-structured R code to reproduce those findings
4.  Focus on replicable quantitative sentences (not the maps)

**Important:** The data files from the 2020 U.S. Religion Census are in the `data/` folder.

## The Data

The report analyzes seven non-Christian faith traditions: Baha'i Faith, Buddhism, Hinduism, Islam, Judaism, Sikhism, and Zoroastrianism. It examines their geographic dispersion across U.S. metropolitan areas, classifying metro areas by "gateway city" type.

**Key definitions:**

- **High-concentration area:** A metropolitan area with at least 6 houses of worship for a given faith tradition
- **Gateway city types** (from Singer et al., 2008): Former, Continuous, Post-WWII, Emerging, Re-emerging, Pre-emerging

## Load Libraries and Data

```{r}
library(tidyverse)
library(readxl)
```

```{r}
# Load the 2020 U.S. Religion Census data
# Group Detail has individual religious group data by geographic level
# Summaries has aggregate totals by geographic level
groups_nation <- read_excel("data/2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by Nation")
groups_metro  <- read_excel("data/2020_USRC_Group_Detail.xlsx", sheet = "2020 Group by Metro")

glimpse(groups_metro)
```

```{r}
# The report examines 7 non-Christian faith traditions.
# We need to map Group Codes from the Religion Census to each tradition.
#
# Key group codes:
# Baha'i:   056 (Baha'i Faith USA)
# Buddhist: 890 (Mahayana), 891 (Theravada), 892 (Vajrayana)
# Hindu:    895 (Hindu Temples) -- the report focuses on temples specifically
# Muslim:   267 (Muslim Estimate)
# Jewish:   270 (Conservative), 271 (Reform), 275 (Reconstructionist),
#           276 (Orthodox), FGR (Independent), FGS (Chabad)
# Sikh:     416 (American Sikh Council)
# Zoroastrian: 490

faith_mapping <- tribble(
  ~group_code, ~faith,
  "056", "Bahai",
  "890", "Buddhist",
  "891", "Buddhist",
  "892", "Buddhist",
  "895", "Hindu",
  "267", "Muslim",
  "270", "Jewish",
  "271", "Jewish",
  "275", "Jewish",
  "276", "Jewish",
  "FGR", "Jewish",
  "FGS", "Jewish",
  "416", "Sikh",
  "490", "Zoroastrian"
)
```

## Define Gateway City Classifications

```{r}
# Gateway city classifications from Singer et al. (2008)
# These patterns match the metro area names in the Religion Census data.
# Established gateways: Former, Continuous, Post-WWII
# 21st-century gateways: Emerging, Re-emerging, Pre-emerging

gateway_cities <- tribble(
  ~pattern, ~gateway_type,
  "Baltimore-Columbia", "Former",
  "Buffalo-Cheektowaga", "Former",
  "Cleveland-Elyria, OH", "Former",
  "Detroit-Warren", "Former",
  "Milwaukee-Waukesha", "Former",
  "Philadelphia-Camden", "Former",
  "Pittsburgh, PA", "Former",
  "St. Louis, MO", "Former",
  "Boston-Cambridge", "Continuous",
  "New York-Newark", "Continuous",
  "Chicago-Naperville", "Continuous",
  "San Francisco-Oakland", "Continuous",
  "Houston-The Woodlands", "Post-WWII",
  "Los Angeles-Long Beach", "Post-WWII",
  "Riverside-San Bernardino", "Post-WWII",
  "San Diego-Chula Vista", "Post-WWII",
  "Miami-Fort Lauderdale", "Post-WWII",
  "Atlanta-Sandy Springs", "Emerging",
  "Dallas-Fort Worth", "Emerging",
  "Las Vegas-Henderson", "Emerging",
  "Orlando-Kissimmee", "Emerging",
  "Washington-Arlington", "Emerging",
  "Minneapolis-St. Paul", "Re-emerging",
  "Denver-Aurora", "Re-emerging",
  "Phoenix-Mesa", "Re-emerging",
  "Portland-Vancouver-Hillsboro, OR", "Re-emerging",
  "Sacramento-Roseville", "Re-emerging",
  "San Jose-Sunnyvale", "Re-emerging",
  "Tampa-St. Petersburg", "Re-emerging",
  "Seattle-Tacoma", "Re-emerging",
  "Raleigh-Cary", "Pre-emerging",
  "Charlotte-Concord", "Pre-emerging",
  "Salt Lake City", "Pre-emerging",
  "Austin-Round Rock", "Pre-emerging"
)

# Helper function: match a metro name to its gateway type
assign_gateway <- function(metro_name) {
  for (i in 1:nrow(gateway_cities)) {
    if (str_detect(metro_name, fixed(gateway_cities$pattern[i]))) {
      return(gateway_cities$gateway_type[i])
    }
  }
  return(NA_character_)
}
```

## Finding 1: National Count of Houses of Worship by Faith Tradition

**Report context:** The report examines seven non-Christian faith traditions in the U.S. and discusses their total numbers of houses of worship. Buddhism has 3 branches tracked separately; Judaism has 6 branches. The report focuses specifically on "Hindu Temples" (not yoga/meditation centers) for Hinduism.

```{r}
# Count national totals for each of the 7 faiths
national_counts <- groups_nation %>%
  inner_join(faith_mapping, by = c("Group Code" = "group_code")) %>%
  group_by(faith) %>%
  summarise(
    total_congregations = sum(Congregations, na.rm = TRUE),
    total_adherents = sum(Adherents, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_congregations))

national_counts

cat("\nKey numbers from the report:\n")
cat("Muslim mosques: 2,771 nationally (267 code)\n")
cat("Buddhist temples: 1,984 nationally (890 + 891 + 892)\n")
cat("Jewish synagogues: 3,306+ nationally (6 branches combined)\n")
cat("Baha'i communities: 983 nationally\n")
cat("Hindu temples: 433 nationally (code 895 only)\n")
cat("Sikh gurdwaras: 307 nationally\n")
cat("Zoroastrian temples: 33 nationally\n")
```

**Your interpretation:** Which non-Christian faith tradition has the most houses of worship? Which has the fewest? How do these numbers reflect each group's size and history in the U.S.?

## Finding 2: Houses of Worship in High-Concentration Metro Areas

**Report claims:** The report counts how many houses of worship are in "high-concentration" metro areas (those with 6 or more congregations of that faith). For example, there are 1,545 Buddhist temples in high-concentration metros, 230 Hindu temples, 2,082 Jewish synagogues, 2,296 mosques, and 129 Sikh gurdwaras.

```{r}
# Aggregate congregations by faith and metro area,
# then filter to high-concentration areas (6+ congregations)
metro_by_faith <- groups_metro %>%
  inner_join(faith_mapping, by = c("Group Code" = "group_code")) %>%
  group_by(faith, `Metro Name`) %>%
  summarise(congregations = sum(Congregations, na.rm = TRUE), .groups = "drop")

high_conc <- metro_by_faith %>%
  filter(congregations >= 6)

# Count of high-concentration metros and total congregations per faith
high_conc_summary <- high_conc %>%
  group_by(faith) %>%
  summarise(
    n_high_conc_metros = n(),
    total_congregations_in_hc = sum(congregations),
    .groups = "drop"
  ) %>%
  arrange(desc(total_congregations_in_hc))

high_conc_summary

cat("\nReport comparison:\n")
cat("Buddhist temples in HC metros: 1,545 (our result:",
    high_conc_summary %>% filter(faith == "Buddhist") %>% pull(total_congregations_in_hc), ")\n")
cat("Hindu temples in HC metros: 230 (our result:",
    high_conc_summary %>% filter(faith == "Hindu") %>% pull(total_congregations_in_hc), ")\n")
cat("Sikh gurdwaras in HC metros: 129 (our result:",
    high_conc_summary %>% filter(faith == "Sikh") %>% pull(total_congregations_in_hc), ")\n")
```

**Your interpretation:** Which faiths have their houses of worship concentrated in a few metros vs. spread across many? Does the number of high-concentration metros tell you something about geographic dispersion?

## Finding 3: Buddhist Temple Distribution Across Gateway City Types

**Report claim (Table 2):** Of the 1,545 Buddhist temples in high-concentration metros, 45.6% are in established gateways (Former 6.0%, Continuous 21.3%, Post-WWII 18.3%) and 25.2% are in 21st-century gateways (Emerging 9.1%, Re-emerging 14.0%, Pre-emerging 2.1%).

```{r}
# Get Buddhist temples in high-concentration metros
buddhist_hc <- high_conc %>%
  filter(faith == "Buddhist") %>%
  mutate(gateway_type = map_chr(`Metro Name`, assign_gateway))

# Calculate the distribution across gateway types
buddhist_gateway <- buddhist_hc %>%
  group_by(gateway_type) %>%
  summarise(
    n_metros = n(),
    total_congregations = sum(congregations),
    .groups = "drop"
  ) %>%
  mutate(pct = round(total_congregations / sum(buddhist_hc$congregations) * 100, 1))

buddhist_gateway

# Calculate established vs 21st-century totals
established <- buddhist_gateway %>%
  filter(gateway_type %in% c("Former", "Continuous", "Post-WWII")) %>%
  summarise(total_pct = sum(pct))

twenty_first <- buddhist_gateway %>%
  filter(gateway_type %in% c("Emerging", "Re-emerging", "Pre-emerging")) %>%
  summarise(total_pct = sum(pct))

cat("\nEstablished gateway total:", established$total_pct, "% (report claims: 45.6%)\n")
cat("21st-century gateway total:", twenty_first$total_pct, "% (report claims: 25.2%)\n")
```

**Your interpretation:** Does the data match the article's Table 2? Are Buddhist temples more concentrated in established or 21st-century gateways? What does this tell you about the history of Buddhist immigration?

## Finding 4: Hindu Temple Distribution Shows Highest Established Gateway Concentration

**Report claim (Table 3):** Of the 230 Hindu temples in high-concentration metros, 76.9% are in established gateways (Former 11.7%, Continuous 45.2%, Post-WWII 20.0%) and 23.1% are in 21st-century gateways (Emerging 17.4%, Re-emerging 5.7%, Pre-emerging 0.0%). This is the highest established-gateway concentration among the faiths studied.

```{r}
# Hindu temples in high-concentration metros
hindu_hc <- high_conc %>%
  filter(faith == "Hindu") %>%
  mutate(gateway_type = map_chr(`Metro Name`, assign_gateway))

hindu_gateway <- hindu_hc %>%
  group_by(gateway_type) %>%
  summarise(
    n_metros = n(),
    total_congregations = sum(congregations),
    .groups = "drop"
  ) %>%
  mutate(pct = round(total_congregations / sum(hindu_hc$congregations) * 100, 1))

hindu_gateway

established_hindu <- hindu_gateway %>%
  filter(gateway_type %in% c("Former", "Continuous", "Post-WWII")) %>%
  summarise(total_pct = sum(pct))

cat("\nHindu established gateway total:", established_hindu$total_pct,
    "% (report claims: 76.9%)\n")
cat("Continuous gateways alone:",
    hindu_gateway %>% filter(gateway_type == "Continuous") %>% pull(pct),
    "% (report claims: 45.2%)\n")
```

**Your interpretation:** Does the data confirm that Hindu temples have the highest concentration in established gateways? Nearly half are in continuous gateways (Boston, New York, Chicago, San Francisco) -- what might explain this?

## Finding 5: Comparing Established vs. 21st-Century Gateway Shares Across All Faiths

**Report context:** The report finds a range: Judaism is the most concentrated in established gateways (84.1%), while Baha'i has the highest share in 21st-century gateways (52.7%). This reflects each faith's immigration history -- older immigration waves settled in established gateways, while newer waves dispersed to 21st-century gateways.

```{r}
# Calculate established vs 21st-century shares for ALL faiths
all_faiths_gateway <- high_conc %>%
  mutate(gateway_type = map_chr(`Metro Name`, assign_gateway)) %>%
  mutate(gateway_era = case_when(
    gateway_type %in% c("Former", "Continuous", "Post-WWII") ~ "Established",
    gateway_type %in% c("Emerging", "Re-emerging", "Pre-emerging") ~ "21st Century",
    TRUE ~ "Non-gateway"
  )) %>%
  group_by(faith, gateway_era) %>%
  summarise(total_cong = sum(congregations), .groups = "drop") %>%
  group_by(faith) %>%
  mutate(pct = round(total_cong / sum(total_cong) * 100, 1)) %>%
  ungroup()

# Show established gateway share for each faith
established_comparison <- all_faiths_gateway %>%
  filter(gateway_era == "Established") %>%
  select(faith, established_pct = pct) %>%
  arrange(desc(established_pct))

established_comparison

cat("\nReport claims for established gateway shares:\n")
cat("Jewish: 84.1%\n")
cat("Hindu: 76.9%\n")
cat("Muslim: 51.2%\n")
cat("Sikh: 50.3%\n")
cat("Buddhist: 45.6%\n")
cat("Bahai: 45.5%\n")

# Show 21st-century shares
twentyfirst_comparison <- all_faiths_gateway %>%
  filter(gateway_era == "21st Century") %>%
  select(faith, twentyfirst_pct = pct) %>%
  arrange(desc(twentyfirst_pct))

cat("\n21st-century gateway shares:\n")
twentyfirst_comparison
```

**Your interpretation:** Which faith is most concentrated in traditional/established gateway cities? Which has dispersed the most into newer gateway cities? How does this reflect each group's immigration history?

## Summary

Summarize what you were able to reproduce and any discrepancies you found. The key findings involve:

1. Counting national houses of worship for each non-Christian faith tradition
2. Identifying "high-concentration" metro areas (6+ congregations) for each faith
3. Calculating the percentage of congregations in each gateway city type
4. Comparing established vs. 21st-century gateway concentration across faiths

Note: Some counts may differ slightly from the report due to data preparation choices (e.g., which Jewish branches to include, whether to count Hindu yoga/meditation centers). The gateway city percentage distributions should match closely because they use the same underlying metro-level data.

## Challenges and Limitations

Describe any challenges you encountered:

- How did you handle the mapping of individual Religious Census group codes to the broader faith traditions?
- The report uses "Hindu Temples" (code 895) specifically, not including "Hindu Yoga and Meditation" (code 896). Why might this matter?
- What about the Zoroastrian analysis? The report uses a threshold of 2+ temples instead of 6+ because there are so few Zoroastrian temples nationally (only 33).
- Not all high-concentration metros are in gateway cities -- the "non-gateway" percentage explains why the gateway totals don't add up to 100%.

## Data Sources

-   2020 U.S. Religion Census: <https://www.usreligioncensus.org/>
-   Gateway city classifications based on Singer et al. (2008)
