---
title: "Reverse Engineering: Non-Christian Faith Dispersion"
author: "Student Name"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze Mike McMullen's report "The Rise and Geographic Dispersion of Non-Christian Faith Traditions in the United States" and reproduce key quantitative findings using data from the 2020 U.S. Religion Census.

**Your task:**

1.  Read the report carefully (PDF file in this folder)
2.  Identify key facts and statistics that were produced through data analysis
3.  Write well-commented, well-structured R code to reproduce those findings
4.  Focus on replicable quantitative sentences (not the maps)

**Important:** This is NOT a step-by-step walkthrough. You are expected to independently: - Identify which claims in the report can be verified with data - Determine what data you need and how to obtain it - Clean and analyze the data appropriately - Reproduce the findings with your own code

## Load Libraries

```{r}
# Load the tidyverse and any other libraries you need


```

## Load Data

```{r}
# Load the 2020 U.S. Religion Census data
# You will need to download this from usreligioncensus.org
# and determine which variables/tables to use


```

## Questions to Guide Your Analysis

As you read the report, consider these questions:

1.  **What are the main quantitative claims?** Look for specific percentages, counts, or comparisons in the text and tables.

2.  **What data sources were used?** The report primarily uses the 2020 U.S. Religion Census data aggregated at the county level.

3.  **What is the methodology?** How did the author define "high-concentration areas" (6+ houses of worship per metro area)? How were gateway cities classified?

4.  **What can you reproduce?** Focus on the percentage breakdowns in Tables 1-7 that show the distribution of houses of worship across different gateway city types.

5.  **What calculations are needed?** You'll likely need to:

    -   Count houses of worship by metropolitan area
    -   Filter for high-concentration areas (6+ houses of worship)
    -   Classify metropolitan areas by gateway type
    -   Calculate percentages of houses of worship in each gateway category

## Your Analysis

Write your code below to reproduce key findings from the report. For each finding: - State the claim from the report - Show your code with clear comments - Present your results - Compare to what the report stated

```{r}
# Your analysis here


```

## Findings

Summarize what you were able to reproduce and any discrepancies you found.

## Challenges and Limitations

Describe any challenges you encountered: - What data was difficult to obtain or prepare? - What assumptions did you have to make? - What findings could not be reproduced and why?

## Data Sources

-   2020 U.S. Religion Census: <https://www.usreligioncensus.org/>
-   Gateway city classifications based on Singer et al. (2008)
