---
title: "Reverse Engineering: The Assemblies of God in 2024"
author: "Student Name"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from the Assemblies of God (AG) denominational statistical reports to reproduce key findings from Ryan Burge's article: ["The Assemblies of God in 2024"](https://www.graphsaboutreligion.com/p/the-assemblies-of-god-in-2024) published on his Substack newsletter "Graphs about Religion."

The article examines the health of the Assemblies of God denomination across multiple metrics: membership trends over time, worship attendance patterns, conversion and baptism inflows, and the changing racial composition of the denomination.

A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1. **Read the article carefully** and identify the key facts and figures the reporter presents
2. **Examine the data** provided in the CSV files in this folder
3. **Write your own code** to reproduce those key findings
4. **Document your work** with clear comments explaining your analysis

Your notebook should be well-organized, clearly commented, and demonstrate that you can independently reproduce data journalism findings.

**Important note:** The data files in this folder were reconstructed from values reported in the article and the R script. Some specific year-by-year values (especially for earlier decades) are approximations, but the key data points highlighted in the article are accurate.

## The Data

The data comes from the AG's denominational statistical reports, which the denomination publishes annually. There are four CSV files:

- **`ag_membership.csv`**: Total AG membership by year from 1975-2024. Columns: `year`, `members`.
- **`ag_attendance.csv`**: Weekly in-person worship attendance by year from 1978-2024. Columns: `year`, `attendance`.
- **`ag_race.csv`**: Racial composition of AG adherents for selected years. Columns: `year`, `white`, `black`, `hispanic`, `asian_pacific`, `native_american`, `other_mixed`, `total`.
- **`ag_conversions.csv`**: Conversion inflows from 2020-2024 broken down by type. Columns: `year`, `water_baptisms`, `spirit_baptisms`, `other`, `total`.

## Load Libraries and Data

```{r}
# Load required libraries
library(tidyverse)

# Load all four data files
membership <- read_csv("ag_membership.csv")
attendance <- read_csv("ag_attendance.csv")
race <- read_csv("ag_race.csv")
conversions <- read_csv("ag_conversions.csv")

# Preview each dataset
glimpse(membership)
glimpse(attendance)
glimpse(race)
glimpse(conversions)
```

## Finding 1: Peak Membership and the COVID Drop

**Article claim:** AG membership peaked in 2019 at about 3.3 million members. COVID caused a loss of roughly 330,000 members in a single year (2020). The 2024 total sits at 3,059,461 -- just slightly above the 3.04 million the AG reported in 2011.

```{r}
# Find the peak membership year and value
peak <- membership %>% filter(members == max(members))
cat("Peak membership year:", peak$year, "\n")
cat("Peak membership:", format(peak$members, big.mark = ","), "\n")

# Calculate the 2020 membership drop
members_2019 <- membership %>% filter(year == 2019) %>% pull(members)
members_2020 <- membership %>% filter(year == 2020) %>% pull(members)
covid_drop <- members_2019 - members_2020
cat("\n2019 membership:", format(members_2019, big.mark = ","), "\n")
cat("2020 membership:", format(members_2020, big.mark = ","), "\n")
cat("COVID drop:", format(covid_drop, big.mark = ","), "\n")

# Compare 2024 to 2011
members_2024 <- membership %>% filter(year == 2024) %>% pull(members)
members_2011 <- membership %>% filter(year == 2011) %>% pull(members)
cat("\n2024 membership:", format(members_2024, big.mark = ","), "\n")
cat("2011 membership:", format(members_2011, big.mark = ","), "\n")
cat("2024 is", format(members_2024 - members_2011, big.mark = ","), "more than 2011\n")

# The net change from 2014-2019 (to confirm the pandemic wiped out more than twice those gains)
members_2014 <- membership %>% filter(year == 2014) %>% pull(members)
gain_2014_2019 <- members_2019 - members_2014
cat("\nGains from 2014-2019:", format(gain_2014_2019, big.mark = ","), "\n")
cat("COVID drop was", round(covid_drop / gain_2014_2019, 1), "times the 2014-2019 gains\n")
```

**Your interpretation:** Does the data confirm a peak around 3.3 million in 2019, a ~330,000 member loss in 2020, and a 2024 total near the 2011 level?

## Finding 2: Attendance-to-Membership Ratio Over Time

**Article claim:** In 1979, 72% of members attended worship. That ratio dropped to 63% by 2018, plunged to 54% during the pandemic low, and has recovered to about 64% in 2024.

```{r}
# Join membership and attendance data to calculate the ratio
ratio <- membership %>%
  inner_join(attendance, by = "year") %>%
  mutate(ratio = attendance / members)

# Look at the specific years mentioned in the article
key_years <- ratio %>%
  filter(year %in% c(1979, 2018, 2020, 2021, 2024))

key_years %>%
  select(year, members, attendance, ratio) %>%
  mutate(ratio_pct = round(ratio * 100, 1))

# Find the lowest ratio (pandemic low)
lowest <- ratio %>% filter(ratio == min(ratio))
cat("\nLowest ratio year:", lowest$year, "at", round(lowest$ratio * 100, 1), "%\n")

# 1979 ratio
cat("1979 ratio:", round(ratio$ratio[ratio$year == 1979] * 100, 1), "%\n")
cat("2018 ratio:", round(ratio$ratio[ratio$year == 2018] * 100, 1), "%\n")
cat("2024 ratio:", round(ratio$ratio[ratio$year == 2024] * 100, 1), "%\n")
```

**Your interpretation:** Does the data confirm the ratios mentioned in the article? What does the trend in this ratio suggest about member engagement?

## Finding 3: 2024 Was a Banner Year for Conversions

**Article claim:** Total conversion inflows reached about 530,000 in 2024. Water baptisms nearly doubled from 94,000 in 2020 to 169,000 in 2024. Spirit baptisms grew from 62,000 to 84,000. AG churches conducted roughly 3,250 water baptisms every weekend in 2024.

```{r}
# Display the full conversions table
conversions

# 2024 totals
conv_2024 <- conversions %>% filter(year == 2024)
cat("2024 total inflows:", format(conv_2024$total, big.mark = ","), "\n")
cat("2024 water baptisms:", format(conv_2024$water_baptisms, big.mark = ","), "\n")
cat("2024 spirit baptisms:", format(conv_2024$spirit_baptisms, big.mark = ","), "\n")
cat("2024 other:", format(conv_2024$other, big.mark = ","), "\n")

# Water baptism growth from 2020 to 2024
conv_2020 <- conversions %>% filter(year == 2020)
cat("\nWater baptisms in 2020:", format(conv_2020$water_baptisms, big.mark = ","), "\n")
cat("Water baptisms in 2024:", format(conv_2024$water_baptisms, big.mark = ","), "\n")
cat("Growth:", format(conv_2024$water_baptisms - conv_2020$water_baptisms, big.mark = ","), "\n")
cat("Percent increase:", round((conv_2024$water_baptisms - conv_2020$water_baptisms) / conv_2020$water_baptisms * 100, 0), "%\n")

# Average weekly water baptisms in 2024
weekly_baptisms <- round(conv_2024$water_baptisms / 52)
cat("\nAverage weekly water baptisms in 2024:", format(weekly_baptisms, big.mark = ","), "\n")

# Spirit baptism growth
cat("\nSpirit baptisms in 2020:", format(conv_2020$spirit_baptisms, big.mark = ","), "\n")
cat("Spirit baptisms in 2024:", format(conv_2024$spirit_baptisms, big.mark = ","), "\n")
```

**Your interpretation:** Does the data confirm the conversion numbers from the article? Is 2024 really a banner year?

## Finding 4: The AG's Changing Racial Composition

**Article claim:** In 2001, 71% of AG members were White, 16% were Hispanic, and 6% were Black. By 2024, only 55% are White, nearly 23% are Hispanic, and about 11% are Black.

```{r}
# Display the racial composition data
race

# Calculate percentage-point changes from 2001 to the most recent year
race_2001 <- race %>% filter(year == min(year))
race_latest <- race %>% filter(year == max(year))

cat("Year comparison:", race_2001$year, "to", race_latest$year, "\n\n")

cat("White: ", round(race_2001$white * 100, 0), "% ->", round(race_latest$white * 100, 0), "%",
    "(change:", round((race_latest$white - race_2001$white) * 100, 0), "pp)\n")
cat("Hispanic:", round(race_2001$hispanic * 100, 0), "% ->", round(race_latest$hispanic * 100, 0), "%",
    "(change:", round((race_latest$hispanic - race_2001$hispanic) * 100, 0), "pp)\n")
cat("Black:  ", round(race_2001$black * 100, 0), "% ->", round(race_latest$black * 100, 0), "%",
    "(change:", round((race_latest$black - race_2001$black) * 100, 0), "pp)\n")

# Calculate the non-white share over time
race %>%
  mutate(nonwhite = 1 - white) %>%
  select(year, white, nonwhite) %>%
  mutate(white_pct = round(white * 100, 0),
         nonwhite_pct = round(nonwhite * 100, 0))
```

**Your interpretation:** Does the data confirm the shift from 71% to 55% White? How does the AG's 2024 composition compare to the U.S. population (58% White, 19% Hispanic, 12% Black)?

## Finding 5: Membership Growth from 1975 to Pre-Pandemic

**Article claim:** AG membership grew from about 1 million in 1975 to a peak of about 3.3 million in 2019, adding roughly 2 million members over about 45 years. The growth was remarkably steady at about 2-3% per year.

```{r}
# 1975 vs 2019 membership
members_1975 <- membership %>% filter(year == 1975) %>% pull(members)
members_2019 <- membership %>% filter(year == 2019) %>% pull(members)

cat("1975 membership:", format(members_1975, big.mark = ","), "\n")
cat("2019 membership:", format(members_2019, big.mark = ","), "\n")
cat("Total growth:", format(members_2019 - members_1975, big.mark = ","), "\n")
cat("Growth multiple:", round(members_2019 / members_1975, 1), "x\n")

# Calculate year-over-year growth rates
growth_rates <- membership %>%
  filter(year >= 1975, year <= 2019) %>%
  mutate(
    prev_members = lag(members),
    growth_rate = (members - prev_members) / prev_members * 100
  ) %>%
  filter(!is.na(growth_rate))

cat("\nAverage annual growth rate (1975-2019):", round(mean(growth_rates$growth_rate), 1), "%\n")
cat("Median annual growth rate:", round(median(growth_rates$growth_rate), 1), "%\n")
cat("Min growth rate:", round(min(growth_rates$growth_rate), 1), "% in", growth_rates$year[which.min(growth_rates$growth_rate)], "\n")
cat("Max growth rate:", round(max(growth_rates$growth_rate), 1), "% in", growth_rates$year[which.max(growth_rates$growth_rate)], "\n")

# How many years had growth between 1% and 4%?
steady_growth <- growth_rates %>% filter(growth_rate >= 1, growth_rate <= 4)
cat("\nYears with 1-4% growth:", nrow(steady_growth), "of", nrow(growth_rates), "years\n")
```

**Your interpretation:** Does the data support the claim of steady 2-3% annual growth? Was the growth truly "remarkably steady"?

## Summary

Write a paragraph summarizing your findings. Were you able to replicate the article's key claims? What patterns did you find most noteworthy? What additional data or context would have been useful for this analysis?

## Data Sources

- AG Denominational Statistical Reports, 2024
- Article: Burge, Ryan. "The Assemblies of God in 2024." Graphs about Religion.
