---
title: "Reverse Engineering: Changing Faiths"
author: "Student Name"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from Pew Research Center's 2014 Religious Landscape Study to reproduce key findings from FiveThirtyEight's analysis: ["Evangelical Protestants Are The Biggest Winners When People Change Faiths"](http://fivethirtyeight.com/datalab/evangelical-protestants-are-the-biggest-winners-when-people-change-faiths/) by Leah Libresco.

The article uses survey data on religious switching patterns and fertility rates to project what America's religious makeup would look like if current trends continued for many generations. A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1. **Read the article carefully** and identify the key facts and figures the reporter presents
2. **Examine the data** provided in the `data/` folder
3. **Write your own code** to reproduce those key findings
4. **Document your work** with clear comments explaining your analysis

Your notebook should be well-organized, clearly commented, and demonstrate that you can independently reproduce data journalism findings.

## The Data

The data comes from Pew's 2014 Religious Landscape Study, which surveyed more than 35,000 Americans about their religious beliefs and practices. The key question: what religion were you raised in, and what religion are you now?

You have four data files in the `data/` folder:

- **transition_matrix.csv**: A 12x12 matrix of switching probabilities. Each row represents a religion someone was raised in; each column represents their current religion. Values are probabilities (each row sums to approximately 1.0).

- **initial_distribution.csv**: The starting share of each religion in the U.S. population (the proportion of Americans raised in each tradition).

- **fertility_rates.csv**: Average number of children per woman for each religious group.

- **current.csv**: The output of FiveThirtyEight's simulation showing projected religious distribution over 101 iterations as the model approaches equilibrium.

The 12 religious categories are: Buddhist, Catholic, Evangelical Protestant, Hindu, Historically Black Protestant, Jehovah's Witness, Jewish, Mainline Protestant, Mormon, Muslim, Orthodox Christian, and Unaffiliated.

## Load Libraries and Data

```{r}
# Load required libraries
library(tidyverse)

# Load the datasets


```

## Analysis

Now it's your turn. Read through the FiveThirtyEight article and identify the key facts and claims made. Then write well-commented code below to reproduce those findings using the data provided.

Some questions to guide your work:
- What does the article say about retention rates for different religious groups?
- What patterns of religious switching does the article highlight?
- What does the model project about the future religious makeup of America?
- How do fertility rates factor into the projections?

Structure your analysis with clear section headers and comments explaining what you're testing and what you found.

```{r}
# Your analysis code here

```

## Findings

Write a summary of your findings here. Did your analysis confirm what the article reported? Were there any discrepancies? What did you learn about religious switching patterns in America?

