---
title: "Reverse Engineering: Changing Faiths"
author: "Student Name"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from Pew Research Center's 2014 Religious Landscape Study to reproduce key findings from FiveThirtyEight's analysis: ["Evangelical Protestants Are The Biggest Winners When People Change Faiths"](http://fivethirtyeight.com/datalab/evangelical-protestants-are-the-biggest-winners-when-people-change-faiths/) by Leah Libresco.

The article uses survey data on religious switching patterns and fertility rates to project what America's religious makeup would look like if current trends continued for many generations. A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1. **Read the article carefully** and identify the key facts and figures the reporter presents
2. **Examine the data** provided in the `data/` folder
3. **Write your own code** to reproduce those key findings
4. **Document your work** with clear comments explaining your analysis

Your notebook should be well-organized, clearly commented, and demonstrate that you can independently reproduce data journalism findings.

## The Data

The data comes from Pew's 2014 Religious Landscape Study, which surveyed more than 35,000 Americans about their religious beliefs and practices. The key question: what religion were you raised in, and what religion are you now?

You have four data files in the `data/` folder:

- **transition_matrix.csv**: A 12x12 matrix of switching probabilities. Each row represents a religion someone was raised in; each column represents their current religion. Values are probabilities (each row sums to approximately 1.0).

- **initial_distribution.csv**: The starting share of each religion in the U.S. population (the proportion of Americans raised in each tradition).

- **fertility_rates.csv**: Average number of children per woman for each religious group.

- **current.csv**: The output of FiveThirtyEight's simulation showing projected religious distribution over 101 iterations as the model approaches equilibrium.

The 12 religious categories are: Buddhist, Catholic, Evangelical Protestant, Hindu, Historically Black Protestant, Jehovah's Witness, Jewish, Mainline Protestant, Mormon, Muslim, Orthodox Christian, and Unaffiliated.

## Load Libraries and Data

```{r}
# Load required libraries
library(tidyverse)

# Load the datasets
transition <- read_csv("data/transition_matrix.csv")
initial <- read_csv("data/initial_distribution.csv")
fertility <- read_csv("data/fertility_rates.csv")

# The current.csv has no headers - it's 101 rows x 12 columns
# Each row is one generation, each column is a religion in alphabetical order
religions <- c("Buddhist", "Catholic", "Evangelical Protestant", "Hindu",
               "Historically Black Protestant", "Jehovahs Witness", "Jewish",
               "Mainline Protestant", "Mormon", "Muslim", "Orthodox Christian",
               "Unaffiliated")

equilibrium <- read_csv("data/current.csv", col_names = religions)
```

## Finding 1: Retention Rates - Who Stays in Their Faith?

**Article claim:** People raised evangelical Protestant have a 65% chance of staying evangelical. Catholics retain about 59% of those raised Catholic. Mainline Protestants have only a 45% retention rate. And the unaffiliated retain about 53%.

The retention rate is the diagonal value in the transition matrix - the probability that someone raised in a faith stays in that faith.

```{r}
# Extract the diagonal of the transition matrix (retention rates)
# The diagonal represents the probability of staying in the faith you were raised in
retention <- transition %>%
  mutate(retention_rate = case_when(
    raised_in == "Buddhist" ~ Buddhist,
    raised_in == "Catholic" ~ Catholic,
    raised_in == "Evangelical Protestant" ~ `Evangelical Protestant`,
    raised_in == "Hindu" ~ Hindu,
    raised_in == "Historically Black Protestant" ~ `Historically Black Protestant`,
    raised_in == "Jehovahs Witness" ~ `Jehovahs Witness`,
    raised_in == "Jewish" ~ Jewish,
    raised_in == "Mainline Protestant" ~ `Mainline Protestant`,
    raised_in == "Mormon" ~ Mormon,
    raised_in == "Muslim" ~ Muslim,
    raised_in == "Orthodox Christian" ~ `Orthodox Christian`,
    raised_in == "Unaffiliated" ~ Unaffiliated
  )) %>%
  select(raised_in, retention_rate) %>%
  arrange(desc(retention_rate))

retention

# Check the specific claims from the article
cat("Evangelical retention rate:", round(retention$retention_rate[retention$raised_in == "Evangelical Protestant"] * 100, 1), "%\n")
cat("Catholic retention rate:", round(retention$retention_rate[retention$raised_in == "Catholic"] * 100, 1), "%\n")
cat("Unaffiliated retention rate:", round(retention$retention_rate[retention$raised_in == "Unaffiliated"] * 100, 1), "%\n")
cat("Mainline Protestant retention rate:", round(retention$retention_rate[retention$raised_in == "Mainline Protestant"] * 100, 1), "%\n")
```

**Your interpretation:** Does the data confirm ~65% for evangelicals, ~59% for Catholics, ~45% for mainline, and ~53% for unaffiliated?

## Finding 2: Catholic-to-Evangelical vs. Evangelical-to-Catholic Switching

**Article claim:** About 10% of people raised Catholic wind up as evangelicals, but just 2% of people born as evangelicals wind up Catholic.

```{r}
# What percent of those raised Catholic become evangelical?
catholic_to_evangelical <- transition %>%
  filter(raised_in == "Catholic") %>%
  pull(`Evangelical Protestant`)

cat("Percent of Catholics who become Evangelical:", round(catholic_to_evangelical * 100, 1), "%\n")

# What percent of those raised evangelical become Catholic?
evangelical_to_catholic <- transition %>%
  filter(raised_in == "Evangelical Protestant") %>%
  pull(Catholic)

cat("Percent of Evangelicals who become Catholic:", round(evangelical_to_catholic * 100, 1), "%\n")

# What is the ratio?
cat("Ratio:", round(catholic_to_evangelical / evangelical_to_catholic, 1), "Catholics become evangelical for every 1 evangelical who becomes Catholic\n")
```

**Your interpretation:** Does the data show roughly a 5-to-1 ratio in favor of evangelicals gaining from Catholics?

## Finding 3: Mainline-to-Evangelical vs. Evangelical-to-Mainline Switching

**Article claim:** 19% of mainline Protestants wind up becoming evangelical, while only 12% of those raised evangelical end up in mainline congregations.

```{r}
# What percent of mainline Protestants become evangelical?
mainline_to_evangelical <- transition %>%
  filter(raised_in == "Mainline Protestant") %>%
  pull(`Evangelical Protestant`)

cat("Percent of Mainline who become Evangelical:", round(mainline_to_evangelical * 100, 1), "%\n")

# What percent of evangelicals become mainline?
evangelical_to_mainline <- transition %>%
  filter(raised_in == "Evangelical Protestant") %>%
  pull(`Mainline Protestant`)

cat("Percent of Evangelicals who become Mainline:", round(evangelical_to_mainline * 100, 1), "%\n")

# Evangelicals gain more than they lose to mainline
cat("Net advantage for evangelicals:", round((mainline_to_evangelical - evangelical_to_mainline) * 100, 1), "percentage points\n")
```

**Your interpretation:** Does the data confirm that evangelicals gain nearly twice as many from mainline as they lose?

## Finding 4: Initial vs. Current vs. Equilibrium Population Shares

**Article claim:** Unaffiliated Americans make up 22.8% of the current population (up from 9.2% raised unaffiliated). At equilibrium (conversion only, no fertility), the unaffiliated would reach 29%, evangelicals would grow from 26% to 32%, and Catholics would shrink from 21% to 8%.

```{r}
# The initial distribution shows what share of Americans were RAISED in each religion
initial %>% arrange(desc(share))

# Row 1 of equilibrium is approximately the current distribution
# The last row is the equilibrium
current_shares <- equilibrium %>% slice(1)
equilibrium_shares <- equilibrium %>% slice(n())

# Compare initial (raised), current, and equilibrium
comparison <- tibble(
  religion = religions,
  raised_in = initial$share,
  current = as.numeric(current_shares[1,]),
  equilibrium = as.numeric(equilibrium_shares[1,])
) %>%
  mutate(
    raised_in_pct = round(raised_in * 100, 0),
    current_pct = round(current * 100, 0),
    equilibrium_pct = round(equilibrium * 100, 0)
  ) %>%
  select(religion, raised_in_pct, current_pct, equilibrium_pct) %>%
  arrange(desc(current_pct))

comparison

# Highlight the key article claims
cat("\nUnaffiliated: raised =", comparison$raised_in_pct[comparison$religion == "Unaffiliated"],
    "%, current =", comparison$current_pct[comparison$religion == "Unaffiliated"],
    "%, equilibrium =", comparison$equilibrium_pct[comparison$religion == "Unaffiliated"], "%\n")

cat("Evangelical: raised =", comparison$raised_in_pct[comparison$religion == "Evangelical Protestant"],
    "%, current =", comparison$current_pct[comparison$religion == "Evangelical Protestant"],
    "%, equilibrium =", comparison$equilibrium_pct[comparison$religion == "Evangelical Protestant"], "%\n")

cat("Catholic: raised =", comparison$raised_in_pct[comparison$religion == "Catholic"],
    "%, current =", comparison$current_pct[comparison$religion == "Catholic"],
    "%, equilibrium =", comparison$equilibrium_pct[comparison$religion == "Catholic"], "%\n")
```

**Your interpretation:** Does the data confirm that evangelicals are the biggest winners (growing from 26% to 32%) and Catholics are the biggest losers (shrinking from 21% to 8%) at equilibrium?

## Finding 5: Fertility Rates - Mormons Highest, Unaffiliated Lowest

**Article claim:** Mormons have the highest fertility rate at 3.4 children per woman. Unaffiliated Americans have the lowest at 1.7 children per woman (below replacement level of ~2.1).

```{r}
# Display fertility rates sorted from highest to lowest
fertility %>% arrange(desc(children_per_woman))

# Key comparisons from the article
cat("\nMormon fertility rate:", fertility$children_per_woman[fertility$religion == "Mormon"], "\n")
cat("Muslim fertility rate:", fertility$children_per_woman[fertility$religion == "Muslim"], "\n")
cat("Unaffiliated fertility rate:", fertility$children_per_woman[fertility$religion == "Unaffiliated"], "\n")
cat("Mainline Protestant fertility rate:", fertility$children_per_woman[fertility$religion == "Mainline Protestant"], "\n")

# How many groups are below replacement level (2.1)?
below_replacement <- fertility %>% filter(children_per_woman < 2.1)
cat("\nGroups below replacement level (2.1):", nrow(below_replacement), "\n")
cat("They are:", paste(below_replacement$religion, collapse = ", "), "\n")
```

**Your interpretation:** Does the data confirm that Mormons have the highest fertility (3.4) and unaffiliated the lowest (1.7)? What does this suggest about the long-term projections when fertility is included?

## Summary

Write a summary of your findings here. Did your analysis confirm what the article reported? Were there any discrepancies? What did you learn about religious switching patterns in America?
