---
title: "pre_lab_07.Rmd"
author: "derek willis"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Task 1: Load libraries and settings

**Task** Run the following code in the gray-colored codeblock below to load the tidyverse library and the tidycensus library (remove the comment in front of install.packages if you need to install the latter)

```{r}
# turn off sci notation
options(scipen=999)
library(tidyverse)
# if needed, install tidycensus, then load it.
# install.packages("tidycensus")
library(tidycensus)
```

# Intro to APIs: The Census

We're going to use a library called `tidycensus` which makes calls to the Census API in a very tidy way, and gives you back tidy data. That means we don't have to go through the process of importing the data from a file. I can't tell you how amazing this is, speaking from experience.

To use the API, you need to [an API key from the Census Bureau](https://api.census.gov/data/key_signup.html). It takes a few minutes and you need to activate your key via email. Once you have your key, you need to set that for this session. Just FYI: Your key is your key. Do not share it.

### Task 2: Sign up for an API Key and set it

**Task** Replace YOUR KEY HERE in the codeblock below with your Census API Key (leave the quotemarks).

```{r echo=FALSE}
#census_api_key("YOUR KEY HERE", install=TRUE)
```

The `install=TRUE` part saves your key for future use.

So to give you some idea of how complicated the data is, let's pull up just one file from the decennial Census. We'll use Summary File 1, or SF1. That has the major population and housing stuff.

### Task 3: Examining Census data

**Task** Run the following codeblock and look at the resulting variable.

```{r}
sf1 <- load_variables(2010, "sf1", cache = TRUE)
View(sf1)
```

Note: There are thousands of variables in SF1. That's not a typo. Open it in your environment by double clicking. As you scroll down, you'll get an idea of what you've got to choose from.

The `name` variable is the Census code for each kind of data, while `label` describes it and `concept` is a broader category.

If you think that's crazy, try the SF3 file from 2000.

**Task** Run the following codeblock and look at the resulting variable.

```{r}
sf3 <- load_variables(2000, "sf3", cache = TRUE)
```

### Task 4: Calculate the fastest-growing state

So let's try to answer a question using the Census. What is the fastest growing state since 2000?

To answer this, we need to pull the total population by state in each of the decennial census.

**Task** Run the following codeblock and examine the 2010 results

```{r}
p00 <- get_decennial(geography = "state", variables = "P001001", year = 2000)
p10 <- get_decennial(geography = "state", variables = "P001001", year = 2010)
```

Let's take a peek at 2010.

```{r}
p10
```

As you can see, we have a GEOID, NAME, then variable and value. Variable and value are going to be the same. Because those are named the same thing, to merge them together, we need to rename them. Remember how we used `rename` in the last lab? We're going to use it some more.

**Task** Run the following codeblock to rename the population total columns in each dataframe:

```{r}
p10 |> select(GEOID, NAME, value) |> rename(Population2010=value) -> p10

p00 |> select(GEOID, NAME, value) |> rename(Population2000=value) -> p00

```

Now we join the data together.

**Task** Run the following codeblock to combine the two dataframes into a single variable called `alldata`:

```{r}
alldata <- p00 |> inner_join(p10)
```

And now we calculate the percent change.

**Task** Run the following codeblock to calculate the percentage change and put the state with the largest percentage growth below. **Answer**

```{r}
alldata |> mutate(change = ((Population2010-Population2000)/Population2000)*100) |> arrange(desc(change))
```

You may be asking: hey, wasn't there a 2020 Census? Where's that data? The answer is that it's coming - the Census Bureau has a [schedule of releases](https://www.census.gov/programs-surveys/popest/about/schedule.html). There are some variables available:

```{r}
pl <- load_variables(2020, "pl")
View(pl)
```

## The ACS

In 2010, the Census Bureau replaced SF3 with the American Community Survey. The Good News is that the data would be updated on a rolling basis. The bad news is that it's more complicated because it's more like survey data with a large sample. That means there's margins of error and confidence intervals to worry about.

What is Maryland's richest county?

We can measure this by median household income. That variable is `B19013_001`, so we can get that data like this (I'm narrowing it to the top 20 for simplicity):

### Task 5: Answer Questions with ACS Data

**Task** Run the following codeblock to find the Maryland county with the highest median household income and put the answer below. **Answer**

```{r}
md <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2021)

md <- md |> arrange(desc(estimate)) |> top_n(20, estimate)

md
```

What do the top counties all have in common? Lots of suburban flight from D.C and Baltimore. But do the margins of error let us say one county is richer than the other? We can find this out visually using error bars. Don't worry much about the code here -- we'll cover that soon enough.

### Task 6: Margins of Error

**Task** Run the following codeblock to create a chart showing the margin of error for each county. The current headline isn't very descriptive of the data. What should it be? **Answer**

```{r}
md |>
  mutate(NAME = gsub(" County, Maryland", "", NAME)) |>
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red") +
  labs(title = "Household income by county in Maryland",
       subtitle = "2017-2021 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```

As you can see, some of the error bars are quite wide. Some are narrow. But if the bars overlap, it means the difference between the two counties is within the margin of error, and the differences aren't statistically significant.

Is the difference between Calvert and Montgomery significant? **Answer**

Is the difference between Howard and everyone else significant? **Answer**

### Task 7: Joining Multiple Years Together

Let's ask another question of the ACS -- did any counties lose income from the time of the global financial crisis to the current 5-year window?

Let's re-label our first household income data.

**Task** Run the following codeblock to retrieve Maryland county median income from 2010 and from 2021.

```{r}
md21 <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2021)

md10 <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2010)
```

What we're going to do next is a lot, but each step is simple. We're going to join the data together, so each county has one line of data. Then we're going to rename some fields that repeat. Then we're going to calculate the minimum and maximum value of the estimate using the margin of error. That'll help us later. After that, we're going to calculate a percent change and sort it by that change.

**Task** Run the following codeblock to join the data, rename the population estimates and then calculate the values we need. Notice how we select *after* we do the initial calculations so we only see the columns we're interested in (and drop the ones we don't want to see). Also note how the select function has the minus sign (-) in front of some columns? That means we don't want to see them in our result. Which jurisdictions saw the largest increases in median income? Did any counties see their median incomes fall from 2010-2021? **Answer**

```{r}
md10 |>
  # join md10 to md21 using the shared GEOID and NAME columns
  inner_join(md21, by=c("GEOID", "NAME")) |>
  # rename the first estimate to clarify that it's from 2010 and the second estimate to clarify that it's from 2021
  rename(estimate2010=estimate.x, estimate2021=estimate.y) |>
  mutate(min2010 = estimate2010-moe.x, max2010 = estimate2010+moe.x, min2020 = estimate2021-moe.y, max2021 = estimate2021+moe.y) |>
  select(-variable.x, -variable.y, -moe.x, -moe.y) |>
  mutate(change = ((estimate2021-estimate2010)/estimate2010)*100) |>
  arrange(desc(change))
```

## Joining Census Data with U.S. Religion Census Data

Now that we've learned how to work with Census API data, let's combine it with data from the 2020 U.S. Religion Census. This will let us explore the relationship between demographic characteristics and religious adherence patterns.

The key to joining these datasets is the FIPS code - a standardized geographic identifier used by the Census Bureau. State FIPS codes are 2 digits, and county FIPS codes are 5 digits (state + county).

### Task 8: Load U.S. Religion Census Data

**Task** Run the following codeblock to load the state and county religion census data. Examine the structure - what columns do you see? **Answer**

```{r}
religion_state <- read_csv("data/usreligioncensus_state_2020.csv")
religion_county <- read_csv("data/usreligioncensus_county_2020.csv")

glimpse(religion_state)
glimpse(religion_county)
```

Notice that the state data has a "State Code" column (FIPS code) and the county data has a "FIPS" column (5-digit county FIPS). These are the keys we'll use to join with Census data.

### Task 9: Join State-Level Census and Religion Data

Let's get state-level population data from the 2020 Census and join it with the religion census data to explore which states have the highest percentage of religious adherents.

**Task** Run the following code to get 2020 state populations and join them with religion data. Notice that we need to make sure the FIPS codes match in format (both as character types). Which state has the highest percentage of adherents? Which has the lowest? **Answer**

```{r}
# Get 2020 state population from Census
state_pop_2020 <- get_decennial(
  geography = "state",
  variables = "P1_001N",  # Total population
  year = 2020
)

# Clean the religion data and prepare for joining
religion_state_clean <- religion_state |>
  # Make sure State Code is character for joining
  mutate(`State Code` = as.character(`State Code`)) |>
  # Pad single-digit state codes with leading zero
  mutate(`State Code` = str_pad(`State Code`, width = 2, pad = "0"))

# Join the datasets
state_religion_combined <- state_pop_2020 |>
  inner_join(religion_state_clean, by = c("GEOID" = "State Code"))

# View the combined data
state_religion_combined |>
  select(NAME, value, `State Name`, Adherents, `Adherents as % of Population`, `Diversity Index Value`) |>
  arrange(desc(`Adherents as % of Population`))
```

### Task 10: Explore State Religious Diversity

The U.S. Religion Census calculates a diversity index that measures how religiously diverse each state is. Let's see if there's a relationship between a state's population and its religious diversity.

**Task** Run the following code to examine religious diversity by state. Do larger states tend to be more or less religiously diverse? **Answer**

```{r}
state_religion_combined |>
  select(NAME, value, `Diversity Index Value`, Adherents, `Adherents as % of Population`) |>
  arrange(desc(`Diversity Index Value`)) |>
  head(10)

# Compare to least diverse
state_religion_combined |>
  select(NAME, value, `Diversity Index Value`, Adherents, `Adherents as % of Population`) |>
  arrange(`Diversity Index Value`) |>
  head(10)
```

### Task 11: Join County-Level Data

Now let's work with county-level data to explore Maryland counties specifically. We'll join ACS income data with religion census data to see if there's a relationship between median household income and religious adherence.

**Task** Run the following code to get Maryland county data from both sources and join them. What patterns do you notice? **Answer**

```{r}
# Get Maryland county median income (we already have this from earlier, but let's be explicit)
md_income <- get_acs(
  geography = "county",
  variables = c(medincome = "B19013_001"),
  state = "MD",
  year = 2021
)

# Filter religion data for Maryland counties only
md_religion <- religion_county |>
  filter(`State Name` == "Maryland") |>
  mutate(FIPS = as.character(FIPS))

# Join the data
md_combined <- md_income |>
  inner_join(md_religion, by = c("GEOID" = "FIPS"))

# Examine the relationship between income and religious adherence
md_combined |>
  select(NAME, estimate, `County Name`, `Adherents as % of Population`, `Diversity Index Value`, Congregations) |>
  arrange(desc(estimate))
```

### Task 12: Analyze Maryland Counties

Let's dig deeper into the Maryland data. We can look at which counties have the most religious diversity and how that relates to other characteristics.

**Task** Run the following code to find the most and least religiously diverse counties in Maryland. Then look at the relationship between median income and religious diversity. What do you observe? **Answer**

```{r}
# Most diverse Maryland counties
md_combined |>
  select(NAME, estimate, `Diversity Index Value`, `Adherents as % of Population`, `Largest Body by # of Adherents`) |>
  arrange(desc(`Diversity Index Value`)) |>
  head(10)

# Least diverse Maryland counties
md_combined |>
  select(NAME, estimate, `Diversity Index Value`, `Adherents as % of Population`, `Largest Body by # of Adherents`) |>
  arrange(`Diversity Index Value`) |>
  head(10)

# Calculate correlation between income and diversity
md_combined |>
  summarize(
    correlation = cor(estimate, `Diversity Index Value`, use = "complete.obs")
  )
```

### Task 13: Congregations Per Capita

Let's explore whether wealthier counties have more or fewer congregations per capita.

**Task** Run the following code to calculate and examine the relationship between median income and congregations per 100,000 population. What pattern emerges? **Answer**

```{r}
md_combined |>
  select(NAME, estimate, `Congregations Per 100,000 Population`, `Adherents as % of Population`) |>
  arrange(desc(`Congregations Per 100,000 Population`)) |>
  head(10)

# Look at correlation
md_combined |>
  summarize(
    correlation = cor(estimate, `Congregations Per 100,000 Population`, use = "complete.obs")
  )
```

### Task 14: Compare Across States

Finally, let's use what we've learned to compare religious characteristics across states. We'll get ACS data for median household income by state and join it with the religion census data.

**Task** Run the following code to get state income data and combine it with religion data. Which states have high adherence rates but lower incomes? Which have high incomes but lower adherence? **Answer**

```{r}
# Get state median household income
state_income <- get_acs(
  geography = "state",
  variables = c(medincome = "B19013_001"),
  year = 2021
)

# Join with religion data
state_combined_full <- state_income |>
  inner_join(religion_state_clean, by = c("GEOID" = "State Code"))

# High adherence, lower income
state_combined_full |>
  filter(`Adherents as % of Population` >= "60%") |>
  select(NAME, estimate, `Adherents as % of Population`, `Diversity Index Value`, `Largest Body by # of Adherents`) |>
  arrange(estimate) |>
  head(10)

# High income, lower adherence
state_combined_full |>
  filter(`Adherents as % of Population` <= "40%") |>
  select(NAME, estimate, `Adherents as % of Population`, `Diversity Index Value`) |>
  arrange(desc(estimate)) |>
  head(10)
```

### Task 15: Your Own Analysis

**Task** Come up with your own question that combines Census API data with U.S. Religion Census data. Write the question, the code to answer it, and your interpretation. **Answer**

```{r}
# Your code here
```
