---
title: "lab_06"
author: "Derek Willis"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries

```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use clergy contribution data from multiple election cycles and United Methodist church membership data to do this.

## Load Data

First, let's load all the data we'll need for this lab.

```{r}
# Load the five clergy contribution files for different election cycles




# Load the two United Methodist data files


```

## Answer questions

Q1. You want to understand how much clergy members have contributed across all five election cycles combined. Load and combine the five clergy contribution files into a single dataframe. Then calculate the total amount contributed by each person across all cycles, and identify the top 10 contributors by total amount. Who are they, and how much did each contribute in total? What do you notice about the top contributors?

A1.

```{r}
```

Q2. Now you're interested in understanding giving patterns by election cycle. Using your combined dataframe from Q1, calculate the total dollar amount contributed in each election cycle, and the total number of contributions in each cycle. Which cycle had the most money contributed? Which had the most individual contributions? What's the difference between those two results, and what might explain it?

A2.

```{r}
```

Q3. You're interested in understanding which contributors gave to the same committees across multiple election cycles. First, create a dataframe of contributions from 2020 only, with contributor name and committee name. Then create another dataframe of contributions from 2024 only, with the same fields. Join these two dataframes to find contributors who gave to the same committee in both 2020 and 2024. How many contributor-committee combinations appear in both cycles? Who are some of the contributors who stayed loyal to the same committees?

A3.

```{r}
```

Q4. Some clergy members have been more consistent donors across election cycles. Using the combined contribution data, create a dataframe that shows each contributor and how many different election cycles they contributed in (someone who gave in 2016, 2018, and 2024 would have 3 cycles). Who contributed in all five election cycles? How many people contributed in only one cycle vs. multiple cycles? What patterns do you notice?

A4.

```{r}
```

Q5. You're interested in whether there are geographic patterns in clergy giving. Using the combined contribution data, calculate the total amount contributed by clergy in each city (`contributor_city`). What are the top 5 cities by total contribution amount? Then calculate the number of unique contributors in each of those top 5 cities. Are the cities with the most money also the cities with the most contributors, or are there differences?

A5.

```{r}
```

Q6. Use `pivot_wider` to create a summary showing each contributor and their total contributions in the 2020 and 2024 election cycles only (create a dataframe with columns for contributor_name, total_2020, and total_2024). Then calculate a new column showing the change in contribution amount from 2020 to 2024. Which contributors increased their giving the most? Which decreased the most? What percentage of contributors increased vs. decreased?

A6.

```{r}
```

Q7. Now let's work with the United Methodist data. Join the `umdata_people` and `umdata_people_all` datasets using `GCFAId` as the key. Before joining, select only the unique columns from `umdata_people_all` (ConferenceId and ConferenceName) to avoid duplicate columns. After joining, filter for only clergy (not laity) and then identify the top 5 conferences by number of clergy members. Which conferences have the most clergy?

A7.

```{r}
```

Q8. Using the joined United Methodist data from Q7, compare the number of Active clergy vs. Deceased clergy in the dataset. Which conference has the highest number of Active clergy? What about Deceased clergy? Calculate the ratio of Active to Deceased clergy for the top 5 conferences by total clergy count. What patterns do you notice?

A8.

```{r}
```

Q9. You want to explore whether there's any overlap between clergy who made political contributions and clergy in the United Methodist database. Using the combined contribution data and the UM data, identify any contributors whose names (first and last name) match clergy in the `umdata_people` file. You'll need to join on both first and last names. How many matches do you find? What challenges does this kind of name-based matching present?

Hint: You may need to use `inner_join` with multiple conditions, or create a combined name field in both datasets to join on.

A9.

```{r}
```

Q10. Choose your own adventure: Come up with a question that requires you to use at least one join and one aggregation (sum, mean, count, etc.) using either the clergy contribution data, the UM data, or both. Write the question, then write the code to answer it, and provide an answer.

A10: **Your question here**

```{r}
```
