---
title: "lab_06"
author: "Derek Willis"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use clergy contribution data from multiple election cycles and United Methodist church membership data to do this.

## Load Data

First, let's load all the data we'll need for this lab.

```{r}
# Load the five clergy contribution files for different election cycles

# Load the two United Methodist data files
```

## Answer questions

Q1. Now you're interested in understanding giving patterns by election cycle. Combine the five clergy contribution files into a single dataframe, then calculate the total dollar amount contributed in each election cycle, and the total number of contributions in each cycle. Which cycle had the most money contributed? Which had the most individual contributions? What's the difference between those two results, and what might explain it?

A1. 

```{r}
```

Q2. Some clergy members have been more consistent donors across election cycles. Using the combined contribution data, create a dataframe that shows each contributor and how many different election cycles they contributed in (someone who gave in 2016, 2018, and 2024 would have 3 cycles). You can use group_by and summarize to do this. Who contributed in all five election cycles? How many people contributed in only one cycle vs. multiple cycles?

A2. 

```{r}
```

Q3. You're interested in whether there are geographic patterns in clergy giving. Using the combined contribution data, calculate the total amount contributed by clergy in each city (`contributor_city`). What are the top 5 cities by total contribution amount? Then calculate the number of unique contributors in each of those top 5 cities. Are the cities with the most money also the cities with the most contributors, or are there differences?

A3. 

```{r}
```

Q4. Use `pivot_wider` to create a summary showing each contributor and their total contributions in the 2020 and 2024 election cycles only (so create a dataframe with columns for contributor_name, total_2020, and total_2024). Then calculate a new column showing the change in contribution amount from 2020 to 2024. Which contributors increased their giving the most? Which decreased the most?

A4. 

```{r}
```

Q5. Now let's work with the United Methodist data. Join the `umdata_people` and `umdata_people_all` datasets using `GCFAId` as the key. Before joining, select only the unique columns from `umdata_people_all` (ConferenceId and ConferenceName) to avoid duplicate columns. After joining, filter for only clergy (not laity) and count them by conference and AccountStatus. Which conferences have the most clergy, and how do Active vs. Deceased clergy compare?

A5.

```{r}
```

Q6. You want to explore whether there's any overlap between clergy who made political contributions and clergy in the United Methodist database. Using the combined contribution data and the UM data, identify any contributors whose names (first and last name) match clergy in the `umdata_people` file. How many matches do you find? What challenges does this kind of name-based matching present? Are there ways to improve the accuracy of matching?

A6. 

```{r}
```

Q7. Choose your own adventure: Come up with a question that requires you to use at least one join and one aggregation (sum, mean, count, etc.) using either the clergy contribution data, the UM data, or both. Write the question, then write the code to answer it, and provide an answer.

A7: 

```{r}
```
