---
title: "Pre-Lab 13: Using LLMs with Religion Data"
author: "Derek Willis"
date: "2026-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Large language models (LLMs) like Claude are not just chatbots — they are tools journalists can use to extract, classify and summarize information from documents that are too messy or unstructured for ordinary data processing. In this pre-lab you will learn how to call LLMs directly from R using the `ellmer` package, which provides a clean interface to Anthropic's Claude and other providers.

We will work through three uses that are directly relevant to religion data journalism:

1. **Basic chat** — asking an LLM a question and getting a plain-text response
2. **Structured extraction from text** — pulling specific facts out of a research article into a tidy dataframe
3. **Structured extraction from a PDF** — sending a PDF directly to an LLM and extracting structured claims

Along the way you will practice a critical skill: verifying LLM output against its source. LLMs can extract the wrong number from a table with complete confidence. A journalist's job is to check.

## Setup

First, install `ellmer` if you have not already. Run this in your console, not in a code chunk:

```
install.packages("ellmer")
```

```{r}
library(ellmer)
library(tidyverse)
```

### API Key

To call Claude, you need an Anthropic API key. Store it as an environment variable so it never appears in your code or gets accidentally committed to GitHub. Run this line **in your console** and then restart R:

```
Sys.setenv(ANTHROPIC_API_KEY = "your-key-here")
```

Once set, `ellmer` will find it automatically whenever you call `chat_anthropic()`. You can verify it is available with:

```{r}
# Should print TRUE if your key is set — never print the key itself
nchar(Sys.getenv("ANTHROPIC_API_KEY")) > 0
```

## Part 1: Basic Chat

The central object in `ellmer` is a **Chat**. You create one with a provider function like `chat_anthropic()`, optionally giving it a system prompt that shapes how it behaves. Then you call `$chat()` to send a message and get a response.

```{r}
chat <- chat_anthropic(
  system_prompt = "You are a helpful assistant with expertise in U.S. religion data and sociology of religion."
)
```

Let's ask it something relevant to what we've been studying:

```{r}
chat$chat("What is Simpson's Diversity Index and how is it used to measure religious diversity at the county level?")
```

Notice that the response streamed to your console as it arrived. You can also capture it silently as a string by setting `echo = "none"`:

```{r}
chat2 <- chat_anthropic(echo = "none")
response <- chat2$chat("In one sentence, name the three most common Christian denominations in the United States.")
cat(response)
```

Each `Chat` object holds its full conversation history. That means you can ask follow-up questions and the model remembers what was said. But it also means each turn adds to your token usage. For data extraction tasks - where you just want a structured result, not a dialogue - it is better to create a fresh chat for each task.

Let's see the token cost of our first conversation so far:

```{r}
chat$get_cost()
```

## Part 2: Structured Extraction from Text

`$chat()` returns free text, which is hard to work with programmatically. The `$extract_data()` method is different: you define the exact structure you want using `type_object()`, `type_array()`, `type_string()`, and friends, and the model returns JSON that `ellmer` automatically converts into R objects.

We will work with the chapter "Religious Diversity in the United States" by Dale E. Jones from the 2020 U.S. Religion Census, which is available as a plain text file in our course data repository.

```{r}
jones_text <- read_lines(
  "https://raw.githubusercontent.com/dwillis/religion-data/refs/heads/main/ReligiousDiversity-Jones1.txt"
)

# How long is the document?
length(jones_text)
```

```{r}
# Collapse into a single string for the prompt
jones_full <- paste(jones_text, collapse = "\n")
```

Now let's extract Table 1 from the chapter — the most and least diverse major metropolitan areas — as a structured dataframe.

```{r}
chat3 <- chat_anthropic()

metro_data <- chat3$chat_structured(
  jones_full,
  type = type_array(
    items = type_object(
      .description = "One metropolitan area entry from Table 1.",
      metro_area = type_string("Full name of the metropolitan area as it appears in the table"),
      rank       = type_integer("The rank number (1 through 5 for most diverse, 52 through 56 for least diverse)"),
      tier       = type_enum(c("most diverse", "least diverse"), "Whether this metro is among the most or least religiously diverse")
    )
  )
)

metro_data
```

`ellmer` automatically converted the array of objects into a dataframe.

### Verify the extraction

This is the most important step. Open the original text file and find Table 1. Does every row in `metro_data` match what is actually written there? Check: Are any metros missing? Are any ranks wrong? Did the model hallucinate a city that is not in the table?

**ANSWER HERE**

## Part 3: Structured Extraction from a PDF

One of the most powerful uses of LLMs in journalism is extracting tables from PDFs — annual reports, government documents, academic papers — that are not available as clean CSVs. Claude can read PDF content directly when you use `content_pdf_url()`.

We will extract Table 6 from the Dollhopf chapter: the ten states with the largest increase in adherents as a percentage of population between 2010 and 2020. The table has four data columns: adherents as a percent of population in 2010, adherents as a percent of population in 2020, and the net difference.

```{r}
chat4 <- chat_anthropic()

pdf_url <- "https://raw.githubusercontent.com/dwillis/religion-data/refs/heads/main/Trends2010to2020-Dollhopf.pdf"

table6_data <- chat4$chat_structured(
  content_pdf_url(pdf_url),
  type = type_array(
    type_object(
      .description = "One row from Table 6: States with the Largest Increase in Adherents as a % of Population, 2010-2020.",
      rank           = type_integer("The rank number, 1 through 10"),
      state          = type_string("The name of the state"),
      pct_2010       = type_number("Adherents as a percentage of population in 2010 (e.g. 39.13 for 39.13%)"),
      pct_2020       = type_number("Adherents as a percentage of population in 2020 (e.g. 47.08 for 47.08%)"),
      net_difference = type_number("Net difference in percentage points between 2010 and 2020 (e.g. 7.95 for 7.95%)")
    )
  )
)

table6_data
```

```{r}
# Should be 10 — one row per state in the table
nrow(table6_data)
```

```{r}
# Sort by net difference to confirm rank order
table6_data |> arrange(desc(net_difference))
```

```{r}
# Compute pct_2020 - pct_2010 ourselves and check against net_difference
table6_data |>
  mutate(computed_diff = pct_2020 - pct_2010) |>
  select(rank, state, net_difference, computed_diff)
```

### Verify the extraction

The `computed_diff` column above should match `net_difference` for each state. But the percentages themselves came from the PDF — and LLMs can mis-read numbers in tables.

Open the PDF and find Table 6. Does each state's `pct_2010`, `pct_2020`, and `net_difference` match exactly what is printed in the table?

**ANSWER HERE**

## Part 4: The Limits of LLM Extraction

The examples above show LLMs working well. But there are predictable failure modes every journalist should know.

**Hallucination of numbers.** LLMs are trained to produce plausible text. When a table is complex or a number appears ambiguous, the model may substitute a plausible-sounding value. Always cross-reference numeric claims against the source.

**Context collapse.** A finding that is true of one denomination may be attributed to another when the model processes a long document with many similar-sounding facts.

**Incomplete extraction.** `extract_data()` with a long document may miss some rows of a table, especially if the table is split across pages in the PDF.

**Prompt sensitivity.** Small wording changes in your `type_object()` descriptions can produce meaningfully different results. If your extraction looks wrong, try rephrasing the field descriptions before assuming the data is bad.

Let's test prompt sensitivity by re-running the Table 6 extraction without telling the model which table to use. Instead of naming Table 6 specifically, we'll ask for states and percentage data in general terms:

```{r}
chat5 <- chat_anthropic()

table6_vague <- chat5$chat_structured(
  content_pdf_url(pdf_url),
  type = type_array(
    type_object(
      .description = "A row about a state and its religious adherent percentages.",
      rank           = type_integer("A rank number"),
      state          = type_string("A state name"),
      pct_2010       = type_number("A percentage from 2010"),
      pct_2020       = type_number("A percentage from 2020"),
      net_difference = type_number("The change between the two percentages")
    )
  )
)

table6_vague
```

Now compare the two extractions:

```{r}
# Did we get the same number of rows?
nrow(table6_data)
nrow(table6_vague)
```

```{r}
# Are the states and values the same?
table6_data |> arrange(rank) |> select(rank, state, net_difference)
table6_vague |> arrange(rank) |> select(rank, state, net_difference)
```

Do the two extractions return the same states and values? If they differ, which rows changed and what does that tell you about how specific your type descriptions need to be?
