---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the Maryland churches data and the megachurches data.

**Task** Run the codeblock below to load the data, which is contained in CSV files.

```{r}
md_churches <- read_csv("churches_md.csv")
megachurches <- read_csv("megachurches.csv")
```

## Answer questions

**Q1** Which NTEE classification code (the NTEE_CD column) has the most churches in Maryland? Show the classification codes (you can find out what they mean here: https://www.irs.gov/pub/irs-soi/eo-info.pdf) with the highest count first. Is this result useful? Why or why not? **A1**

```{r}
```

**Q2** Find out which denomination has the greatest average (mean) size for its megachurches. Your code must show the denomination with the greatest average size first. **A2**

```{r}
```

**Q3** Count the number of megachurches for each state and city combination. Which state/city combination has the most megachurches? Which city is the most surprising, and why? **A3**

```{r}
```

**Q4** Given the results for Q3, what's a story that you could tell with this information? Think broadly, and consider what additional information you might need to make your story stronger. You also should think about what cities and states aren't in the results. **A4**

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**

```{r}
```
