---
title: "Reverse Engineering: Streaming Church"
author: "Student Name"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

In this reverse engineering assignment, you will analyze data from the Pew Religious Landscape Study to reproduce key findings from Ryan Burge's article: ["The People Streaming Church Aren't Who You Think"](https://www.graphsaboutreligion.com/p/the-people-streaming-church-arent) published on his Substack newsletter "Graphs about Religion."

The article examines who attends religious services online vs. in-person, revealing surprising patterns related to education, income, race, and generation. A copy of the article has been saved as an HTML file in this folder.

### Your Assignment

1. **Read the article carefully** and identify the key facts and figures the reporter presents
2. **Examine the data** provided in the `data/` folder
3. **Write your own code** to reproduce those key findings
4. **Document your work** with clear comments explaining your analysis

## The Data

The data comes from the Pew Religious Landscape Study and contains about 37,000 respondents. Each row is a survey respondent.

Key variables:
- `weight`: Survey weight for representative estimates
- `attndperrls`: In-person attendance frequency (1-2 = weekly+, 3-4 = occasionally, 5 = seldom, 6 = never)
- `attndonrls`: Online attendance frequency (same coding)
- `educrec`: Education level (1 = HS or less, 3 = some college/bachelor's, 4 = post-grad)
- `inc_sdt1`: Income category (1-3 = < $50k, 4-5 = $50k-$100k, 6-8 = $100k+)
- `hisp`: Hispanic identification (1 = yes, 2 = no)
- `racecmb`: Race (1 = White, 2 = Black, 3 = Asian, 4-5 = Other)
- `gender`: Gender (1 = male, 2 = female)
- `birthdecade`: Birth decade (1 = 1940s through 7 = 2000s)

## Load Libraries and Data

```{r}
library(tidyverse)

rls <- read_csv("data/rls_streaming.csv")
glimpse(rls)
```

## Finding 1: Education Gap in In-Person vs. Online Attendance

**Article claim:** Those with a high school education or less attend in-person services at higher rates (~25%) than college graduates (~19-20%). But the pattern reverses for online attendance -- college graduates stream services at higher rates (~17%) than those with HS or less (~13-14%).

```{r}
# Create education categories and weekly attendance indicators
education_attendance <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    in_person_weekly = attndperrls %in% c(1, 2),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2)) %>%
  group_by(educ2) %>%
  summarise(
    in_person_pct = weighted.mean(in_person_weekly, weight, na.rm = TRUE),
    online_pct = weighted.mean(online_weekly, weight, na.rm = TRUE),
    .groups = "drop"
  )

education_attendance %>%
  mutate(
    in_person_pct = round(in_person_pct * 100, 1),
    online_pct = round(online_pct * 100, 1)
  )

# Calculate the gaps
cat("\nIn-person gap (HS or Less minus College):",
    round((education_attendance$in_person_pct[education_attendance$educ2 == "HS or Less"] -
           education_attendance$in_person_pct[education_attendance$educ2 == "College"]) * 100, 1),
    "percentage points\n")

cat("Online gap (College minus HS or Less):",
    round((education_attendance$online_pct[education_attendance$educ2 == "College"] -
           education_attendance$online_pct[education_attendance$educ2 == "HS or Less"]) * 100, 1),
    "percentage points\n")
```

**Your interpretation:** Does the data confirm that less-educated Americans attend in-person more, but college-educated Americans stream more?

## Finding 2: The "Never/Never" Group - Most People Never Attend Either Way

**Article claim:** The largest group of Americans never attends services either in-person or online. Among college-educated Americans, approximately 35-37% never attend in either mode, compared to about 28-30% of those with HS or less.

```{r}
# Calculate percentages for the "Never/Never" combination by education
never_never <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    never_in_person = attndperrls == 6,
    never_online = attndonrls == 6,
    never_either = never_in_person & never_online
  ) %>%
  filter(!is.na(educ2)) %>%
  group_by(educ2) %>%
  summarise(
    never_either_pct = weighted.mean(never_either, weight, na.rm = TRUE),
    never_in_person_pct = weighted.mean(never_in_person, weight, na.rm = TRUE),
    never_online_pct = weighted.mean(never_online, weight, na.rm = TRUE),
    .groups = "drop"
  )

never_never %>%
  mutate(across(where(is.numeric), ~ round(. * 100, 1)))

cat("\nPercentage who never attend either in-person or online:\n")
cat("HS or Less:", round(never_never$never_either_pct[never_never$educ2 == "HS or Less"] * 100, 1), "%\n")
cat("College:", round(never_never$never_either_pct[never_never$educ2 == "College"] * 100, 1), "%\n")
```

**Your interpretation:** Is the "never attend anything" group really the largest single category? Are college-educated Americans more likely to fall into this group?

## Finding 3: Combined Weekly Attendance (In-Person or Online) by Education

**Article claim:** Among those with a high school diploma, about 34% attend either in person or online weekly or more. Among those with a bachelor's degree, about 30% attend either mode weekly.

```{r}
# Calculate percentage attending weekly in EITHER mode by education
combined_weekly <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    in_person_weekly = attndperrls %in% c(1, 2),
    online_weekly = attndonrls %in% c(1, 2),
    either_weekly = in_person_weekly | online_weekly
  ) %>%
  filter(!is.na(educ2)) %>%
  group_by(educ2) %>%
  summarise(
    in_person_only = round(weighted.mean(in_person_weekly & !online_weekly, weight, na.rm = TRUE) * 100, 1),
    online_only = round(weighted.mean(!in_person_weekly & online_weekly, weight, na.rm = TRUE) * 100, 1),
    both_weekly = round(weighted.mean(in_person_weekly & online_weekly, weight, na.rm = TRUE) * 100, 1),
    either_weekly = round(weighted.mean(either_weekly, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

combined_weekly
```

**Your interpretation:** Does the data show about 34% of HS-or-less and 30% of college grads attending weekly in either mode?

## Finding 4: Income Does Not Explain the Education Gap in Online Attendance

**Article claim:** Even when controlling for income, college-educated Americans still stream services at higher rates than those with less education at every income level. The education gap persists across income brackets.

```{r}
# Analyze online attendance by education and income
income_education_online <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    income3 = case_when(
      inc_sdt1 %in% 1:3 ~ "< $50k",
      inc_sdt1 %in% 4:5 ~ "$50k-$100k",
      inc_sdt1 %in% 6:8 ~ "$100k+",
      TRUE ~ NA_character_
    ),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2), !is.na(income3)) %>%
  group_by(educ2, income3) %>%
  summarise(
    online_pct = round(weighted.mean(online_weekly, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = educ2, values_from = online_pct) %>%
  mutate(college_advantage = College - `HS or Less`)

income_education_online

cat("\nCollege advantage in online attendance by income level:\n")
for (i in 1:nrow(income_education_online)) {
  cat(income_education_online$income3[i], ":",
      income_education_online$college_advantage[i], "percentage points\n")
}
```

**Your interpretation:** Does the college advantage in online attendance persist at every income level? Does income explain the education gap?

## Finding 5: Black Americans Attend at Higher Rates in Both Modes

**Article claim:** Black Americans, regardless of education level, attend both in-person and online services at substantially higher rates than White Americans. College-educated Black Americans attend in-person at about 30-32% and online at about 25-27%, roughly double the rates for White Americans.

```{r}
# Analyze attendance by race and education
race_attendance <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    race_cat = case_when(
      hisp == 1 ~ "Hispanic",
      hisp == 2 & racecmb == 1 ~ "White",
      hisp == 2 & racecmb == 2 ~ "Black",
      TRUE ~ NA_character_
    ),
    in_person_weekly = attndperrls %in% c(1, 2),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2), !is.na(race_cat)) %>%
  group_by(race_cat, educ2) %>%
  summarise(
    in_person_pct = round(weighted.mean(in_person_weekly, weight, na.rm = TRUE) * 100, 1),
    online_pct = round(weighted.mean(online_weekly, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

race_attendance

# Focus on Black vs White comparison among college-educated
cat("\nCollege-educated weekly attendance rates:\n")
bw_college <- race_attendance %>% filter(educ2 == "College", race_cat %in% c("Black", "White"))
for (i in 1:nrow(bw_college)) {
  cat(bw_college$race_cat[i], "- In-person:", bw_college$in_person_pct[i], "%,",
      "Online:", bw_college$online_pct[i], "%\n")
}
```

**Your interpretation:** Does the data confirm that Black Americans attend at substantially higher rates in both modes? How large are the gaps?

## Summary

Write a summary of how these five findings demonstrate the article's key insight: the relationship between education and religious attendance is more complex than a simple "more education = less religion" story. The online/streaming dimension reveals patterns that challenge conventional assumptions.
