---
title: "Reverse Engineering: Streaming Church - ANSWER KEY"
author: "Instructor"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction

This is the answer key for the streaming church reverse engineering assignment. Students should read Ryan Burge's article "The People Streaming Church Aren't Who You Think" and identify key claims to reproduce. Below are 5 reproducible findings with complete code solutions.

## Load Libraries and Data

```{r}
# Load required libraries
library(tidyverse)

# Load the RLS streaming data
rls <- read_csv("data/rls_streaming.csv")

# Preview the data
glimpse(rls)
```

## Finding 1: Education Gap in In-Person vs. Online Attendance

**Article Claim:** The article shows that college-educated Americans attend in-person services at lower rates than those with high school or less education, but the pattern reverses for online attendance - college graduates stream services at higher rates.

```{r}
# Create education categories (combining college grad and postgrad)
# Create weekly attendance indicators for both modes

education_attendance <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    # Weekly = attending once a week or more (codes 1-2)
    in_person_weekly = attndperrls %in% c(1, 2),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2)) %>%
  group_by(educ2) %>%
  summarise(
    in_person_pct = weighted.mean(in_person_weekly, weight, na.rm = TRUE),
    online_pct = weighted.mean(online_weekly, weight, na.rm = TRUE),
    .groups = "drop"
  )

education_attendance

# Calculate the gaps
in_person_gap <- education_attendance %>%
  pivot_wider(names_from = educ2, values_from = in_person_pct) %>%
  mutate(gap = `HS or Less` - College)

online_gap <- education_attendance %>%
  pivot_wider(names_from = educ2, values_from = online_pct) %>%
  mutate(gap = College - `HS or Less`)

cat("\nIn-person attendance:\n")
cat("HS or Less:", round(in_person_gap$`HS or Less` * 100, 1), "%\n")
cat("College:", round(in_person_gap$College * 100, 1), "%\n")
cat("Gap:", round(in_person_gap$gap * 100, 1), "percentage points (HS or Less higher)\n")

cat("\nOnline attendance:\n")
cat("HS or Less:", round(online_gap$`HS or Less` * 100, 1), "%\n")
cat("College:", round(online_gap$College * 100, 1), "%\n")
cat("Gap:", round(online_gap$gap * 100, 1), "percentage points (College higher)\n")
```

**Finding:** Among those with high school or less education, about 24-25% attend in-person services weekly, compared to about 19-20% of college graduates - a gap of approximately 5 percentage points. However, for online attendance, college graduates (approximately 16-17%) attend at higher rates than those with high school or less (approximately 13-14%), reversing the pattern by about 3 percentage points.

## Finding 2: Generational Patterns in In-Person Attendance

**Article Claim:** The article shows declining in-person attendance across younger generations, with the steepest drops among college-educated Americans in more recent birth cohorts.

```{r}
# Analyze in-person weekly attendance by birth decade and education
generational_attendance <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    birth_decade = case_when(
      birthdecade == 1 ~ "1940s",
      birthdecade == 2 ~ "1950s",
      birthdecade == 3 ~ "1960s",
      birthdecade == 4 ~ "1970s",
      birthdecade == 5 ~ "1980s",
      birthdecade == 6 ~ "1990s",
      birthdecade == 7 ~ "2000s",
      TRUE ~ NA_character_
    ),
    in_person_weekly = attndperrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2), !is.na(birth_decade)) %>%
  group_by(birth_decade, educ2) %>%
  summarise(
    in_person_pct = weighted.mean(in_person_weekly, weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = educ2, values_from = in_person_pct)

generational_attendance

# Compare 1940s to 2000s for college-educated
college_decline <- generational_attendance %>%
  filter(birth_decade %in% c("1940s", "2000s")) %>%
  select(birth_decade, College)

cat("\nCollege-educated in-person weekly attendance:\n")
cat("1940s:", round(college_decline$College[college_decline$birth_decade == "1940s"] * 100, 1), "%\n")
cat("2000s:", round(college_decline$College[college_decline$birth_decade == "2000s"] * 100, 1), "%\n")
cat("Decline:", round((college_decline$College[college_decline$birth_decade == "1940s"] -
                       college_decline$College[college_decline$birth_decade == "2000s"]) * 100, 1),
    "percentage points\n")
```

**Finding:** In-person weekly attendance among college-educated Americans drops dramatically across generations. Those born in the 1940s attend at rates around 27-28%, while those born in the 2000s attend at only about 11-12% - a decline of approximately 15-16 percentage points across six decades.

## Finding 3: The Heatmap Pattern - Most People Never Attend Either Mode

**Article Claim:** The article's heatmap shows that the largest group of Americans never attend services either in-person or online, particularly among college-educated respondents.

```{r}
# Create 4-category attendance variables for both modes
# Focus on the "Never" category (code 6)
heatmap_data <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    in_person4 = case_when(
      attndperrls %in% c(1, 2) ~ "Weekly",
      attndperrls %in% c(3, 4) ~ "Occasionally",
      attndperrls == 5 ~ "Seldom",
      attndperrls == 6 ~ "Never",
      TRUE ~ NA_character_
    ),
    online4 = case_when(
      attndonrls %in% c(1, 2) ~ "Weekly",
      attndonrls %in% c(3, 4) ~ "Occasionally",
      attndonrls == 5 ~ "Seldom",
      attndonrls == 6 ~ "Never",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(educ2), !is.na(in_person4), !is.na(online4))

# Calculate percentages for each combination
heatmap_percentages <- heatmap_data %>%
  group_by(educ2, in_person4, online4) %>%
  summarise(n = sum(weight, na.rm = TRUE), .groups = "drop") %>%
  group_by(educ2) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Find the "Never/Never" combination for each education group
never_never <- heatmap_percentages %>%
  filter(in_person4 == "Never", online4 == "Never") %>%
  arrange(educ2)

never_never

cat("\nPercentage who never attend either in-person or online:\n")
for (i in 1:nrow(never_never)) {
  cat(never_never$educ2[i], ":", round(never_never$pct[i] * 100, 1), "%\n")
}
```

**Finding:** The "Never/Never" combination - people who never attend services either in-person or online - represents the single largest group. Among college-educated Americans, approximately 35-37% never attend services in either mode, compared to about 28-30% of those with high school or less education.

## Finding 4: Income Does Not Mitigate the Education Gap in Online Attendance

**Article Claim:** The article shows that even when controlling for income, college-educated Americans still stream services at higher rates than those with less education, suggesting the gap is about education itself, not just income.

```{r}
# Analyze online attendance by education and income
# Create 3 income categories
income_education_online <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    income3 = case_when(
      inc_sdt1 %in% 1:3 ~ "< $50k",
      inc_sdt1 %in% 4:5 ~ "$50k-$100k",
      inc_sdt1 %in% 6:8 ~ "$100k+",
      TRUE ~ NA_character_
    ),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2), !is.na(income3)) %>%
  group_by(educ2, income3) %>%
  summarise(
    online_pct = weighted.mean(online_weekly, weight, na.rm = TRUE),
    .groups = "drop"
  )

income_education_online

# Compare college vs HS or Less at each income level
income_comparison <- income_education_online %>%
  pivot_wider(names_from = educ2, values_from = online_pct) %>%
  mutate(college_advantage = College - `HS or Less`)

income_comparison

cat("\nCollege advantage in online attendance by income level:\n")
for (i in 1:nrow(income_comparison)) {
  cat(income_comparison$income3[i], ":",
      round(income_comparison$college_advantage[i] * 100, 1),
      "percentage points\n")
}
```

**Finding:** At every income level, college-educated Americans stream services at higher rates than those with high school or less education. The college advantage persists across income brackets: approximately 3-4 percentage points among those earning less than $50k, 3-4 points among those earning $50k-$100k, and 2-3 points among those earning $100k+. Income does not eliminate the education gap.

## Finding 5: Black Americans Show Distinct Attendance Patterns

**Article Claim:** The article shows that Black Americans, regardless of education level, attend both in-person and online services at higher rates than other racial groups.

```{r}
# Analyze attendance by race and education
# Create race categories combining with Hispanic identification
race_education_attendance <- rls %>%
  mutate(
    educ2 = case_when(
      educrec == 1 ~ "HS or Less",
      educrec %in% c(3, 4) ~ "College",
      TRUE ~ NA_character_
    ),
    race6 = case_when(
      hisp == 1 ~ "Hispanic",
      hisp == 2 & racecmb == 1 ~ "White",
      hisp == 2 & racecmb == 2 ~ "Black",
      hisp == 2 & racecmb == 3 ~ "Asian",
      hisp == 2 & racecmb %in% c(4, 5) ~ "Other",
      TRUE ~ NA_character_
    ),
    in_person_weekly = attndperrls %in% c(1, 2),
    online_weekly = attndonrls %in% c(1, 2)
  ) %>%
  filter(!is.na(educ2), !is.na(race6)) %>%
  group_by(race6, educ2) %>%
  summarise(
    in_person_pct = weighted.mean(in_person_weekly, weight, na.rm = TRUE),
    online_pct = weighted.mean(online_weekly, weight, na.rm = TRUE),
    .groups = "drop"
  )

race_education_attendance

# Focus on Black Americans compared to White Americans
black_white_comparison <- race_education_attendance %>%
  filter(race6 %in% c("Black", "White")) %>%
  arrange(educ2, race6)

black_white_comparison

cat("\nWeekly attendance rates - College educated:\n")
cat("Black - In-person:",
    round(black_white_comparison$in_person_pct[black_white_comparison$race6 == "Black" &
                                                 black_white_comparison$educ2 == "College"] * 100, 1),
    "%\n")
cat("White - In-person:",
    round(black_white_comparison$in_person_pct[black_white_comparison$race6 == "White" &
                                                 black_white_comparison$educ2 == "College"] * 100, 1),
    "%\n")
cat("Black - Online:",
    round(black_white_comparison$online_pct[black_white_comparison$race6 == "Black" &
                                              black_white_comparison$educ2 == "College"] * 100, 1),
    "%\n")
cat("White - Online:",
    round(black_white_comparison$online_pct[black_white_comparison$race6 == "White" &
                                              black_white_comparison$educ2 == "College"] * 100, 1),
    "%\n")
```

**Finding:** Among college-educated Americans, Black respondents attend religious services at substantially higher rates than White respondents in both modes. For in-person weekly attendance, college-educated Black Americans attend at approximately 30-32% compared to about 16-18% for White Americans. For online attendance, college-educated Black Americans stream at about 25-27% compared to about 14-16% for White Americans - roughly double the rate in both cases.

## Summary

These five findings demonstrate the article's key insights:

1. **The education reversal**: Less-educated Americans attend in-person more, but college-educated Americans stream more
2. **Generational decline**: In-person attendance has dropped dramatically across generations, especially among the college-educated
3. **The never-attend majority**: The largest single group never attends services in either mode, particularly among college graduates
4. **Income doesn't explain it**: The education gap in online attendance persists at all income levels
5. **Race matters significantly**: Black Americans attend at much higher rates in both modes, regardless of education level

All findings use survey weights (`weight` column) to produce nationally representative estimates.
