---
title: "Reverse Engineering: Money and Leadership in the PCA"
author: "Your Name"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

## Introduction

This reverse engineering project is based on Ryan Burge's analysis ["Money and Leadership in the Presbyterian Church in America"](https://www.graphsaboutreligion.com/p/money-and-leadership-in-the-presbyterian) published on January 1, 2026.

The analysis examines over 1,700 PCA churches to answer practical questions: How many leaders should a church have? What's normal for per capita giving? How do finances vary by church size?

A copy of the article has been saved as an HTML file in this folder.

## Load Libraries and Data

```{r}
library(tidyverse)
options(scipen = 999)
```

```{r}
# Load the cleaned CSV (derived from the original pca_stats.xlsx)
# Columns: city, state, church, total_members, additions, losses,
#   ruling_elders, deacons, tithes_offerings, building_fund_offering,
#   total_contributions, other_income, total_church_income,
#   current_expenses, building_fund_disbursed, grand_total_disbursements,
#   per_capita_giving
pca <- read_csv("pca_clean.csv")

glimpse(pca)
```

```{r}
# Create size categories and total leaders variable
pca <- pca %>%
  mutate(
    total_leaders = ruling_elders + deacons,
    size_cat = case_when(
      total_members <= 25 ~ "1-25",
      total_members <= 50 ~ "26-50",
      total_members <= 100 ~ "51-100",
      total_members <= 250 ~ "101-250",
      total_members <= 500 ~ "251-500",
      total_members > 500 ~ "500+",
      TRUE ~ NA_character_
    ),
    size_cat = factor(size_cat, levels = c("1-25", "26-50", "51-100", "101-250", "251-500", "500+"))
  )
```

## Finding 1: Average Number of Leaders by Church Size

**Article claim:** In the very smallest churches, you'll find about 3 people serving in formal leadership roles. That rises to about 4 in churches with 26-50 members, about 10 in churches with 100-250 members, and roughly 37 in the largest PCA churches.

```{r}
# Calculate total leaders (ruling elders + deacons) by church size category
pca %>%
  filter(!is.na(size_cat)) %>%
  group_by(size_cat) %>%
  summarise(
    n_churches = n(),
    avg_leaders = round(mean(total_leaders, na.rm = TRUE), 1),
    avg_elders = round(mean(ruling_elders, na.rm = TRUE), 1),
    avg_deacons = round(mean(deacons, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

**Your interpretation:** Does the data support the article's claims about 3 leaders in the smallest churches, 4 in 26-50, 10 in 100-250, and 37 in the largest?

## Finding 2: Consistent Elder-to-Deacon Ratio

**Article claim:** For every one deacon, there's one elder. It's a near-perfect 50/50 split that barely wobbles across the entire dataset.

```{r}
# Calculate the elder and deacon share by size category
pca %>%
  filter(!is.na(size_cat)) %>%
  group_by(size_cat) %>%
  summarise(
    avg_elders = round(mean(ruling_elders, na.rm = TRUE), 1),
    avg_deacons = round(mean(deacons, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  mutate(
    elder_share_pct = round(avg_elders / (avg_elders + avg_deacons) * 100, 1)
  )

# Overall ratio
overall <- pca %>%
  summarise(
    total_elders = sum(ruling_elders, na.rm = TRUE),
    total_deacons = sum(deacons, na.rm = TRUE)
  ) %>%
  mutate(elder_share = round(total_elders / (total_elders + total_deacons) * 100, 1))

cat("Overall elder share:", overall$elder_share, "%\n")
cat("Overall deacon share:", 100 - overall$elder_share, "%\n")
```

**Your interpretation:** Is the ratio consistently around 50/50 across all church sizes?

## Finding 3: Members-per-Leader Ratio

**Article claim:** Across the entire sample, the average church has one leader for every twenty members. In a really small church, there's often one leader for every seven members. Among churches with 50-100 members, roughly one for sixteen. In the largest PCA churches, about one leader for every thirty-three members.

```{r}
# Overall members-per-leader ratio
pca %>%
  filter(total_leaders > 0) %>%
  mutate(members_per_leader = total_members / total_leaders) %>%
  summarise(
    avg_ratio = round(mean(members_per_leader, na.rm = TRUE), 0),
    median_ratio = round(median(members_per_leader, na.rm = TRUE), 0)
  )

# By size category
pca %>%
  filter(total_leaders > 0, !is.na(size_cat)) %>%
  mutate(members_per_leader = total_members / total_leaders) %>%
  group_by(size_cat) %>%
  summarise(
    n = n(),
    avg_members_per_leader = round(mean(members_per_leader, na.rm = TRUE), 0),
    .groups = "drop"
  )
```

**Your interpretation:** Is the overall ratio close to 20? Do the smallest churches really have about 1 leader per 7 members, and the largest about 1 per 33?

## Finding 4: Per Capita Giving Distribution

**Article claim:** About 7% have per-capita giving below $1,000. About 5% have giving above $7,000. The most common range is $2,000 to $4,000, which includes nearly half of all congregations. The median is $3,314.

```{r}
# Filter to churches with valid per capita giving data
pca_giving <- pca %>%
  filter(per_capita_giving > 0, !is.na(per_capita_giving))

n_total <- nrow(pca_giving)
cat("Total churches with valid giving data:", n_total, "\n\n")

# Below $1,000
below_1k <- sum(pca_giving$per_capita_giving < 1000)
cat("Below $1,000:", below_1k, "(", round(below_1k / n_total * 100, 1), "%) - article claims ~7%\n")

# Above $7,000
above_7k <- sum(pca_giving$per_capita_giving > 7000)
cat("Above $7,000:", above_7k, "(", round(above_7k / n_total * 100, 1), "%) - article claims ~5%\n")

# Between $2,000 and $4,000
mid_range <- sum(pca_giving$per_capita_giving >= 2000 & pca_giving$per_capita_giving <= 4000)
cat("$2,000-$4,000:", mid_range, "(", round(mid_range / n_total * 100, 1), "%) - article claims ~half\n")

# Median
cat("\nMedian per capita giving: $", format(round(median(pca_giving$per_capita_giving), 0), big.mark = ","),
    "- article claims $3,314\n")

# Mean
cat("Mean per capita giving: $", format(round(mean(pca_giving$per_capita_giving), 0), big.mark = ","), "\n")
```

**Your interpretation:** Do the percentages and median match the article's claims?

## Finding 5: Building Fund Prevalence

**Article claim:** Exactly one-third of all PCA churches had a line item for building funds. About 34% of shrinking churches and 41% of growing churches had building funds.

```{r}
# Calculate percentage of churches with building fund offerings
pca <- pca %>%
  mutate(has_building_fund = building_fund_offering > 0 & !is.na(building_fund_offering))

cat("Churches with building funds:", sum(pca$has_building_fund),
    "of", nrow(pca),
    "(", round(mean(pca$has_building_fund) * 100, 1), "%) - article claims 33%\n\n")

# By growth status (additions minus losses)
pca_growth <- pca %>%
  filter(!is.na(additions), !is.na(losses)) %>%
  mutate(
    net_change = additions - losses,
    growth_status = case_when(
      net_change > 0 ~ "Growing",
      net_change < 0 ~ "Shrinking",
      net_change == 0 ~ "No Change",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(growth_status))

pca_growth %>%
  group_by(growth_status) %>%
  summarise(
    n = n(),
    with_fund = sum(has_building_fund),
    pct_with_fund = round(mean(has_building_fund) * 100, 1),
    .groups = "drop"
  )
```

**Your interpretation:** Is the percentage close to 33%? Do growing churches have building funds at about 41% and shrinking churches at about 34%?

## Conclusion

Summarize which findings you were able to replicate and which (if any) differed from the article. What did you learn about PCA church finances and leadership patterns?

**Your conclusion:**
