---
title: "Reverse Engineering: Money and Leadership in the PCA"
author: "Your Name"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This reverse engineering project is based on Ryan Burge's analysis ["Money and Leadership in the Presbyterian Church in America"](https://www.graphsaboutreligion.com/p/money-and-leadership-in-the-presbyterian) published on January 1, 2026.

The analysis examines over 1,700 PCA churches to answer practical questions: How many leaders should a church have? What's normal for per capita giving? How do finances vary by church size?

## Load Libraries

```{r}
library(tidyverse)
library(readxl)
options(scipen = 999)
```

## Load Data

Read the PCA statistics Excel file. Note: Skip the first 2 rows to get proper column headers.

```{r}
pca <- read_excel("pca_stats.xlsx", skip = 2)

# Examine the structure
glimpse(pca)
```

## Finding 1: Leadership Patterns by Church Size

**Claim from article**: "In the very smallest churches, you'll usually find about three people serving in formal leadership roles. That number rises to around four deacons and elders in churches with 26–50 members... A church with 100–250 members averages about ten leaders, and among the largest PCA churches, that number climbs to roughly 37."

Calculate the average number of total leaders (ruling elders + deacons) by church size category.

```{r}
# Create size categories and calculate total leaders


# Group by size category and calculate means


# Display results

```

**Your interpretation**:

Does the data support the article's claims about 3 leaders in smallest churches, 4 in 26-50 member churches, 10 in 100-250, and 37 in largest churches?

## Finding 2: Consistent Elder-to-Deacon Ratio

**Claim from article**: "For every one deacon, there's one elder. It's a near-perfect 50/50 split that barely wobbles across the entire dataset."

Calculate the average number of elders and deacons separately by church size to verify this ratio.

```{r}
# Calculate mean elders and mean deacons by size category


# Calculate the ratio


# Display results

```

**Your interpretation**:

Is the ratio consistently around 1:1 across all church sizes?

## Finding 3: Leader-to-Member Ratio Overall

**Claim from article**: "Across the entire sample, the average church has one leader for every twenty members."

Calculate the members-per-leader ratio for all churches.

```{r}
# Calculate total leaders (elders + deacons)


# Calculate members per leader ratio


# Calculate mean ratio

```

**Your interpretation**:

Is the overall ratio close to 20 members per leader?

## Finding 4: Smallest Churches Have Tighter Ratios

**Claim from article**: "In a really small church, there's often one elder or deacon for every seven members."

Calculate the leader-to-member ratio for very small churches (25 members or fewer).

```{r}
# Filter to smallest churches


# Calculate members per leader ratio


# Display results

```

**Your interpretation**:

Do the smallest churches really have about 1 leader per 7 members?

## Finding 5: Ratio Increases with Size

**Claim from article**: "Among churches with 50–100 members, there's roughly one leader for every sixteen people... In the largest PCA churches, there's about one leader for every thirty-three members."

Calculate how the members-per-leader ratio changes by church size.

```{r}
# Calculate members per leader ratio by size category


# Display results

```

**Your interpretation**:

Does the ratio roughly double from small to medium churches, and double again to large churches?

## Finding 6: Per Capita Giving Distribution

**Claim from article**: "About 7% have a per-capita giving rate below $1,000 per person... about 5% of PCA congregations" have giving above $7,000. "The most common range is $2,000 to $4,000 per member, which includes nearly half of all congregations." The median is "$3,314."

Calculate the distribution of per capita giving.

```{r}
# Filter to valid per capita giving values


# Calculate percentage below $1,000


# Calculate percentage above $7,000


# Calculate percentage between $2,000 and $4,000


# Calculate median


# Display results

```

**Your interpretation**:

Do the percentages match the article's claims? Is the median close to $3,314?

## Finding 7: Per Capita Giving by Church Size

**Claim from article**: "Per-capita giving is about $3,500 per person in smaller congregations and rises only to around $3,850 among those at the upper end—a bump of roughly 10%." The largest churches have "about $4,343 per person."

Calculate mean per capita giving by church size category.

```{r}
# Calculate mean per capita giving by size


# Display results

```

**Your interpretation**:

Are the dollar amounts close to what the article reports? Is there only about a 10% increase from small to medium churches?

## Finding 8: Expense-to-Income Ratio Overall

**Claim from article**: "Across the average PCA church, about 86% of all income goes out through current expenses."

Calculate the ratio of current expenses to total church income.

```{r}
# Filter to churches with valid income and expense data


# Calculate expense ratio


# Calculate mean expense ratio

```

**Your interpretation**:

Is the mean expense ratio around 86% (or 0.86)?

## Finding 9: Small Churches Overspend

**Claim from article**: "Small churches are spending more than they're taking in. In fact, their expenses exceed their income by nearly 30%."

Calculate expense ratios for the smallest churches.

```{r}
# Filter to smallest churches (under 25 members)


# Calculate expense ratio


# Display results

```

**Your interpretation**:

Do the smallest churches have expense ratios above 100%? By how much?

## Finding 10: Large Churches Save Significantly

**Claim from article**: "Among churches with at least 500 members, every three dollars that comes in translates to just two dollars going out for expenses. That's a savings rate north of 30%."

Calculate expense ratios for the largest churches.

```{r}
# Filter to largest churches (500+ members)


# Calculate expense ratio


# Calculate savings rate (1 - expense ratio)


# Display results

```

**Your interpretation**:

Is the expense ratio around 0.67 (two-thirds)? Is the savings rate above 30%?

## Finding 11: Expense Ratio by Church Size

**Claim from article**: "Congregations with 26 to 250 members spend between 85% and 88% of their income each year."

Calculate mean expense ratios for different size categories.

```{r}
# Calculate mean expense ratio by size category


# Display results

```

**Your interpretation**:

Do medium-sized churches (26-250 members) have expense ratios in the 85-88% range?

## Finding 12: Building Fund Prevalence

**Claim from article**: "Exactly one-third of all PCA churches had a line item for building funds."

Calculate what percentage of churches have building fund offerings.

```{r}
# Count churches with building funds


# Calculate percentage


# Display results

```

**Your interpretation**:

Is the percentage close to 33%?

## Finding 13: Growth Status and Building Funds

**Claim from article**: "About 34% of churches that reported more subtractions than additions in the prior year still contributed to a building fund. Among those that grew in the past year, the figure was 41%."

Calculate building fund percentages by growth category.

```{r}
# Calculate net change (additions - losses)


# Create growth categories


# Calculate building fund percentage by growth category


# Display results

```

**Your interpretation**:

Do declining churches have building funds at about 34%, and growing churches at about 41%?

## Finding 14: Income Formula

**Claim from article**: "Total Church Income = (3,519.40 × Total Members) - 42,716. So, for example, a church with 100 members would have an expected income of about $309,000."

Build a linear model predicting total church income from total membership.

```{r}
# Filter to valid data (remove outliers above 99th percentile)


# Build linear model


# Display model summary


# Calculate predicted income for 100 members

```

**Your interpretation**:

Are the model coefficients close to what the article reports (slope ~3,519, intercept ~-42,716)? Does a 100-member church predict to around $309,000?

## Conclusion

Summarize which findings you were able to replicate and which (if any) differed from the article. What did you learn about PCA church finances and leadership patterns?

**Your conclusion**:
