---
title: "lab_10"
author: "Derek Willis"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries

```

Let's explore religious life in Maryland and make some maps!

## Questions

**Q1.** Let's look at the relationship between the number of congregations in a county and its number of adherents. Using the `md_religion_choropleth.csv` file in the data folder, calculate two rates for each county: congregations per 10,000 population and adherents per 10,000 population. You'll need to use tidycensus to get county population data for Maryland (the variable for total population is B01001_001; use 2021 and be sure to get the geometry, too).

Join the religion data to the population data, calculate both rates, then rank each county by both rates and calculate the difference in rankings. A county that ranks high on congregations but low on adherents has many small churches; the reverse suggests fewer but larger ones.

Make a county-level choropleth map of the congregations per 10,000 rate AND a second map of the adherents per 10,000 rate. What patterns do you see? Which counties have many small congregations vs. fewer large ones?

**A1.**

```{r}
```

**Q2.** Using the `churches_md.csv` file in the data folder, perform spatial joins to identify churches in Baltimore City and in Montgomery County. You'll need the Maryland zip code shapefile from the pre_lab (in `pre_labs/pre_lab_10/data/md_zips/`) and the county boundaries from tigris.

Count the number of churches per zip code in each jurisdiction and make a choropleth map for each one, using the same color scale. What's the story about church density in an urban jurisdiction (Baltimore City) vs. a suburban one (Montgomery County)? Where are the clusters, and what might explain the differences?

**A2.**

```{r}
```

**Q3.** Is there a relationship between religious adherence and political voting patterns? Using the `md_statewide_county.csv` file in the data folder (which has 2022 gubernatorial election results), calculate the percentage of the vote won by Dan Cox (the Republican candidate) in each county. Then join that to the religion data from Q1, which should already have the congregations per 10,000 rate and the adherent percentage.

Make a county-level choropleth map of the Cox vote percentage AND a second map of the congregations per 10,000 rate (or adherent percentage). Compare the two maps. Do counties with higher rates of religious congregations tend to vote more Republican? What are the exceptions, and what might explain them? What additional data would you want to explore this further?

HINT: The election data has a GEOID column and the religion data has a FIPS column -- these should match. You'll need to calculate total votes in the governor's race to get Cox's percentage.

**A3.**

```{r}
```

**Q4.** You've been assigned a story about the relationship between demographics and religious life in Maryland. Use tidycensus to retrieve a demographic variable of your choice at the county level for Maryland with geometry -- for example, percent of the population that is Black (B02001_003 divided by B01001_001), median age (B01002_001), or percent with a bachelor's degree (B15003_022 divided by B15003_001). You can pick any variable you think would be interesting.

Join that demographic data to the religion data from Q1 (or reload it). Make a county-level choropleth of your chosen demographic variable AND a second map of a religion variable (adherents rate, congregations rate, or adherent percentage). Compare the two maps.

What geographic patterns do you see? Do any counties stand out as having an unexpected combination of high/low demographics and high/low religious adherence? What would your next reporting steps be?

**A4.**

```{r}
```
